---
title: "Day 8 (PM) Lab: Logistic Regression -- Solutions"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```

---

# Exercise 1 -- Simple Logistic Regression with a Binary Predictor

## Part a: Tabulate CHD by smoking status

```{r}
framingham |>
  count(currentSmoker, TenYearCHD) |>
  group_by(currentSmoker) |>
  mutate(prop = round(n / sum(n), 3)) |>
  filter(TenYearCHD == "Yes")
```

About 15% of non-smokers develop CHD within 10 years compared with about 18%
of smokers. The raw proportions are similar, which will make the unadjusted OR
modest.

## Part b: Fit the simple logistic regression

```{r}
m_smoker <- glm(TenYearCHD ~ currentSmoker,
                data = framingham,
                family = binomial)

summary(m_smoker)
```

The coefficient on `currentSmokerSmoker` is positive (approximately 0.23),
indicating that smokers have higher log-odds of CHD than non-smokers. The
z-statistic and p-value indicate this difference is statistically significant.

## Part c: Odds ratio and 95% CI

```{r}
exp(cbind(OR = coef(m_smoker), confint(m_smoker))) |> round(3)
```

The OR for smoking is approximately 1.11 (95% CI roughly 0.94--1.32). Smokers
have about 11% higher odds of developing CHD within 10 years compared with
non-smokers. The confidence interval includes 1, so the association is
not statistically significant at alpha = 0.05.



## Part d: Verify against the 2x2 table

```{r}
tab <- table(framingham$currentSmoker, framingham$TenYearCHD)
tab

or_table <- (tab[2,2] * tab[1,1]) / (tab[1,2] * tab[2,1])
round(or_table, 3)
```

The cross-product OR from the 2x2 table matches the `exp(coef())` value from
the logistic model to at least three decimal places, confirming that simple
logistic regression with a single binary predictor is equivalent to computing
the odds ratio directly from a contingency table.

---

# Exercise 2 -- Simple Logistic Regression with a Continuous Predictor

## Part a: Inspect the relationship visually

```{r}
#| fig-width: 7
#| fig-height: 4.5
#| message: false
framingham |>
  mutate(CHD_numeric = as.numeric(TenYearCHD) - 1) |>
  ggplot(aes(x = sysBP, y = CHD_numeric)) +
  geom_jitter(alpha = 0.06, height = 0.04, color = "steelblue") +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE, linewidth = 1) +
  labs(x = "Systolic BP (mmHg)",
       y = "Probability of CHD within 10 years",
       title = "Empirical CHD Probability by Systolic BP") +
  theme_minimal()
```

The loess curve rises steadily with systolic BP. The relationship is
approximately monotonic with no sharp inflection points, supporting the use of
a single linear term on the log-odds scale.

## Part b: Fit the model, report coefficient, OR, and CI

```{r}
m_bp <- glm(TenYearCHD ~ sysBP,
            data = framingham,
            family = binomial)

summary(m_bp)

exp(cbind(OR = coef(m_bp), confint(m_bp))) |> round(4)
```

The coefficient on `sysBP` is approximately 0.024. The per-1-mmHg OR is about
1.024, with a confidence interval that excludes 1. The z-statistic is large and
the p-value is essentially zero, indicating very strong evidence that systolic
BP is associated with CHD risk.

The plot does indicate that the relationship may be quadratic. Let's take a look:

```{r}
m_bp_quad <- glm(TenYearCHD ~ poly(sysBP,2),
                data = framingham,
                family = binomial)

summary(m_bp_quad)
anova(m_bp,m_bp_quad)
```

The linear model provides a better fit to the data.

## Part c: Rescale to 20 mmHg

```{r}
exp(20 * coef(m_bp)["sysBP"])
exp(20 * confint(m_bp)["sysBP", ])
```

A 20 mmHg increase in systolic blood pressure is associated with approximately
a 62% increase in the odds of developing CHD within 10 years (OR approximately 1.5 to 2.0). This is a clinically meaningful unit: 20 mmHg is
roughly one standard deviation of systolic BP in this cohort and corresponds to
the difference between, say, 120 and 140 mmHg.

## Part d: Plot the fitted logistic curve

```{r}
#| fig-width: 7
#| fig-height: 4.5
#| message: false
framingham |>
  mutate(CHD_numeric = as.numeric(TenYearCHD) - 1) |>
  ggplot(aes(x = sysBP, y = CHD_numeric)) +
  geom_jitter(alpha = 0.06, height = 0.04, color = "steelblue") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "firebrick", se = TRUE, linewidth = 1) +
  labs(x = "Systolic BP (mmHg)",
       y = "Probability of CHD within 10 years",
       title = "Fitted Logistic Regression: CHD on Systolic BP") +
  theme_minimal()
```

The fitted curve is S-shaped and bounded between 0 and 1, rising from near 5%
at very low BP values to near 40% at very high values. Unlike a linear
regression line, it cannot predict probabilities outside the valid range.

---

# Exercise 3 -- Multiple Logistic Regression and Confounding

## Part a: Unadjusted OR for smoking

```{r}
m_smoke_only <- glm(TenYearCHD ~ currentSmoker,
                    data = framingham,
                    family = binomial)

exp(cbind(OR = coef(m_smoke_only), confint(m_smoke_only))) |> round(3)
```

The unadjusted OR for smoking is approximately 1.11 (same as Exercise 1).

## Part b: Adjusted OR for smoking, controlling for age

```{r}
m_smoke_age <- glm(TenYearCHD ~ currentSmoker + age,
                   data = framingham,
                   family = binomial)

exp(cbind(OR = coef(m_smoke_age), confint(m_smoke_age))) |> round(3)
```

After adjusting for age, the OR for smoking increases to approximately 1.54.

## Part c: Compare unadjusted and adjusted ORs

```{r}
framingham |>
  group_by(currentSmoker) |>
  summarize(mean_age = mean(age), .groups = "drop")
```

The adjusted OR for smoking (approximately 1.54) is **larger** than the
unadjusted OR (approximately 1.11). This is an example of negative confounding.
The mean age check reveals that smokers in this cohort are younger than
non-smokers by about 2--3 years. Since younger age is associated with lower CHD
risk, the younger age of smokers was partially masking (attenuating) the true
effect of smoking on CHD. Once we hold age constant, the smoking effect becomes
larger. Confounding can inflate or deflate an association depending on the direction of the confounder's relationship with both the exposure and the outcome.

## Part d: Likelihood ratio test for age

```{r}
anova(m_smoke_only, m_smoke_age)
```

The LRT p-value is essentially zero, providing very strong evidence that age
adds significant predictive information beyond smoking status alone. The large
chi-squared statistic reflects the fact that age is one of the strongest
predictors of CHD in the Framingham cohort.

---

# Exercise 4 -- Full Workflow: Building a CHD Risk Model

## Part a: Fit the full model

```{r}
vars_used <- c("TenYearCHD", "age", "male", "currentSmoker",
               "sysBP", "totChol", "BMI", "diabetes", "prevalentHyp")

framingham_cc <- framingham |>
  drop_na(all_of(vars_used))

nrow(framingham) - nrow(framingham_cc)

m_full <- glm(TenYearCHD ~ age + male + currentSmoker + sysBP +
                            totChol + BMI + diabetes + prevalentHyp,
              data = framingham_cc,
              family = binomial)

summary(m_full)
```

Age, sex (male), smoking, systolic BP, and diabetes all appear strongly
associated with CHD. Total cholesterol shows a positive association. BMI and
prevalent hypertension have weaker or less significant effects once the other
variables are in the model.

## Part b: AIC-based model selection

```{r}
m_step <- step(m_full, direction = "both", trace = 0)
summary(m_step)
AIC(m_full, m_step)
```

The stepwise procedure typically retains age, sex, smoking, sysBP, totChol,
and diabetes, and drops BMI and prevalentHyp. The selected model has a lower
AIC, indicating a modest improvement in fit per parameter.

## Part c: LRT for dropped variables

```{r}
anova(m_step, m_full, test = "LRT")
```

The LRT p-value for the variables dropped by stepwise selection (BMI and
prevalentHyp) is typically large (p > 0.05), confirming that these variables
do not significantly improve the model beyond the retained predictors. The AIC
and LRT are in agreement here.

## Part d: Adjusted ORs for the final model

```{r}
chd_final <- m_step

or_table <- exp(cbind(OR = coef(chd_final), confint(chd_final))) |>
  round(3)

or_table
```

**Interpretations:**

- **age (per 10 years):**
```{r}
exp(10 * coef(chd_final)["age"])
exp(10 * confint(chd_final)["age", ])
```
Each decade of age roughly doubles the odds of CHD after adjusting for the
other risk factors.

- **maleMale:** Males have substantially higher odds of CHD than females of the
  same age, smoking status, blood pressure, cholesterol, and diabetes status.

- **currentSmokerSmoker:** After adjusting for age and other factors, smokers
  have markedly higher CHD odds than non-smokers. Note this adjusted OR is
  larger than the unadjusted OR from Exercise 1, consistent with the negative
  confounding by age found in Exercise 3.

- **sysBP (per 20 mmHg):**
```{r}
exp(20 * coef(chd_final)["sysBP"])
exp(20 * confint(chd_final)["sysBP", ])
```
A 20 mmHg higher systolic BP is associated with roughly 33% higher odds
of CHD, holding the other variables constant.

- **totChol (per 40 mg/dL):**
```{r}
exp(100 * coef(chd_final)["totChol"])
exp(100 * confint(chd_final)["totChol", ])
```
A 40 mg/dL increase in total cholesterol is associated with a modest increase
in CHD odds.

- **diabetesYes:** Diabetic participants have substantially higher CHD odds even
  after adjusting for all other risk factors in the model.

## Part e: Predicted probabilities

```{r}
framingham_cc <- framingham_cc |>
  mutate(p_final = predict(chd_final, newdata = framingham_cc,
                           type = "response"))

framingham_cc |>
  select(TenYearCHD, p_final) |>
  summary()
```

```{r}
framingham_cc |>
  mutate(CHD_numeric = as.numeric(TenYearCHD) - 1) |>
  ggplot(aes(x = p_final, fill = TenYearCHD)) +
  geom_histogram(binwidth = 0.02, position = "identity",
                 alpha = 0.6, color = "white") +
  scale_fill_manual(values = c("No" = "steelblue", "Yes" = "firebrick")) +
  labs(x = "Predicted probability of CHD",
       y = "Count",
       fill = "Actual CHD",
       title = "Distribution of Predicted Probabilities by Actual Outcome") +
  theme_minimal()
```

The predicted probabilities range from near 0% to about 60--70%. The
distributions for the CHD and non-CHD groups overlap considerably, which is
typical for clinical risk models. The CHD group tends to have higher predicted
probabilities on average, but there is no clean separation. This overlap is
what motivates tomorrow's topic: how do we choose a threshold to turn these
probabilities into yes/no predictions, and how do we evaluate the tradeoffs?
