---
title: "Day 8 (PM) Lab: Logistic Regression"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This lab puts the Day 8 morning material into practice. You will fit, interpret,
and evaluate logistic regression models on the Framingham Heart Study data.

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** Follow along as the instructor works
  through a complete simple logistic regression with a different predictor than
  the morning example.
- **Part 2 (on your own, ~90 min):** Six exercises of increasing difficulty.
  Create a new `.qmd` file, work through each exercise, render when done.

The **Quick Reference Card** in Section 3 has the key functions.

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(this.path)
library(pROC)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```

## Step 1: Check Variable Form

We want to predict CHD risk (`TenYearCHD`) from age. First, inspect the
empirical CHD probability across age.

```{r}
#| message: false
framingham |>
  mutate(CHD_numeric = as.numeric(TenYearCHD) - 1) |>
  ggplot(aes(x = age, y = CHD_numeric)) +
  geom_jitter(alpha = 0.06, height = 0.04, color = "steelblue") +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE, linewidth = 1) +
  labs(x = "Age (years)",
       y = "Probability of CHD within 10 years",
       title = "Empirical CHD Probability by Age") +
  theme_minimal()
```

The loess curve rises steadily from low risk in younger participants to higher
risk in older participants. A logistic regression with age as a linear term on
the log-odds scale looks appropriate.

## Step 2: Fit the Model and Interpret

```{r}
chd_age <- glm(TenYearCHD ~ age,
               data = framingham,
               family = binomial)

summary(chd_age)
```

The coefficient on `age` is in log-odds units. To get the odds ratio:

```{r}
exp(cbind(OR = coef(chd_age), confint(chd_age)))
```

Each additional year of age multiplies the odds of CHD by about 1.10. Rescaled
to a decade:

```{r}
exp(10 * coef(chd_age)["age"])
exp(10 * confint(chd_age)["age", ])
```

A 10-year age difference is associated with roughly a 2.5-fold increase in the
odds of CHD over 10 years of follow-up.

## Step 3: Visualize the Fitted Curve

```{r}
#| message: false
framingham |>
  mutate(CHD_numeric = as.numeric(TenYearCHD) - 1) |>
  ggplot(aes(x = age, y = CHD_numeric)) +
  geom_jitter(alpha = 0.06, height = 0.04, color = "steelblue") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "firebrick", se = TRUE, linewidth = 1) +
  labs(x = "Age (years)",
       y = "Probability of CHD within 10 years",
       title = "Fitted Logistic Regression: CHD on Age") +
  theme_minimal()
```

## Step 4: Predicted Probabilities and Classification

```{r}
new_ages <- tibble(age = c(35, 45, 55, 65))
new_ages |>
  mutate(p_chd = predict(chd_age, newdata = new_ages, type = "response"))
```

Predicted probability of CHD rises from about 7% at age 35 to about 28% at age
65.

```{r}
framingham <- framingham |>
  mutate(p_age = predict(chd_age, type = "response"))

roc_age <- roc(framingham$TenYearCHD, framingham$p_age,
               levels = c("No","Yes"), direction = "<")
auc(roc_age)
```

The AUC for age alone is around 0.68. Age is a reasonable single predictor but
far from perfect. Adding more predictors should improve discrimination.

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document: File → New File → Quarto Document.
Title it `"Day 8 Lab: Logistic Regression"`, select HTML, click **Create**.

**Step 2.** Replace the template content with this YAML and setup chunk:

````markdown
---
title: "Day 8 Lab: Logistic Regression"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{{r}}
library(tidyverse)
library(this.path)
library(pROC)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```
````

**Step 3.** Work through Exercises 1--6 in order. For each one:

- Add a new code chunk (`Ctrl+Alt+I` / `Cmd+Option+I`)
- Write and run your code
- Add 2--3 sentences below each result describing what you found in plain
  language, including the clinical interpretation of any odds ratios

**Step 4.** Render (`Ctrl+Shift+K` / `Cmd+Shift+K`) when finished and fix
any errors before you are done.

::: {.callout-tip}
**Before asking for help**, check two things: (1) the **Quick Reference Card**
in Section 3, and (2) whether your factor variables are coded with the right
reference level. For `TenYearCHD`, "No" should be the reference (level 1) so
that `glm()` models the probability of "Yes".
:::

::: {.callout-note}
Exercises 1--4 cover the core material. Exercise 5 puts everything together in
a complete model-building workflow. Exercise 6 is a challenge problem on
classification performance.
:::

---

**Exercise 1 -- Simple Logistic Regression with a Binary Predictor.**

a. Use `count()` to tabulate `TenYearCHD` by `currentSmoker`. What proportion
   of smokers develop CHD? What proportion of non-smokers do?

b. Fit a simple logistic regression model predicting `TenYearCHD` from
   `currentSmoker`. Report the coefficient on `currentSmokerSmoker` from
   `summary()`.

c. Compute the odds ratio (OR) and its 95% confidence interval using
   `exp(cbind(OR = coef(model), confint(model)))`. Interpret the OR in plain
   language: how do the odds of CHD compare for smokers versus non-smokers?

d. Confirm that this OR matches what you would get from a 2x2 table:
   ```r
   tab <- table(framingham$currentSmoker, framingham$TenYearCHD)
   tab
   (tab[2,2] * tab[1,1]) / (tab[1,2] * tab[2,1])
   ```
   Does it match the OR from `glm()`?

```{r}
# your code here
```

---

**Exercise 2 -- Simple Logistic Regression with a Continuous Predictor.**

a. Use `geom_jitter()` and `geom_smooth(method = "loess")` to inspect the
   relationship between `sysBP` and `TenYearCHD`. Convert `TenYearCHD` to numeric
   with `as.numeric(TenYearCHD) - 1` before plotting.

b. Fit a simple logistic regression of `TenYearCHD` on `sysBP`. Report the
   coefficient, OR, and 95% CI.

c. The OR per 1 mmHg is hard to interpret. Compute the OR for a **20 mmHg
   increase** in systolic BP using `exp(20 * coef(model)["sysBP"])`. Compute
   the corresponding CI using `exp(20 * confint(model)["sysBP", ])`. Write
   one sentence interpreting this rescaled OR clinically.

d. Plot the fitted logistic curve overlaid on the data using
   `geom_smooth(method = "glm", method.args = list(family = "binomial"))`.
   Roughly what predicted probability does the model give for a person with
   a systolic BP of 180?

```{r}
# your code here
```

---

**Exercise 3 -- Continuous vs Categorical Predictors.**

This exercise demonstrates the same point we made for linear regression:
categorizing a continuous predictor wastes information.

a. Create a new variable called `age_cat` using `cut()` with breaks
   `c(30, 45, 60, 75)` and labels `c("30-45", "45-60", "60-75")`. Use
   `include.lowest = TRUE`.

b. Fit two logistic regression models predicting `TenYearCHD`:
   - `m_continuous`: uses `age` as a continuous predictor
   - `m_categorical`: uses `age_cat` as a factor

c. Compare AIC values with `AIC(m_continuous, m_categorical)`. Which fits
   better?

d. From `m_categorical`, what is the OR for the oldest age group (60-75)
   compared with the youngest (30-45)? How does this compare with the OR
   estimated by `m_continuous` for a 22.5-year age difference (the midpoint
   gap between the two extreme groups)?

```{r}
# your code here
```

---

**Exercise 4 -- Multiple Logistic Regression and Confounding.**

a. Fit a model predicting `TenYearCHD` from `currentSmoker` alone. Note the OR
   for smoking.

b. Now fit a model predicting `TenYearCHD` from both `currentSmoker` and `age`.
   Compute the **adjusted** OR for smoking (controlling for age).

c. Compare the unadjusted and adjusted ORs for smoking. Did the OR get larger,
   smaller, or stay the same? Why do you think this happened? (Hint: are
   smokers in this cohort younger or older on average than non-smokers? Check
   with `framingham |> group_by(currentSmoker) |> summarize(mean_age = mean(age))`.)

d. Test whether age is a useful addition to the model using
   `anova(model_smoker, model_smoker_age, test = "LRT")`. Interpret the
   p-value.

```{r}
# your code here
```

---

**Exercise 5 -- Full Workflow: Building a CHD Risk Model.**

Build a multivariable logistic regression model for CHD risk.

a. Start with a full model including: `age`, `male`, `currentSmoker`, `sysBP`,
   `totChol`, `BMI`, `diabetes`, and `prevalentHyp`. Fit it with `glm()` and
   examine `summary()`. Which predictors appear most strongly associated with
   CHD?

b. Use `step(model_full, direction = "both", trace = 0)` to perform AIC-based
   model selection. Which variables are retained?

c. Use `anova(reduced, full, test = "LRT")` to formally test whether the
   variables dropped by `step()` could be removed. Does the LRT agree with the
   AIC-based selection?

d. Report the adjusted ORs and 95% CIs for your final model. For each predictor
   retained, write one sentence interpreting the OR in clinical terms. Rescale
   continuous predictors to meaningful units (10 years for age, 20 mmHg for
   sysBP, 40 mg/dL for totChol).

e. Generate predicted probabilities for the final model and add them to the
   dataset as `p_final`.

```{r}
# your code here
```

---

**Exercise 6 (Challenge) -- Classification Performance and ROC.**

Continuing with the final model from Exercise 5, evaluate its predictive
performance.

a. Build a confusion matrix at three thresholds: 0.10, 0.20, and 0.50. Use
   `table(predicted, actual)` with predicted classes from
   `if_else(p_final > threshold, "Yes", "No")`.

b. For each threshold, compute sensitivity and specificity by hand from the
   confusion matrix. As the threshold increases, what happens to each? Does
   this match the tradeoff described in the morning lecture?

c. Compute the ROC curve and AUC using the `pROC` package:
   ```r
   roc_final <- roc(framingham$TenYearCHD, framingham$p_final,
                    levels = c("No","Yes"), direction = "<")
   plot(roc_final, col = "firebrick", lwd = 2)
   auc(roc_final)
   ```
   How would you describe the model's discriminative ability (poor,
   acceptable, excellent, outstanding)?

d. Use `coords(roc_final, "best", best.method = "youden")` to find the
   threshold that maximizes Youden's J statistic (sensitivity + specificity - 1).
   What is the "optimal" threshold by this criterion? How does it compare to
   the baseline event rate (about 0.15)?

e. **Discussion question.** The AUC you computed is on the same data used to
   fit the model. Why is this an optimistic estimate of how the model will
   perform on new data? What technique from earlier in the course could give
   a more honest estimate?

```{r}
# your code here
```

---

# Quick Reference Card

## Setup

```r
library(tidyverse)
library(this.path)
library(pROC)
setwd(this.dir())
framingham <- read_csv("../../data/framingham.csv")
```

## Checking Variable Form on a Binary Outcome

```r
# Empirical probability curve for one predictor
df |>
  mutate(Y_numeric = as.numeric(Y) - 1) |>
  ggplot(aes(x = predictor, y = Y_numeric)) +
  geom_jitter(alpha = 0.05, height = 0.04) +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE)
```

## Fitting Models

| Function | What it does |
|---|---|
| `glm(y ~ x, data = df, family = binomial)` | Simple logistic regression |
| `glm(y ~ x1 + x2, data = df, family = binomial)` | Multiple logistic regression |
| `summary(model)` | Coefficients, z-tests, deviance |
| `coef(model)` | Just the estimated coefficients |
| `confint(model)` | 95% CIs for coefficients (on log-odds scale) |
| `exp(coef(model))` | Odds ratios |
| `exp(confint(model))` | 95% CIs for ORs |
| `exp(cbind(OR = coef(m), confint(m)))` | OR table in one go |
| `AIC(m1, m2)` | Compare models, lower is better |
| `anova(m_small, m_large, test = "LRT")` | Likelihood ratio test for nested models |
| `step(model, direction = "both", trace = 0)` | AIC-based stepwise selection |

## Predicted Probabilities

```r
# Add fitted probabilities to the dataset
df <- df |> mutate(p_hat = predict(model, type = "response"))

# Predict for new individuals
new_data <- tibble(age = c(45, 55), sysBP = c(120, 140))
predict(model, newdata = new_data, type = "response")
```

## Plotting the Fitted Curve

```r
df |>
  mutate(Y_numeric = as.numeric(Y) - 1) |>
  ggplot(aes(x = predictor, y = Y_numeric)) +
  geom_jitter(alpha = 0.05, height = 0.04) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "firebrick", se = TRUE)
```

## Classification and Performance

```r
# Build a confusion matrix at a chosen threshold
df <- df |>
  mutate(pred_class = factor(if_else(p_hat > 0.15, "Yes", "No"),
                             levels = c("No", "Yes")))
tab <- table(Predicted = df$pred_class, Actual = df$Y)
tab

# Sensitivity = TP / (TP + FN)
# Specificity = TN / (TN + FP)

# ROC and AUC
roc_obj <- roc(df$Y, df$p_hat, levels = c("No", "Yes"), direction = "<")
plot(roc_obj, col = "firebrick", lwd = 2)
auc(roc_obj)

# Best threshold by Youden's J
coords(roc_obj, "best", best.method = "youden")
```

## Rescaling Odds Ratios to Meaningful Units

```r
# For a continuous predictor, the per-1-unit OR is often uninterpretable.
# Multiply the coefficient by the meaningful unit before exponentiating.
exp(10 * coef(model)["age"])           # 10-year OR
exp(20 * confint(model)["sysBP", ])    # 20 mmHg CI for sysBP
exp(40 * coef(model)["totChol"])       # 40 mg/dL OR
```

## Common Interpretive Phrases

| Quantity | How to say it |
|---|---|
| OR = 1.5 | "1.5-fold higher odds" or "50% higher odds" |
| OR = 2.0 | "double the odds" |
| OR = 0.5 | "half the odds" or "50% lower odds" |
| OR = 1.0 (CI includes 1) | "no evidence of an association" |
| Adjusted OR | "after controlling for [other variables]" |
