---
title: "Logistic Regression Analysis Examples in R"

format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    smooth-scroll: true
    
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This document demonstrates how to perform logistic regression analysis in R using generalized linear models (`glm`) with a binomial distribution and logit link function. The analyses include:

-   Creating datasets
-   Fitting logistic regression models
-   Interpreting model summaries
-   Computing odds ratios and confidence intervals
-   Evaluating goodness-of-fit
-   Generating predicted probabilities
-   Visualizing logistic regression curves


## Load Required Package

The following packages are used for the ROC and the Hosmer-Lemeshow goodness-of-fit test.

```{r}
# Install packages if necessary
# install.packages("ResourceSelection")
# install.packages("pROC")

# Load library
library(ResourceSelection)
library(pROC)
```

------------------------------------------------------------------------

# Example 1: Logistic Regression Using Framingham Dataset

## Read In the Dataset

We will use the Framingham dataset again for this example. As a reminder, it contains demographic and cardiovascular risk factor information collected for the Framingham Heart Study, a longitudinal study investigating predictors of coronary heart disease (CHD). Variables include:

- `age`: Age of the patient (in years)
- `sysBP`: Systolic blood pressure (mmHg)
- `totChol`: Total cholesterol level (mg/dL)
- `cigsPerDay`: Number of cigarettes smoked per day
- `currentSmoker`: Indicator of whether the patient is a current smoker (0 = no, 1 = yes)
- `diabetes`: Whether the patient has diabetes (0 = no, 1 = yes)
- `TenYearCHD`: Binary outcome indicating whether the patient developed coronary heart disease within 10 years:
  - `0`: No coronary heart disease developed within 10 years
  - `1`: Coronary heart disease developed within 10 years

```{r}
framingham <- read.csv("framingham.csv")

head(framingham)
```


## Categorical vs. Continuous Predictors

One important consideration in logistic regression is deciding whether a variable should be treated as:

- **Continuous** (numeric relationship across values)
- **Categorical** using `factor()` (separate groups/categories)

Using the wrong variable type can lead to misleading interpretations and poor model performance.

### When to Treat a Variable as Continuous

A variable should generally remain numeric when:

- The values represent a measurable quantity
- The spacing between values is meaningful
- A one-unit increase has a reasonable interpretation

Examples include variables like age, blood pressure, cholesterol, or BMI.

### When to Use `factor()`

A variable should be converted to a factor when:

- The variable represents categories or groups
- Numeric labels do not imply meaningful distance
- Categories should have separate effects

Examples include variables like sex, smoking status, treatment group, education level, or disease stage.

### What Can Go Wrong

Suppose our `age` variable was incorrectly converted to a factor:

```{r}

framingham$age <- factor(framingham$age)

glm(
  TenYearCHD ~ age,
  data = framingham,
  family = binomial(link = "logit")
)

```
Issues arise including:

- A parameter being created for every age value
- More difficulty interpreting the model
- The model becomes unnecessarily large
- Decreased statistical power
- Predictions become unstable

Continuous variables should typically remain numeric unless there is a strong scientific reason to group them.

## Fit the Logistic Regression Model

This model predicts the probability of coronary heart disease within 10 years (`TenYearCHD`) using systolic blood pressure (`sysBP`).

The logistic regression equation is:

$\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1X$

```{r}
logit_model <- glm(
  TenYearCHD ~ sysBP,
  data = framingham,
  family = binomial(link = "logit")
)

summary(logit_model)
```

------------------------------------------------------------------------

## Compute Odds Ratios and Confidence Intervals

The coefficients from logistic regression are in log-odds units. Exponentiating the coefficients converts them into odds ratios.


```{r}

odds_ratios <- exp(coef(logit_model))

# Compute 95% confidence intervals
conf_int <- exp(confint(logit_model))

cat("\nOdds Ratios:\n")
print(odds_ratios)

cat("\n95% Confidence Intervals:\n")
print(conf_int)
```

------------------------------------------------------------------------

## Generate Predicted Probabilities

Predicted probabilities estimate the likelihood of successful task completion for each observation.

```{r}
framingham$Predicted_Prob <- predict(
  logit_model,
  type = "response"
)

head(framingham)
```

------------------------------------------------------------------------

## Plot Observed Outcomes and Logistic Curve

This plot displays:

-   Observed binary outcomes
-   Predicted logistic regression probabilities

```{r}
plot(
  framingham$sysBP,
  framingham$TenYearCHD,
  xlab = "Systolic Blood Pressure",
  ylab = "Probability of CHD Within 10 Years",
  main = "Logistic Regression Model",
  pch = 19,
  col = "blue",
  ylim = c(-0.1, 1.1)
)

# Order values for smooth curve
ordered_index <- order(framingham$sysBP)

# Add logistic curve
lines(
  framingham$sysBP[ordered_index],
  framingham$Predicted_Prob[ordered_index],
  col = "red",
  lwd = 2
)

legend(
  "bottomright",
  legend = c("Observed Y", "Predicted Probability"),
  col = c("blue", "red"),
  pch = c(19, NA),
  lty = c(NA, 1),
  lwd = c(NA, 2)
)
```



------------------------------------------------------------------------

## Deviance Residuals

Deviance residuals are a tool to help us evaluate how well the logistic regression model fits individual observations.

Large residuals could indicate:
-   Poorly predicted observations
-   Outliers
-   Model misspecification

We can find them using the `residuals()` function:

```{r}

dev_resid <- residuals(
  logit_model,
  type = "deviance"
)

head(dev_resid)

```


```{r}

plot(
  dev_resid,
  ylab = "Deviance Residuals",
  main = "Deviance Residual Plot",
  pch = 19,
  col = "darkblue"
)

abline(h = 0, lty = 2, col = "red")

```

Things to look for:

- Residuals should be roughly centered around zero
- Extremely large positive or negative residuals may indicate problematic observations
- Systematic patterns suggest poor model fit

## ROC and AUC

The ROC curve evaluates how well the model distinguishes between individuals who experienced the outcome and individuals who did not.

The AUC (Area Under the Curve) summarizes overall classification performance of the model.

```{r}

roc_curve <- roc(
  framingham$TenYearCHD,
  framingham$Predicted_Prob
)

plot(
  roc_curve,
  col = "blue",
  lwd = 3,
  main = "ROC Curve for Logistic Regression Model"
)

```
```{r}

auc(roc_curve)

```

The blue curve represents the model’s performance across different probability cutoffs used for classification. 

The diagonal gray line represents random guessing. A model whose curve lies close to this line has very little ability to distinguish between the two groups.

An AUC of 0.5 means the model performs no better than random guessing (gray line) when trying to distinguish between the two outcome groups, while an AUC of 1 indicates a perfect model.

Generally, a higher AUC indicates that the model is better at distinguishing between individuals who develop CHD and those who do not.

------------------------------------------------------------------------

## Hosmer-Lemeshow Goodness-of-Fit Test

The Hosmer-Lemeshow test evaluates whether the logistic regression model adequately fits the data.

-   Null hypothesis: Model fits the data well
-   Large p-value: Evidence of good fit

```{r}
hoslem.test(
  framingham$TenYearCHD,
  fitted(logit_model)
)
```


------------------------------------------------------------------------

## Model Fit Statistics

The following statistics help evaluate overall model quality:

-   AIC: Akaike Information Criterion
-   BIC: Bayesian Information Criterion
-   Log-likelihood: Overall model likelihood

These measures are most useful when comparing multiple models:

-   Lower AIC → better fit (penalizes models that include unnecessary predictors)
-   Lower BIC → better fit (BIC tends to favor simpler models compared to AIC)
-   Higher (less negative) log-likelihood → better fit

```{r}
cat("\nAIC:\n")
print(AIC(logit_model))

cat("\nBIC:\n")
print(BIC(logit_model))

cat("\nLog-Likelihood:\n")
print(logLik(logit_model))
```

------------------------------------------------------------------------

# Example 2: Donner Party Survival Dataset

## Create the Dataset

This dataset contains:
-   `Age`: Age of individual
-   `Sex`:
    -   `0 = Female`
    -   `1 = Male`
-   `Survivor`:
    -   `0 = Did not survive`
    -   `1 = Survived`

```{r}
donner <- data.frame(
  Age = c(23,40,40,30,28,40,45,62,65,45,
          25,28,28,23,22,23,28,15,47,57,
          20,18,25,60,25,20,32,32,24,30,
          15,50,21,25,46,32,30,25,25,25,
          30,35,23,24,25),

  Sex = c(1,0,1,1,1,1,0,1,1,0,
          0,1,1,1,0,0,1,0,0,1,
          0,1,1,1,1,1,1,0,0,1,
          1,0,0,1,1,0,1,1,1,1,
          1,1,1,0,1),

  Survivor = c(0,1,1,0,0,0,0,0,0,0,
               0,1,0,0,1,1,1,1,0,0,
               1,1,0,0,1,1,1,1,1,1,
               0,0,1,0,1,1,0,0,0,0,
               0,0,1,0,1)
)

head(donner)
```

------------------------------------------------------------------------

## Logistic Regression Using Age

This model predicts survival probability using age.

```{r}
model_age <- glm(
  Survivor ~ Age,
  data = donner,
  family = binomial(link = "logit")
)

summary(model_age)
```

------------------------------------------------------------------------

## Odds Ratios and Confidence Intervals for Age Model

```{r}
cat("\nOdds Ratios (Age Model):\n")
print(exp(coef(model_age)))

cat("\n95% Confidence Intervals (Age Model):\n")
print(exp(confint(model_age)))
```

------------------------------------------------------------------------

## Deviance Residuals and Goodness-of-Fit for Age Model

```{r}

dev_resid_age <- residuals(
  model_age,
  type = "deviance"
)


plot(
  dev_resid_age,
  ylab = "Deviance Residuals",
  main = "Deviance Residual Plot",
  pch = 19,
  col = "darkblue"
)

abline(h = 0, lty = 2, col = "red")

hoslem.test(
  donner$Survivor,
  fitted(model_age)
)
```


------------------------------------------------------------------------

## Logistic Regression Using Sex

This model predicts survival probability using sex.

```{r}
model_sex <- glm(
  Survivor ~ Sex,
  data = donner,
  family = binomial(link = "logit")
)

summary(model_sex)
```

------------------------------------------------------------------------

## Odds Ratios and Confidence Intervals for Sex Model

```{r}
cat("\nOdds Ratios (Sex Model):\n")
print(exp(coef(model_sex)))

cat("\n95% Confidence Intervals (Sex Model):\n")
print(exp(confint(model_sex)))
```

## Deviance Residuals for Sex Model

```{r}

dev_resid_sex <- residuals(
  model_sex,
  type = "deviance"
)


plot(
  dev_resid_sex,
  ylab = "Deviance Residuals",
  main = "Deviance Residual Plot",
  pch = 19,
  col = "darkblue"
)

abline(h = 0, lty = 2, col = "red")


```

------------------------------------------------------------------------

## Generalized Linear Model

This final model demonstrates a generalized linear model.

```{r}
model_glm <- glm(
  Survivor ~ Sex,
  data = donner,
  family = binomial(link = "logit")
)

cat("\nGeneralized Linear Model Summary:\n")
summary(model_glm)

cat("\nAIC:\n")
print(AIC(model_glm))

cat("\nBIC:\n")
print(BIC(model_glm))

cat("\nLog-Likelihood:\n")
print(logLik(model_glm))
```

------------------------------------------------------------------------

# Conclusion

These examples demonstrated several applications of logistic regression in R, including:

-   Binary outcome modeling
-   Odds ratio interpretation
-   Confidence interval estimation
-   Model diagnostics
-   Goodness-of-fit testing
-   Logistic regression visualization

These methods are commonly used in biostatistics, epidemiology, social sciences, and machine learning applications involving binary outcomes.
