---
title: "Day 8 (PM) Lab: Logistic Regression"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This lab puts the Day 8 morning material into practice. You will fit, interpret,
and compare logistic regression models on the Framingham Heart Study data.

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** Follow along as the instructor works
  through a complete simple logistic regression with a different predictor than
  the morning example.
- **Part 2 (on your own, ~90 min):** Five exercises of increasing difficulty.
  Create a new `.qmd` file, work through each exercise, render when done.

The **Quick Reference Card** in Section 3 has the key functions.

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```

## Step 1: Check Variable Form

We want to predict CHD risk (`TenYearCHD`) from age. First, inspect the
empirical CHD probability across age.

```{r}
#| message: false
framingham |>
  mutate(CHD_numeric = as.numeric(TenYearCHD) - 1) |>
  ggplot(aes(x = age, y = CHD_numeric)) +
  geom_jitter(alpha = 0.06, height = 0.04, color = "steelblue") +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE, linewidth = 1) +
  labs(x = "Age (years)",
       y = "Probability of CHD within 10 years",
       title = "Empirical CHD Probability by Age") +
  theme_minimal()
```

The loess curve rises steadily. A logistic regression with age as a linear term
on the log-odds scale looks appropriate.

## Step 2: Fit the Model and Interpret

```{r}
chd_age <- glm(TenYearCHD ~ age,
               data = framingham,
               family = binomial)

summary(chd_age)
```

The coefficient on `age` is in log-odds units. To get the odds ratio:

```{r}
exp(cbind(OR = coef(chd_age), confint(chd_age)))
```

Each additional year of age multiplies the odds of CHD by about 1.07. Rescaled
to a decade:

```{r}
exp(10 * coef(chd_age)["age"])
exp(10 * confint(chd_age)["age", ])
```

A 10-year age difference is associated with roughly a 2-fold increase in the
odds of CHD.

## Step 3: Visualize the Fitted Curve

```{r}
#| message: false
framingham |>
  mutate(CHD_numeric = as.numeric(TenYearCHD) - 1) |>
  ggplot(aes(x = age, y = CHD_numeric)) +
  geom_jitter(alpha = 0.06, height = 0.04, color = "steelblue") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "firebrick", se = TRUE, linewidth = 1) +
  labs(x = "Age (years)",
       y = "Probability of CHD within 10 years",
       title = "Fitted Logistic Regression: CHD on Age") +
  theme_minimal()
```

## Step 4: Predicted Probabilities

```{r}
new_ages <- tibble(age = c(35, 45, 55, 65))
new_ages |>
  mutate(p_chd = predict(chd_age, newdata = new_ages, type = "response"))
```

Predicted probability of CHD rises from about 7% at age 35 to about 28% at
age 65.

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document: File -> New File -> Quarto Document.
Title it `"Day 8 Lab: Logistic Regression"`, select HTML, click **Create**.

**Step 2.** Replace the template content with this YAML and setup chunk:

````markdown
---
title: "Day 8 Lab: Logistic Regression"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{{r}}
library(tidyverse)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```
````

**Step 3.** Work through Exercises 1--5 in order. For each one, add a code
chunk, run your code, and write 2--3 sentences interpreting the results.

**Step 4.** Render when finished and fix any errors.

::: {.callout-note}
Exercises 1--3 cover the core material. Exercise 4 introduces confounding.
Exercise 5 is a full model-building workflow.
:::

---

**Exercise 1 -- Simple Logistic Regression with a Binary Predictor.**

a. Use `count()` to tabulate `TenYearCHD` by `currentSmoker`. What proportion
   of smokers develop CHD? What proportion of non-smokers do?

b. Fit a simple logistic regression model predicting `TenYearCHD` from
   `currentSmoker`. Report the coefficient from `summary()`.

c. Compute the odds ratio (OR) and its 95% confidence interval using
   `exp(cbind(OR = coef(model), confint(model)))`. Interpret the OR in plain
   language.

d. Confirm that this OR matches what you would get from a 2x2 table:
   ```r
   tab <- table(framingham$currentSmoker, framingham$TenYearCHD)
   tab
   (tab[2,2] * tab[1,1]) / (tab[1,2] * tab[2,1])
   ```

```{r}
# your code here
```

---

**Exercise 2 -- Simple Logistic Regression with a Continuous Predictor.**

a. Use `geom_jitter()` and `geom_smooth(method = "loess")` to inspect the
   relationship between `sysBP` and `TenYearCHD`.

b. Fit a simple logistic regression of `TenYearCHD` on `sysBP`. Report the
   coefficient, OR, and 95% CI.

c. The OR per 1 mmHg is hard to interpret. Compute the OR for a **20 mmHg
   increase** using `exp(20 * coef(model)["sysBP"])` and the corresponding CI.
   Write one sentence interpreting this rescaled OR clinically.

d. Plot the fitted logistic curve using
   `geom_smooth(method = "glm", method.args = list(family = "binomial"))`.

```{r}
# your code here
```

---

**Exercise 3 -- Multiple Logistic Regression and Confounding.**

a. Fit a model predicting `TenYearCHD` from `currentSmoker` alone. Note the OR
   for smoking.

b. Now fit a model predicting `TenYearCHD` from both `currentSmoker` and `age`.
   Compute the **adjusted** OR for smoking (controlling for age).

c. Compare the unadjusted and adjusted ORs for smoking. Did the OR get larger,
   smaller, or stay the same? Why do you think this happened? (Hint: check
   `framingham |> group_by(currentSmoker) |> summarize(mean_age = mean(age))`.)

d. Test whether age is a useful addition to the model using
   `anova(model_smoker, model_smoker_age, test = "LRT")`. Interpret the
   p-value.

```{r}
# your code here
```


---

# Quick Reference Card

## Setup

```r
library(tidyverse)
library(this.path)
setwd(this.dir())
framingham <- read_csv("../../data/framingham.csv")
```

## Checking Variable Form on a Binary Outcome

```r
df |>
  mutate(Y_numeric = as.numeric(Y) - 1) |>
  ggplot(aes(x = predictor, y = Y_numeric)) +
  geom_jitter(alpha = 0.05, height = 0.04) +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE)
```

## Fitting Models

| Function | What it does |
|---|---|
| `glm(y ~ x, data = df, family = binomial)` | Simple logistic regression |
| `glm(y ~ x1 + x2, data = df, family = binomial)` | Multiple logistic regression |
| `summary(model)` | Coefficients, z-tests, deviance |
| `exp(cbind(OR = coef(m), confint(m)))` | OR table in one go |
| `AIC(m1, m2)` | Compare models, lower is better |
| `anova(m_small, m_large, test = "LRT")` | Likelihood ratio test |
| `step(model, direction = "both", trace = 0)` | AIC-based stepwise |
| `predict(model, type = "response")` | Predicted probabilities |
| `drop_na(all_of(vars))` | Complete cases for model comparison |

## Plotting the Fitted Curve

```r
df |>
  mutate(Y_numeric = as.numeric(Y) - 1) |>
  ggplot(aes(x = predictor, y = Y_numeric)) +
  geom_jitter(alpha = 0.05, height = 0.04) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "firebrick", se = TRUE)
```

## Rescaling Odds Ratios to Meaningful Units

```r
exp(10 * coef(model)["age"])           # 10-year OR
exp(20 * confint(model)["sysBP", ])    # 20 mmHg CI
exp(40 * coef(model)["totChol"])       # 40 mg/dL OR
```

## Common Interpretive Phrases

| Quantity | How to say it |
|---|---|
| OR = 1.5 | "1.5-fold higher odds" or "50% higher odds" |
| OR = 2.0 | "double the odds" |
| OR = 0.5 | "half the odds" or "50% lower odds" |
| OR = 1.0 (CI includes 1) | "no evidence of an association" |
| Adjusted OR | "after controlling for [other variables]" |
