---
title: "Day 2 (PM) Lab: Working with Data in R -- Solutions"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Setup

```{r}
library(tidyverse)
library(this.path)

this.dir <- this.dir()
setwd(this.dir)

framingham <- read_csv("../../data/framingham.csv")
```

Variable type corrections (required before all exercises):

```{r}
framingham <- framingham |>
  mutate(
    male            = factor(male,            levels = c(0, 1),
                                              labels = c("Female", "Male")),
    currentSmoker   = factor(currentSmoker,   levels = c(0, 1),
                                              labels = c("Non-smoker", "Smoker")),
    BPMeds          = factor(BPMeds,          levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentStroke = factor(prevalentStroke, levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentHyp    = factor(prevalentHyp,    levels = c(0, 1),
                                              labels = c("No", "Yes")),
    diabetes        = factor(diabetes,        levels = c(0, 1),
                                              labels = c("No", "Yes")),
    TenYearCHD      = factor(TenYearCHD,      levels = c(0, 1),
                                              labels = c("No", "Yes")),
    education       = factor(education,       levels = 1:4,
                                              labels = c("Some HS", "HS Diploma",
                                                         "Some College",
                                                         "College Degree"))
  )
```

---

# Exercise 1

**Cholesterol by diabetes status.**

```{r}
framingham |>
  group_by(diabetes) |>
  summarize(
    n        = n(),
    mean_chol = mean(totChol, na.rm = TRUE),
    sd_chol   = sd(totChol,   na.rm = TRUE),
    med_chol  = median(totChol, na.rm = TRUE),
    .groups  = "drop"
  )
```

Diabetic and non-diabetic participants have similar mean total cholesterol
(roughly 236 vs. 248 mg/dL), with the diabetic group showing a
slightly *higher* mean. The difference is small and does not suggest a strong
univariate association between diabetes status and cholesterol in this
sample.

---

# Exercise 2

**Pulse pressure and BMI category.**



```{r}
framingham <- framingham |>
  mutate(
    pulse_pressure = sysBP - diaBP,
    bmi_category   = case_when(
      BMI < 18.5              ~ "Underweight",
      BMI >= 18.5 & BMI < 25 ~ "Normal",
      BMI >= 25   & BMI < 30 ~ "Overweight",
      BMI >= 30               ~ "Obese",
      .default = NA_character_
    )
  )

framingham |>
  group_by(bmi_category) |>
  summarize(
    n                  = n(),
    mean_pulse_pressure = mean(pulse_pressure, na.rm = TRUE)
  )
```

Mean pulse pressure increases monotonically with BMI category, rising from
about 47 mmHg in the Normal group to about 54 mmHg in the Obese group. This
is consistent with higher arterial stiffness in individuals with greater body
mass.



---

# Exercise 3

**Female non-smokers over age 55: base R vs. dplyr.**

```{r}
# Base R version
subset_baseR <- framingham[
  framingham$male == "Female" &
  framingham$currentSmoker == "Non-smoker" &
  framingham$age > 55,
]
nrow(subset_baseR)
mean(subset_baseR$glucose, na.rm = TRUE)
```

```{r}
# dplyr version
subset_dplyr <- framingham |>
  filter(
    male          == "Female",
    currentSmoker == "Non-smoker",
    age           > 55
  )
nrow(subset_dplyr)
mean(subset_dplyr$glucose, na.rm = TRUE)
```

Both approaches return the same row count. Mean glucose in
this subgroup is approximately 87 mg/dL, which is within the normal fasting
range. The dplyr version is easier to read and extend; the base R version is
useful to know for contexts where tidyverse is unavailable.

---

# Exercise 4

**Cigarette use among current smokers; heavy smoker flag.**

```{r}
# Summary among current smokers
framingham |>
  filter(currentSmoker == "Smoker") |>
  summarize(
    n            = n(),
    mean_cpd     = mean(cigsPerDay, na.rm = TRUE),
    sd_cpd       = sd(cigsPerDay,   na.rm = TRUE),
    min_cpd      = min(cigsPerDay,  na.rm = TRUE),
    max_cpd      = max(cigsPerDay,  na.rm = TRUE),
    .groups      = "drop"
  )
```

```{r}
# Add heavy_smoker to full dataset and count
framingham |>
  mutate(heavy_smoker = cigsPerDay > 20) |> 
  summarise(
    pct_heavy_smk = mean(heavy_smoker, na.rm = TRUE) * 100,
    num_heavy_smk = sum(heavy_smoker, na.rm = TRUE)
         )
```

Among current smokers, the average is roughly 18 cigarettes per day (close to
a pack), with a wide range. About 11% of people are classified as heavy smokers. Note that `cigsPerDay` contains
some missingness, so `na.rm = TRUE` is essential throughout.

---

# Exercise 5

**10-year CHD rate by education level.**

```{r}
framingham |>
  group_by(education) |>
  summarize(
    n        = n(),
    chd_pct  = mean(TenYearCHD == "Yes", na.rm = TRUE) * 100
  ) |>
  filter(!is.na(education)) |> # Remove the education = NA group.
  arrange(desc(chd_pct))
```

CHD rates are highest among participants with the lowest education (Some HS), 
and the lowest 10-year CHD risk is among those with a HS Diploma. 

---

# Exercise 6 (Challenge)

**Multi-step pipeline: ages 40--65, hypertension-smoker interaction, CHD
rates by subgroup.**

```{r}
framingham |>
  filter(age >= 40, age <= 65) |>
  mutate(
    hyp_smoker = if_else(
      prevalentHyp == "Yes" & currentSmoker == "Smoker",
      "Yes", "No"
    )
  ) |>
  group_by(hyp_smoker, male) |>
  summarize(
    n         = n(),
    mean_sysBP = mean(sysBP,  na.rm = TRUE),
    chd_pct   = mean(TenYearCHD == "Yes", na.rm = TRUE) * 100,
    .groups   = "drop"
  ) |>
  filter(n >= 50) |>
  arrange(desc(chd_pct))
```

Among adults aged 40--65, males with both prevalent hypertension and current
smoking (`hyp_smoker == "Yes"`, `male == "Male"`) show the highest 10-year CHD
rate (31.4%). The combination of hypertension and smoking acts synergistically: each
risk factor alone elevates risk, but their co-occurrence is associated with
substantially higher rates than either alone. 

There are no groups with fewer than 50 participants, so the final filtering
does not have any effect.

