---
title: "Day 9 (PM) Lab: Classification and Predictive Performance"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This lab puts the Day 9 morning material into practice. You will turn predicted
probabilities into classifications, build confusion matrices, compute
performance metrics, construct ROC curves, and compare models.

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** Follow along as the instructor walks
  through a complete classification evaluation using a simple logistic model.
- **Part 2 (on your own, ~90 min):** Six exercises of increasing difficulty.
  Create a new `.qmd` file, work through each exercise, render when done.

The **Quick Reference Card** in Section 3 has the key functions.

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(this.path)
library(pROC)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```

## Step 1: Fit a Model and Get Predicted Probabilities

We re-fit a simple logistic model predicting CHD from `sysBP` and add predicted
probabilities to the dataset.

```{r}
model_bp <- glm(TenYearCHD ~ sysBP,
                data = framingham,
                family = binomial)

framingham <- framingham |>
  mutate(p_bp = predict(model_bp, type = "response"))

framingham |>
  select(TenYearCHD, sysBP, p_bp) |>
  summary()
```

## Step 2: Classify at Two Thresholds

```{r}
# Threshold = 0.50
framingham <- framingham |>
  mutate(pred_50 = factor(if_else(p_bp > 0.50, "Yes", "No"),
                          levels = c("No", "Yes")))

table(Predicted = framingham$pred_50, Actual = framingham$TenYearCHD)
```

Almost nobody gets classified as positive at 0.50. Now try 0.15:

```{r}
framingham <- framingham |>
  mutate(pred_15 = factor(if_else(p_bp > 0.15, "Yes", "No"),
                          levels = c("No", "Yes")))

tab_15 <- table(Predicted = framingham$pred_15, Actual = framingham$TenYearCHD)
tab_15
```

## Step 3: Compute Sensitivity and Specificity

```{r}
TP <- tab_15["Yes", "Yes"]
FP <- tab_15["Yes", "No"]
TN <- tab_15["No",  "No"]
FN <- tab_15["No",  "Yes"]

tibble(
  Sensitivity = TP / (TP + FN),
  Specificity = TN / (TN + FP),
  PPV         = TP / (TP + FP),
  NPV         = TN / (TN + FN)
) |> round(3)
```

## Step 4: ROC Curve and AUC

```{r}
#| fig-width: 5
#| fig-height: 5
roc_bp <- roc(framingham$TenYearCHD, framingham$p_bp,
              levels = c("No", "Yes"), direction = "<")

plot(roc_bp, col = "firebrick", lwd = 2,
     main = "ROC Curve: CHD ~ sysBP")
auc(roc_bp)
```

The AUC for sysBP alone is modest. Adding more predictors (tomorrow's model
comparison) should improve this.

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document. Title it
`"Day 9 Lab: Classification and Predictive Performance"`.

**Step 2.** Replace the template content with this YAML and setup chunk:

````markdown
---
title: "Day 9 Lab: Classification and Predictive Performance"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{{r}}
library(tidyverse)
library(this.path)
library(pROC)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```
````

**Step 3.** Work through Exercises 1--6 in order.

**Step 4.** Render when finished.

::: {.callout-note}
Exercises 1--4 cover the core material. Exercise 5 compares two models
head-to-head. Exercise 6 is a challenge on honest evaluation.
:::

---

**Exercise 1 -- Predicted Probabilities from Two Models.**

a. Restrict to complete cases on `TenYearCHD`, `age`, `currentSmoker`, `sysBP`,
   `totChol`, and `diabetes` using `drop_na()`.

b. Fit two logistic regression models:
   - `model_simple`: `TenYearCHD ~ age`
   - `model_full`: `TenYearCHD ~ age + currentSmoker + sysBP + totChol + diabetes`

c. Add predicted probabilities from both models to the dataset as `p_simple`
   and `p_full` using `predict(..., type = "response")`.

d. Use `summary()` on both columns. What is the range of predicted probabilities
   for each model? Which model produces a wider spread?

```{r}
# your code here
```

---

**Exercise 2 -- Confusion Matrices at Different Thresholds.**

Using the `p_full` probabilities from Exercise 1:

a. Create classifications at three thresholds: 0.10, 0.20, and 0.50, using
   `if_else(p_full > threshold, "Yes", "No")` wrapped in `factor()`.

b. Build a confusion matrix at each threshold with
   `table(Predicted = ..., Actual = ...)`.

c. As the threshold increases from 0.10 to 0.50, what happens to the number
   of predicted positives? What about predicted negatives?

```{r}
# your code here
```

---

**Exercise 3 -- Computing Sensitivity, Specificity, PPV, NPV.**

a. For each of the three confusion matrices from Exercise 2, compute
   sensitivity, specificity, PPV, and NPV by hand. Extract TP, FP, TN, FN
   from the confusion matrix, then compute:
   ```r
   Sensitivity = TP / (TP + FN)
   Specificity = TN / (TN + FP)
   PPV         = TP / (TP + FP)
   NPV         = TN / (TN + FN)
   ```

b. Which threshold gives the highest sensitivity? Which gives the highest
   specificity? Can you have both at the same time?


```{r}
# your code here
```

---

**Exercise 4 -- ROC Curve and AUC.**

a. Use `roc()` from the `pROC` package to build an ROC curve for the full model:
   ```r
   roc_full <- roc(your_data$TenYearCHD, your_data$p_full,
                   levels = c("No", "Yes"), direction = "<")
   ```

b. Plot the ROC curve with `plot(roc_full, col = "firebrick", lwd = 2)`.
   Describe the shape of the curve. Does it bow far from the diagonal?

c. Compute the AUC with `auc(roc_full)`. Using the interpretation table from
   the morning lecture (0.5 = coin flip, 0.7--0.8 = acceptable, 0.8--0.9 =
   excellent, >0.9 = outstanding), how would you describe this model?


```{r}
# your code here
```


---

**Exercise 5 (Challenge) -- Honest AUC via Train/Test Split.**

The AUC values in Exercise 4 were computed on the same data used to fit the
models. This overestimates how well the model will perform on new data.

a. Set a random seed with `set.seed(2026)` for reproducibility. Split your
   complete-cases dataset into 70% training and 30% test:
   ```r
   n <- nrow(your_data)
   train_idx <- sample(seq_len(n), size = round(0.7 * n))
   train <- your_data[train_idx, ]
   test  <- your_data[-train_idx, ]
   ```

b. Fit the full model on the **training** data only.

c. Use `predict()` to get predicted probabilities on the **test** data.

d. Compute the training AUC and the test AUC. Is the test AUC lower? By how
   much?

e. **Discussion.** Why is the test AUC a more honest measure of model
   performance? If the gap between training and test AUC were very large
   (e.g., 0.85 vs. 0.65), what would that suggest about the model?

```{r}
# your code here
```

---

# Quick Reference Card

## Setup

```r
library(tidyverse)
library(this.path)
library(pROC)
setwd(this.dir())
framingham <- read_csv("../../data/framingham.csv")
```

## Predicted Probabilities

```r
# Add fitted probabilities to the dataset
df <- df |> mutate(p_hat = predict(model, type = "response"))

# Predict for new individuals
predict(model, newdata = new_data, type = "response")
```

## Classification at a Threshold

```r
df <- df |>
  mutate(pred_class = factor(if_else(p_hat > 0.15, "Yes", "No"),
                             levels = c("No", "Yes")))

tab <- table(Predicted = df$pred_class, Actual = df$TenYearCHD)
tab
```

## Computing Metrics from a Confusion Matrix

```r
TP <- tab["Yes", "Yes"]
FP <- tab["Yes", "No"]
TN <- tab["No",  "No"]
FN <- tab["No",  "Yes"]

Sensitivity <- TP / (TP + FN)
Specificity <- TN / (TN + FP)
PPV         <- TP / (TP + FP)
NPV         <- TN / (TN + FN)
```

## ROC Curves and AUC

```r
# Build ROC curve
roc_obj <- roc(df$Y, df$p_hat, levels = c("No", "Yes"), direction = "<")

# Plot
plot(roc_obj, col = "firebrick", lwd = 2)

# AUC
auc(roc_obj)

# Best threshold by Youden's J
coords(roc_obj, "best", best.method = "youden")

# Compare two ROC curves
plot(roc1, col = "firebrick", lwd = 2)
plot(roc2, col = "steelblue", lwd = 2, add = TRUE)

# Test whether AUCs differ significantly
roc.test(roc1, roc2)
```

## Train/Test Split

```r
set.seed(2026)
n <- nrow(df)
train_idx <- sample(seq_len(n), size = round(0.7 * n))
train <- df[train_idx, ]
test  <- df[-train_idx, ]

# Fit on training, predict on test
model_train <- glm(Y ~ x1 + x2, data = train, family = binomial)
test$p_hat  <- predict(model_train, newdata = test, type = "response")

# AUC on test
roc_test <- roc(test$Y, test$p_hat, levels = c("No", "Yes"), direction = "<")
auc(roc_test)
```

## AUC Interpretation

| AUC | Discriminative ability |
|---|---|
| $\approx$ 0.5 | No discrimination (coin flip) |
| 0.6 -- 0.7 | Poor |
| 0.7 -- 0.8 | Acceptable |
| 0.8 -- 0.9 | Excellent |
| > 0.9 | Outstanding |
