---
title: "Day 9 (PM) Lab: Classification and Predictive Performance -- Solutions"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(this.path)
library(pROC)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```

---

# Exercise 1 -- Predicted Probabilities from Two Models

## Part a: Restrict to complete cases

```{r}
vars_used <- c("TenYearCHD", "age", "currentSmoker", "sysBP",
               "totChol", "diabetes")

framingham_cc <- framingham |>
  drop_na(all_of(vars_used))

nrow(framingham) - nrow(framingham_cc)
```

A small number of rows are dropped due to missing values on one or more of the
variables. All subsequent exercises use `framingham_cc` so that both models are
always fit on the same observations.

## Part b: Fit two models

```{r}
model_simple <- glm(TenYearCHD ~ age,
                    data = framingham_cc,
                    family = binomial)

model_full <- glm(TenYearCHD ~ age + currentSmoker + sysBP + totChol + diabetes,
                  data = framingham_cc,
                  family = binomial)
```

## Part c: Add predicted probabilities

```{r}
framingham_cc <- framingham_cc |>
  mutate(
    p_simple = predict(model_simple, type = "response"),
    p_full   = predict(model_full,   type = "response")
  )
```

## Part d: Compare the spread of predicted probabilities

```{r}
framingham_cc |>
  select(p_simple, p_full) |>
  summary()
```

Both sets of predicted probabilities are bounded between 0 and 1 and centered
near the overall event rate (about 15%). The full model produces a wider spread
than the simple model: its minimum is lower and its maximum is higher. This
makes intuitive sense because the full model uses more information and can more
confidently assign very low risk to healthy-profile participants and very high
risk to high-risk participants. A model with no predictive ability at all would
output the same probability (the event rate, about 0.15) for everyone. Wider
spread signals that the model is doing more discriminative work.

---

# Exercise 2 -- Confusion Matrices at Different Thresholds

## Part a and b: Classifications and confusion matrices at three thresholds

```{r}
framingham_cc <- framingham_cc |>
  mutate(
    pred_10 = factor(if_else(p_full > 0.10, "Yes", "No"), levels = c("No","Yes")),
    pred_20 = factor(if_else(p_full > 0.20, "Yes", "No"), levels = c("No","Yes")),
    pred_50 = factor(if_else(p_full > 0.50, "Yes", "No"), levels = c("No","Yes"))
  )

tab_10 <- table(Predicted = framingham_cc$pred_10, Actual = framingham_cc$TenYearCHD)
tab_20 <- table(Predicted = framingham_cc$pred_20, Actual = framingham_cc$TenYearCHD)
tab_50 <- table(Predicted = framingham_cc$pred_50, Actual = framingham_cc$TenYearCHD)

cat("Threshold = 0.10\n"); tab_10
cat("\nThreshold = 0.20\n"); tab_20
cat("\nThreshold = 0.50\n"); tab_50
```

## Part c: What happens as the threshold increases?

As the threshold increases from 0.10 to 0.50, the number of predicted positives
decreases substantially and the number of predicted negatives increases. At
0.10, we flag a large fraction of the cohort as predicted CHD cases. At 0.50,
almost nobody clears the threshold because only about 15% of participants
actually develop CHD and most predicted probabilities are well below 0.50. The
total number of observations stays the same; we are simply re-drawing the line
between "positive" and "negative" classifications.

---

# Exercise 3 -- Computing Sensitivity, Specificity, PPV, NPV

## Part a: Compute metrics at each threshold

```{r}
compute_metrics <- function(tab) {
  TP <- tab["Yes", "Yes"]
  FP <- tab["Yes", "No"]
  TN <- tab["No",  "No"]
  FN <- tab["No",  "Yes"]
  tibble(
    Sensitivity = TP / (TP + FN),
    Specificity = TN / (TN + FP),
    PPV         = TP / (TP + FP),
    NPV         = TN / (TN + FN)
  )
}

bind_rows(
  compute_metrics(tab_10) |> mutate(Threshold = 0.10),
  compute_metrics(tab_20) |> mutate(Threshold = 0.20),
  compute_metrics(tab_50) |> mutate(Threshold = 0.50)
) |>
  select(Threshold, everything()) |>
  round(3)
```

## Part b: Which threshold gives the highest sensitivity? Highest specificity?

The lowest threshold (0.10) gives the highest sensitivity: we flag so many
people as positive that we catch most of the true CHD cases, but at the cost
of many false alarms (low specificity). The highest threshold (0.50) gives the
highest specificity: almost everyone is classified as negative, so we rarely
accuse a healthy person of having CHD, but we miss nearly all the true cases
(near-zero sensitivity). Sensitivity and specificity move in opposite directions
as the threshold changes. You cannot maximize both simultaneously with a fixed
model. Improving one always comes at the expense of the other, which is why the
ROC curve is useful: it shows the full tradeoff across every possible threshold
rather than locking you into one.

## Part c: Why is PPV low at every threshold?

PPV is low because CHD is relatively rare in this cohort (about 15% event rate).
Even a model with good sensitivity will generate many false positives when it
flags a large portion of the population, because most of that population is
truly healthy. This is the base-rate problem: if only 1 in 7 people has the
condition and the model flags 1 in 3 as positive, at least half of those
flagged will be false positives even if the model is working well. PPV is not
just a property of the model; it depends on the prevalence of the condition
in the population being tested.

---

# Exercise 4 -- ROC Curve and AUC

## Part a: Build the ROC curve

```{r}
roc_full <- roc(framingham_cc$TenYearCHD, framingham_cc$p_full,
                levels = c("No", "Yes"), direction = "<")
```

## Part b: Plot and describe

```{r}
#| fig-width: 5
#| fig-height: 5
plot(roc_full,
     col  = "firebrick",
     lwd  = 2,
     main = "ROC Curve: Full Logistic Model")
abline(a = 0, b = 1, lty = 2, col = "gray60")
```

The curve bows noticeably above the diagonal, which represents a coin-flip
classifier with no predictive value. The curve rises steeply on the left side,
indicating that the model achieves reasonably high sensitivity before
accumulating many false positives. It does not hug the top-left corner tightly,
which is typical for clinical risk models predicting a complex outcome like CHD:
the event is influenced by many unmeasured factors, so no short set of
predictors can achieve near-perfect discrimination.

## Part c: AUC and interpretation

```{r}
auc(roc_full)
```

The AUC is approximately 0.72--0.75 (exact value updates on render). By the
interpretation table from the morning lecture, this places the model in the
**acceptable** range (0.70--0.80). The model discriminates between CHD cases
and non-cases meaningfully, but it is far from perfect. This is a reasonable
result for a five-predictor model predicting a multifactorial cardiovascular
outcome in a general population cohort.

---

# Exercise 5 (Challenge) -- Honest AUC via Train/Test Split

## Parts a--c: Split, fit, and predict

```{r}
set.seed(2026)
n         <- nrow(framingham_cc)
train_idx <- sample(seq_len(n), size = round(0.7 * n))
train     <- framingham_cc[train_idx, ]
test      <- framingham_cc[-train_idx, ]

cat("Training rows:", nrow(train), "\n")
cat("Test rows:    ", nrow(test),  "\n")
```

```{r}
model_train <- glm(TenYearCHD ~ age + currentSmoker + sysBP + totChol + diabetes,
                   data  = train,
                   family = binomial)

test <- test |>
  mutate(p_test = predict(model_train, newdata = test, type = "response"))
```

## Part d: Training AUC vs. test AUC

```{r}
roc_train <- roc(train$TenYearCHD,
                 predict(model_train, type = "response"),
                 levels = c("No","Yes"), direction = "<")

roc_test <- roc(test$TenYearCHD, test$p_test,
                levels = c("No","Yes"), direction = "<")

tibble(
  Dataset = c("Training", "Test"),
  N       = c(nrow(train), nrow(test)),
  AUC     = c(as.numeric(auc(roc_train)),
              as.numeric(auc(roc_test)))
)
```

```{r}
#| fig-width: 5
#| fig-height: 5
plot(roc_train, col = "steelblue", lwd = 2,
     main = "Training vs. Test ROC Curves")
plot(roc_test,  col = "firebrick", lwd = 2, add = TRUE)
legend("bottomright",
       legend = c(paste0("Training (AUC = ", round(auc(roc_train), 3), ")"),
                  paste0("Test     (AUC = ", round(auc(roc_test),  3), ")")),
       col = c("steelblue", "firebrick"), lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "gray60")
```

The test AUC is typically a few points lower than the training AUC. The gap
here is small (often less than 0.02), which suggests the model is not badly
over-fitting. With only five predictors and over 4,000 observations, there is
not much room for the model to memorize noise in the training data.

## Part e: Discussion

The test AUC is a more honest estimate because the model has never seen the test
observations during fitting. When we evaluate AUC on the training data, the
model has already been optimized to fit those exact observations, so it
inevitably looks a little better than it really is. The test set breaks this
feedback loop by holding out data the model cannot cheat on.

If the gap between training and test AUC were very large (for example, 0.85 vs.
0.65), that would signal severe over-fitting: the model has learned
idiosyncrasies of the training data that do not generalize. This is most likely
to happen when the model has many predictors relative to the number of events,
or when variable selection was performed on the training data. In such cases,
cross-validation (Day 7) provides a more stable estimate than a single
train/test split, because it averages over multiple test sets rather than
depending on one lucky or unlucky random split.
