---
title: "Day 6 (PM) Lab: Linear Regression"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This lab puts the Day 6 morning material into practice. You will work through
the full linear regression workflow — from checking variable form to fitting,
interpreting, and diagnosing models — using the Framingham Heart Study dataset
you already know.

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** Follow along as the instructor works
  through a complete simple regression example using a new outcome variable.
- **Part 2 (on your own, ~90 min):** Six exercises of increasing difficulty.
  Create a new `.qmd` file, work through each exercise, render when done.

The **Quick Reference Card** in Section 3 has the key functions.

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```

## Step 1: Check Variable Form

We want to predict total cholesterol (`totChol`) from BMI. First, inspect the
relationship visually with a loess smoother.

```{r}
#| message: false
ggplot(framingham, aes(x = BMI, y = totChol)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE, linewidth = 1) +
  labs(x = "BMI", y = "Total Cholesterol (mg/dL)",
       title = "Total Cholesterol vs. BMI with Loess Smoother") +
  theme_minimal()
```

The loess curve is approximately linear — a linear model is reasonable here.

## Step 2: Fit the Model and Interpret

```{r}
chol_model <- lm(totChol ~ BMI, data = framingham)
summary(chol_model)
confint(chol_model)
```

Key take-aways from the output:

- The slope tells us how much cholesterol changes per one-unit increase in BMI,
  on average.
- R² tells us the proportion of cholesterol variability explained by BMI alone.
- The confidence interval for the slope tells us the range of plausible values
  for the true population effect.

## Step 3: Plot the Regression Line

```{r}
#| message: false
ggplot(framingham, aes(x = BMI, y = totChol)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "lm", color = "firebrick", se = TRUE, linewidth = 1) +
  labs(x = "BMI", y = "Total Cholesterol (mg/dL)",
       title = "Simple Linear Regression: Total Cholesterol on BMI",
       subtitle = "Shaded band is 95% confidence interval for the mean") +
  theme_minimal()
```

## Step 4: Diagnostics

```{r}
#| fig-width: 8
#| fig-height: 7
par(mfrow = c(2, 2))
plot(chol_model)
par(mfrow = c(1, 1))
```

```{r}
#| fig-width: 7
#| fig-height: 4
cooks_df <- tibble(
  obs   = seq_along(cooks.distance(chol_model)),
  cooks = cooks.distance(chol_model)
)

ggplot(cooks_df, aes(x = obs, y = cooks)) +
  geom_segment(aes(xend = obs, yend = 0), color = "steelblue", alpha = 0.6) +
  geom_point(data = filter(cooks_df, cooks > quantile(cooks, 0.99)),
             color = "firebrick", size = 2) +
  labs(x = "Observation index", y = "Cook's distance",
       title = "Cook's Distance — Cholesterol on BMI") +
  theme_minimal()
```

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document: File → New File → Quarto Document.
Title it `"Day 6 Lab: Linear Regression"`, select HTML, click **Create**.

**Step 2.** Replace the template content with this YAML and setup chunk:

````markdown
---
title: "Day 6 Lab: Linear Regression"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{{r}}
library(tidyverse)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```
````

**Step 3.** Work through Exercises 1--6 in order. For each one:

- Add a new code chunk (`Ctrl+Alt+I` / `Cmd+Option+I`)
- Write and run your code
- Add 2--3 sentences below each result describing what you found in plain
  language

**Step 4.** Render (`Ctrl+Shift+K` / `Cmd+Shift+K`) when finished and fix
any errors before you are done.

::: {.callout-tip}
**Before asking for help**, check two things: (1) the **Quick Reference Card**
in Section 3, and (2) whether you have `na.rm = TRUE` in any summary functions
that precede your model — `lm()` drops missing values automatically, but
`mean()` and `sd()` do not.
:::

::: {.callout-note}
Exercises 1--4 cover the core material. Exercise 5 requires putting together
everything from the morning into a multi-step analysis. Exercise 6 is a
challenge --- don't worry if you don't reach it.
:::

---

**Exercise 1 — Simple Linear Regression: Diastolic BP and Age.**

a. Use `ggplot()` with `geom_smooth(method = "loess")` to inspect the
   relationship between age (`age`) and diastolic blood pressure (`diaBP`).
   Does the relationship look linear?

b. Fit a simple linear regression model predicting `diaBP` from `age`.
   Write out the fitted regression equation using the estimated coefficients.

c. Interpret the slope in plain language: what does it tell you about the
   relationship between age and diastolic blood pressure in this population?

d. How much of the variability in diastolic BP does age alone explain? Is this
   more or less than the R² for the `sysBP ~ age` model from the morning session
   (R² = 15.5%)?

e. Plot the fitted regression line with a 95% confidence band using
   `geom_smooth(method = "lm")`.

```{r}
# your code here
```

---

**Exercise 2 — Model Diagnostics.**

Using the `diaBP ~ age` model from Exercise 1:

a. Produce the four standard diagnostic plots using `par(mfrow = c(2,2))` and
   `plot(model)`. Describe what you see in each panel. Are any model assumptions
   clearly violated?

b. Produce a Cook's distance plot using `ggplot()` with `geom_segment()`.
   Flag the top 1% of observations as red points. Are there any observations
   that stand clearly apart from the rest?

c. Report the 95% confidence interval for the slope using `confint()`. What
   does this interval tell you about the precision of the estimate?

```{r}
# your code here
```

---

**Exercise 3 — Continuous vs. Categorical Predictors.**

a. Create a new variable called `bmi_category` in the Framingham data using
   `mutate()` and `cut()` with three groups: `"<25"` (BMI < 25),
   `"25-30"` (25 ≤ BMI < 30), and `"30+"` (BMI ≥ 30).

b. Fit two models predicting `sysBP`: one using `BMI` as continuous, one using
   `bmi_category` as a factor. Compare their AIC values using `AIC()`. Which
   model fits the data better?

c. Report the R² of both models. What does this comparison tell you about the
   cost of categorizing a continuous predictor?

d. Plot the predicted values from the categorical model on a scatter plot of
   BMI vs. sysBP. You can use `predict()` and `geom_step()` or overlay group
   means using `stat_summary()`. What do the step-function predictions reveal
   about the limitations of categorization?

```{r}
# your code here
```

---

**Exercise 4 — Multiple Regression: Checking Variable Form.**

Before building a multiple regression model for `sysBP`, inspect the
relationship between each continuous predictor and `sysBP` using a loess
smoother.

a. Use `pivot_longer()` and `facet_wrap()` to create a single faceted plot
   showing `sysBP` on the y-axis against each of the following predictors:
   `age`, `BMI`, `totChol`, `diaBP`, `cigsPerDay`, `glucose`. Use
   `geom_smooth(method = "loess")`.

b. Based on the plots, which predictors appear to have a roughly linear
   relationship with `sysBP`? Are there any that show obvious non-linearity
   or very weak relationships?

c. Note: `cigsPerDay` and `glucose` have missing values. Does this affect
   the plot? Check `colSums(is.na(framingham[, c("cigsPerDay","glucose")]))`.

```{r}
# your code here
```

---

**Exercise 5 — Multiple Regression: Full Workflow.**

Build a multiple regression model predicting `sysBP` from a set of continuous
predictors.

a. Start with a full model including: `age`, `BMI`, `totChol`, `diaBP`,
   `cigsPerDay`, and `glucose`. Fit it with `lm()` and examine `summary()`.
   Which predictors appear most strongly associated with `sysBP`?

b. Use `step(model_full, direction = "both", trace = 0)` to perform AIC-based
   model selection. Compare the AIC of the full and selected models using
   `AIC()`. What variables were retained?

c. For any pair of nested models you want to formally compare, use `anova()`
   to run an F-test. Interpret the p-value: does the larger model explain
   significantly more variance?

d. Report and interpret the coefficients of your final model. For each
   predictor retained, write one sentence explaining what the coefficient means
   in clinical terms.

e. Run the four diagnostic plots on your final model. Note any concerns.
   Produce a Cook's distance plot and label any flagged observations.

f. Report the 95% confidence intervals for all coefficients in the final model
   using `confint()`.

```{r}
# your code here
```

---

**Exercise 6 (Challenge) — Non-Linear Relationships in Framingham.**

The relationship between `glucose` and `sysBP` in the Framingham data is worth
examining carefully. Glucose has a skewed distribution and its relationship with
blood pressure may not be linear.

a. Remove missing values from `glucose` and `sysBP` first:
   ```r
   fram_gluc <- framingham |> filter(!is.na(glucose), !is.na(sysBP))
   ```
   How many observations remain?

b. Plot `sysBP ~ glucose` with a loess smoother. Describe the shape of the
   relationship. Does the loess curve bend, or is it roughly straight? Note
   that glucose has a few very high values --- are these affecting the smoother?

c. Fit three competing models predicting `sysBP` from `glucose`:
   - `m_linear`: `sysBP ~ glucose`
   - `m_quad`: `sysBP ~ poly(glucose, 2)`
   - `m_spline`: `sysBP ~ splines::ns(glucose, df = 3)`

   Compare all three with `AIC()`. Which fits best?

d. To visualize the three fitted curves on the same plot, create a prediction
   grid and use `predict()` for each model:
   ```r
   gluc_grid <- tibble(glucose = seq(min(fram_gluc$glucose),
                                     max(fram_gluc$glucose),
                                     length.out = 300))
   ```
   Then use `pivot_longer()` to reshape the predictions and plot them with
   different colors using `geom_line()`.

e. Now build a **multiple regression** model for `sysBP` that includes `age`,
   `BMI`, `diaBP`, and `glucose` — but allow `glucose` to enter as a spline
   term (`splines::ns(glucose, df = 3)`) based on what you found above. Compare
   its AIC to an equivalent model with `glucose` entered as linear. Is the
   non-linear glucose term worth including once the other predictors are
   accounted for?

f. Run diagnostics on the better-fitting model. Does allowing for non-linearity
   in glucose improve the residual plots compared to the linear version?

```{r}
# your code here
```

---

# Quick Reference Card

## Setup and Reading Data

```r
library(tidyverse)
library(this.path)
setwd(this.dir())
framingham <- read_csv("../../data/framingham.csv")
```

## Checking Variable Form

```r
# Single predictor vs. outcome
ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE)

# All predictors at once
df |>
  pivot_longer(cols = c(x1, x2, x3), names_to = "predictor", values_to = "value") |>
  ggplot(aes(x = value, y = outcome)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", color = "firebrick") +
  facet_wrap(~ predictor, scales = "free_x")
```

## Fitting Models

| Function | What it does |
|---|---|
| `lm(y ~ x, data = df)` | Simple linear regression |
| `lm(y ~ x1 + x2 + x3, data = df)` | Multiple linear regression |
| `lm(y ~ poly(x, 2), data = df)` | Quadratic polynomial |
| `lm(y ~ splines::ns(x, df = 3), data = df)` | Natural spline |
| `lm(y ~ factor(x), data = df)` | Categorical predictor |
| `summary(model)` | Coefficients, R², F-test |
| `confint(model)` | 95% CIs for all coefficients |
| `AIC(model1, model2)` | Compare models — lower is better |
| `anova(model_small, model_large)` | F-test for nested models |
| `step(model, direction = "both", trace = 0)` | AIC-based stepwise selection |

## Plotting the Regression Line

```r
ggplot(df, aes(x = predictor, y = outcome)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", color = "firebrick", se = TRUE)
```

## Diagnostics

```r
# Four standard diagnostic plots
par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))

# Cook's distance with ggplot
tibble(obs = seq_along(cooks.distance(model)),
       cooks = cooks.distance(model)) |>
  ggplot(aes(x = obs, y = cooks)) +
  geom_segment(aes(xend = obs, yend = 0), color = "steelblue") +
  geom_point(data = ~ filter(., cooks > quantile(cooks, 0.99)),
             color = "firebrick", size = 2)
```

## Prediction

```r
# New data to predict for
new_data <- tibble(age = c(45, 55), BMI = c(25, 30))

# Confidence interval for the mean response
predict(model, newdata = new_data, interval = "confidence")

# Prediction interval for individual observations (always wider)
predict(model, newdata = new_data, interval = "prediction")
```

## Categorizing a Continuous Variable (to compare — not recommended)

```r
df <- df |>
  mutate(x_cat = cut(x,
                     breaks = c(0, 25, 30, Inf),
                     labels = c("<25", "25-30", "30+"),
                     include.lowest = TRUE))
```

## Interpreting Diagnostic Plots

| Plot | What to look for |
|---|---|
| Residuals vs. Fitted | Random scatter around zero — no curve, no funnel |
| Normal Q-Q | Points follow the diagonal — no heavy tails |
| Scale-Location | Flat red line — no increasing/decreasing spread |
| Residuals vs. Leverage | No points in upper/lower right corners |
