---
title: "Day 6 Lab: Linear Regression"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```




**Exercise 1 — Simple Linear Regression: Diastolic BP and Age.**

a. Use `ggplot()` with `geom_smooth(method = "loess")` to inspect the
   relationship between age (`age`) and diastolic blood pressure (`diaBP`).
   Does the relationship look linear?

```{r}
ggplot(framingham, aes(x = age, y = diaBP)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE, linewidth = 1) +
  labs(x = "Age (years)", y = "Diastolic BP (mmHg)",
       title = "Diastolic BP vs. Age with Loess Smoother",
       subtitle = "Loess curve shown in red; shaded band is 95% confidence interval") +
  theme_minimal()
```

I would not say this has a linear relationship. Based on the plot, the relationship appears to be quadratic. However, a quadratic relationship goes up and then goes down. Here, there is no reason why diastolic BP would go down. As a result, I would probably use a log-transformation (in practice). 

b. Fit a simple linear regression model predicting `diaBP` from `age`.
   Write out the fitted regression equation using the estimated coefficients.
   
   Just to review, the simple linear regression model is:

$$Y = \beta_0 + \beta_1 X + \varepsilon$$

where $Y$ is diastolic blood pressure, $X$ is age, $\beta_0$ is the intercept, $\beta_1$ is the slope, and $\varepsilon$ is the error term.


```{r}
dbp_model <- lm(diaBP ~ age, data = framingham)
summary(dbp_model)
```

The fitted line is:

$\hat Y$ = `r round(coef(dbp_model)[1], 2)` + `r round(coef(dbp_model)[2], 2)`$\times diaBP + \epsilon$



c. Interpret the slope in plain language: what does it tell you about the
   relationship between age and diastolic blood pressure in this population?

**Age ($\hat\beta_1$ = `r round(coef(dbp_model)[2], 2)`):** For each additional year of age, diastolic BP is estimated to increase by `r round(coef(dbp_model)[2], 2)` mmHg on average.

d. How much of the variability in diastolic BP does age alone explain? Is this
   more or less than the R² for the `sysBP ~ age` model from the morning session
   (R² = 15.5%)?
   
   **R²:** The model explains `r round(summary(dbp_model)$r.squared * 100, 1)`% of the variability in diastolic BP. This is notably lower then the R² we found for systolic BP and age.

e. Plot the fitted regression line with a 95% confidence band using
   `geom_smooth(method = "lm")`.

```{r}
#| fig-width: 7
#| fig-height: 5
ggplot(framingham, aes(x = age, y = diaBP)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "lm", color = "firebrick", se = TRUE, linewidth = 1) +
  labs(x = "Age (years)", y = "Diastolic BP (mmHg)",
       title = "Simple Linear Regression: Diastolic BP on Age",
       subtitle = "Shaded band is 95% confidence interval for the mean") +
  theme_minimal()
```

---

**Exercise 2 — Model Diagnostics.**

Using the `diaBP ~ age` model from Exercise 1:

a. Produce the four standard diagnostic plots using `par(mfrow = c(2,2))` and
   `plot(model)`. Describe what you see in each panel. Are any model assumptions
   clearly violated?

Here are the four diagnostic plots for the diastolic BP vs age SLR model:

```{r}
#| fig-width: 8
#| fig-height: 7
par(mfrow = c(2, 2))
plot(dbp_model)
par(mfrow = c(1, 1))
```

If you look closely at the Residuals vs Fitted plot, you'll see that the red mean line is below zero for fitted values that are low and high. This is not surprising given the scatterplot we saw in 1a. It appears that the linear assumption is violated here.

Also, similar to the systolic BP model, there are some indications that the normality assumption is violated as there are some large positive residuals. These values are larger then we would expect if the residuals were actually normally distributed.

Unlike the systolic BP model, the variance of the residuals appears to be relatively consistent.

b. Produce a Cook's distance plot using `ggplot()` with `geom_segment()`.
   Flag the top 1% of observations as red points. Are there any observations
   that stand clearly apart from the rest?

```{r}
#| fig-width: 7
#| fig-height: 4
cooks_df <- tibble(
  obs   = seq_along(cooks.distance(dbp_model)),
  cooks = cooks.distance(dbp_model)
)

ggplot(cooks_df, aes(x = obs, y = cooks)) +
  geom_segment(aes(xend = obs, yend = 0), color = "steelblue", alpha = 0.6) +
  geom_point(data = filter(cooks_df, cooks > quantile(cooks, 0.99)),
             color = "firebrick", size = 2) +
  labs(x = "Observation index", y = "Cook's distance",
       title = "Cook's Distance — Simple Regression of DBP on Age",
       subtitle = "Red points are in the top 1% — worth investigating, not automatically removed") +
  theme_minimal()
```

There's not any one point that is notably larger than the others. This is a good indication that there are no influential points in the data.


c. Report the 95% confidence interval for the slope using `confint()`. What
   does this interval tell you about the precision of the estimate?

```{r}
confint(dbp_model)
```

---


**Exercise 4 — Multiple Regression: Checking Variable Form.**

Before building a multiple regression model for `sysBP`, inspect the
relationship between each continuous predictor and `sysBP` using a loess
smoother.

a. Use `pivot_longer()` and `facet_wrap()` to create a single faceted plot
   showing `sysBP` on the y-axis against each of the following predictors:
   `age`, `BMI`, `totChol`, `diaBP`, `cigsPerDay`, `glucose`. Use
   `geom_smooth(method = "loess")`.

```{r}
#| fig-width: 9
#| fig-height: 9
#| message: false
framingham |>
  pivot_longer(cols = c(age, BMI, totChol, diaBP, cigsPerDay, glucose),
               names_to = "predictor", values_to = "value") |>
  ggplot(aes(x = value, y = sysBP)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE, linewidth = 1) +
  facet_wrap(~ predictor, scales = "free_x", ncol = 2) +
  labs(x = "Predictor value", y = "Systolic BP (Y, mmHG)",
       title = "Relationship Between Each Predictor and Systolic BP",
       subtitle = "Loess smoother — a roughly straight red line supports a linear term") +
  theme_minimal()
```


b. Based on the plots, which predictors appear to have a roughly linear
   relationship with `sysBP`? Are there any that show obvious non-linearity
   or very weak relationships?

From the plots, I would say that `age` and `totChol` appear to have the most linear relationships. Also, I would say that `glucose` and `cigsPerDay` have the weakest relationships.

The relationships between BMI and systolic BP and diastolic and systolic BP are strong. Also, each appears to have some non-linearity. 

- For diastolic and systolic BP, the relationship is linear when diastolic BP is greater than 75 or so. Below 75, there does not appear to be much of a relationship between these variables. 

- BMI and systolic BP, let's ignore the line for BMI larger than 45. There are not many observations out there. For BMI from $\approx 22$ to $45$ I would say there is a consistent linear realtionship. For BMI $<22$ there doesn't appear to be much realationship between the two variables. 

c. Note: `cigsPerDay` and `glucose` have missing values. Does this affect
   the plot? Check `colSums(is.na(framingham[, c("cigsPerDay","glucose")]))`.

```{r}
nrow(framingham)
colSums(is.na(framingham[, c("cigsPerDay","glucose")]))
```

By including `cigsPerDay` and `glucose` we lose about 400 subjects, or roughly 10% of the sample. Based on our scatterplots, it appears these variables have weak relationships with the outcome. As a result, we might want to consider dropping them from our model.

