---
title: "Day 4 Lab: Data Visualization"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(this.path)
library(nycflights13)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```


**Exercise 1 — Distributions.**

Using the Framingham data:

a. Make a histogram of BMI. Experiment with at least two `binwidth` values and
   pick the one that reveals the most useful structure. Add a vertical line at
   BMI = 30 (obesity threshold) using `geom_vline()` and label it with
   `annotate()`.
   
   
   Here are histograms with binwidths equal to 1/2, 1 and 2. Personally, I like the version with 1 the best.
   
```{r}
ggplot(framingham, aes(x = BMI)) +
  geom_histogram(binwidth = 1/2, fill = "steelblue", color = "white") +
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```

```{r}
ggplot(framingham, aes(x = BMI)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```

```{r}
ggplot(framingham, aes(x = BMI)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white")+
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```

Next i'll add the verticle line at 30:

```{r}
ggplot(framingham, aes(x = BMI)) +
  geom_histogram(fill = "steelblue", color = "white") +
  geom_vline(xintercept = 30, color = "red",
             linewidth = 1, linetype = "dashed") +
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```

Finally, i'll add the annotation labeling the line.

```{r}
ggplot(framingham, aes(x = BMI)) +
  geom_histogram(fill = "steelblue", color = "white") +
  geom_vline(xintercept = 30, color = "red",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = 31, y = 600,
           label = "Obesity threshold",
           color = "red", hjust = 0, size = 4) +
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```
   

b. Overlay density plots of BMI separately for diabetic and non-diabetic
   participants using `geom_density()` with `fill` mapped to `diabetes` and
   `alpha = 0.4`. Use `scale_fill_brewer(palette = "Set1")`. What do you
   notice about the distributions?

```{r}
ggplot(framingham, aes(x = BMI, fill = diabetes)) +
  geom_density(alpha = 0.4) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```

c. Make a bar chart showing the count of participants at each education level.
   Tilt the x-axis labels so they don't overlap (`angle = 20, hjust = 1` inside
   `element_text()`). Then make a second version showing the **percentage**
   at each level rather than the count (compute percentages first with `dplyr`,
   then use `geom_col()`).

```{r}
framingham |>
  filter(!is.na(education)) |>
  ggplot(aes(x = education)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Education Level", y = "Percent (%)",
       title = "Education Distribution") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

framingham |>
  filter(!is.na(education)) |>
  count(education) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = education, y = pct)) +
  geom_col(fill = "steelblue") +
  labs(x = "Education Level", y = "Percent (%)",
       title = "Education Distribution (%)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```


---

**Exercise 3 — Scatterplots and Smoothers.**

a. Make a scatterplot of age vs. total cholesterol (`totChol`). With over 4,000
   points there will be overplotting --- use `alpha` to address it. Add a loess
   smoother (`method = "loess"`) in red with `se = TRUE`. Describe the overall
   trend.
   
```{r}
ggplot(framingham, aes(x = age, y = totChol)) +
  geom_point(alpha = 0.2) +
  # method = "loess" lines can show non-linear trends.
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  labs(x = "Age (years)", y = "Total Cholesterol",
       title = "Total Cholesterol vs. Age with Loess Smoother")
```
   

b. Repeat but color points by `currentSmoker` and add **separate** linear trend
   lines for each group (`method = "lm"`, `se = FALSE`). Does the relationship
   between age and cholesterol differ between smokers and non-smokers?


```{r}
ggplot(framingham, aes(x = age, y = totChol, color = currentSmoker)) +
  geom_point(alpha = 0.2) +
  # method = "loess" lines can show non-linear trends.
  geom_smooth(method = "lm",se = FALSE) +
  labs(x = "Age (years)", y = "Total Cholesterol",
       title = "Total Cholesterol vs. Age with Loess Smoother")
```

c. Facet the plot from (b) by sex (`facet_wrap(~ male)`). Add a complete set
   of labels and apply `theme_minimal()`. Are there any sex differences in the
   smoking–cholesterol relationship?

```{r}
ggplot(framingham, aes(x = age, y = totChol, color = currentSmoker)) +
  geom_point(alpha = 0.2) +
  # method = "loess" lines can show non-linear trends.
  geom_smooth(method = "lm",se = FALSE) +
  facet_wrap(~male) +
  labs(x = "Age (years)", y = "Total Cholesterol",
       title = "Total Cholesterol vs. Age with Loess Smoother") +
  theme_minimal()
```

---

**Exercise 5 — Joining Before Plotting.**

This exercise combines Day 3 joins with Day 4 visualization using `nycflights13`.

a. Join `flights` with `weather` on `origin` and `time_hour` to add visibility
   (`visib`) and temperature (`temp`) to each flight. Call this `flights_wx`.

```{r}
library(nycflights13)

flights_wx <- flights |> 
  left_join(weather |> select(origin, time_hour, visib, temp), 
            )
```


b. Compute the mean arrival delay by rounded visibility level
   (`mutate(visib = round(visib))`), grouped by origin airport. Then plot mean
   delay (y) vs. visibility (x) with a separate colored line per origin airport
   using `geom_line()` and `geom_point()`. Use `scale_color_brewer(palette = "Dark2")`.
   Is there a relationship between visibility and delays? Does it differ by airport?
   
   
   First, i'm going to get the data that we we'll end up plotting.
```{r}
mn_delay_vis_origin = flights_wx |> 
  # Creating rounded visibility variable.
  mutate(visib = round(visib)) |> 
  # Grouping by origin and visibility
  group_by(origin, visib) |> 
  # Getting the average delay
  summarise(
    mean_delay = mean(arr_delay, na.rm = TRUE),
    .groups = "drop"
  )
mn_delay_vis_origin
```

Now, I'm going to plot this data:

```{r}
#| warning: false
# Give ggplot the data along with the x, y, and color variables.
  ggplot(mn_delay_vis_origin, aes(x = visib, y = mean_delay, color = origin)) +
  # Add lines and points
  geom_line() + geom_point() +
  # Fancy colors
  scale_color_brewer(palette = "Dark2") +
  # Nicer labels
  labs(x = "Visibility", y = "Average Delay",
       title = "Average Delay vs. Visibility by Airport") +
  theme_minimal()
```
   
   If you were so inclined, you could do the above all at once:
   
```{r}
#| eval: false
flights_wx |> 
  # Creating rounded visibility variable.
  mutate(visib = round(visib)) |> 
  # Grouping by origin and visibility
  group_by(origin, visib) |> 
  # Getting the average delay
  summarise(
    mean_delay = mean(arr_delay, na.rm = TRUE),
    .groups = "drop"
  ) |> 
# Give ggplot the data along with the x, y, and color variables.
  ggplot(aes(x = visib, y = mean_delay, color = origin)) +
  # Add lines and points
  geom_line() + geom_point() +
  # Fancy colors
  scale_color_brewer(palette = "Dark2") +
  # Nicer labels
  labs(x = "Visibility", y = "Average Delay",
       title = "Average Delay vs. Visibility by Airport") +
  theme_minimal()
```   

c. Make a boxplot of departure delay (`dep_delay`) by `origin` airport, limited
   to delays between --20 and 120 minutes (use `filter()` first, or
   `coord_cartesian()`). Order the boxes by median delay, apply
   `theme_classic()`, and add proper labels.

```{r}
flights_wx |> 
  # filter to delays between 20 and 120
  filter(arr_delay >= -20 & arr_delay <= 120) |> 
  # Giving ggplot the x variable (reorder by increaseing median arrival delay) and y variable
  ggplot(aes(x = reorder(origin, arr_delay, FUN = median), y = arr_delay)) +
  # Adding the boxplot
  geom_boxplot() +
  # Classic Theme
  theme_classic() +
  # Adding labels
  labs(x = "Origin Airport", y = "Arrival Delay",
       title = "Boxplot of Arrival Delay by Origin Airport")
```


---

**Exercise 6 (Challenge) — A Complete Figure.**

Build a single polished figure that answers this question: **How does the
10-year coronary heart disease rate in the Framingham cohort vary by age group,
sex, and hypertension status?**

Steps:

a. Create an age group variable using `case_when()`:
   "40--44", "45--49", "50--54", "55--59", "60--64", "65+"
   (or similar grouping of your choosing). Make it a factor with levels in
   the right order.

```{r}
framingham <- framingham |>
  mutate(
    age_group = factor(case_when(
      age < 45 ~ "40-44", 
      age < 50 ~ "45-49", 
      age < 55 ~ "50-54", 
      age < 60 ~ "55-59", 
      age < 65 ~ "60-64", 
      age >= 65 ~ "65+"
    ))
  )
# Check the ordering of the factor
framingham$age_group[1:5]
```


b. Compute the CHD rate (proportion with `TenYearCHD == "Yes"`) for each
   combination of age group, sex, and hypertension status using
   `group_by()` and `summarize()`. Keep only groups with at least 20
   participants.

```{r}
avg_CHD_data <- framingham |> 
  group_by(age_group, male, prevalentHyp) |> 
  summarise(
    N = n(),
    avg_CHD = mean(TenYearCHD == "Yes")
  ) |> 
  filter(N >=20)
```
 
c. Plot the result. A grouped bar chart (`geom_col()` with `position = "dodge"`)
   or a line plot (`geom_line()` + `geom_point()`) with age group on the x-axis
   works well here. Color by hypertension status, facet by sex.

```{r}
ggplot(avg_CHD_data, aes(x = age_group, y = avg_CHD, fill = prevalentHyp)) +
  geom_col(position = "dodge") +
  facet_wrap(~male)
```


d. Add a full set of labels, a meaningful color palette, and a clean theme.

```{r}
ggplot(avg_CHD_data, aes(x = age_group, y = avg_CHD, fill = prevalentHyp)) +
  geom_col(position = "dodge") +
  facet_wrap(~male) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold")
  ) +
  labs(
    title    = "Average 10 Year CHD Rates by Age",
    subtitle = "Framingham Heart Study cohort",
    caption  = "Source: Framingham Heart Study",
    x        = "Age Group (years)",
    y        = "10 CHD Rates",
    fill     = "Hypertension Status"
  )
```

e. Save the figure with `ggsave()` at 300 dpi.

```{r}
ggsave("figures/chd_vs_age.png", width = 7, height = 5, dpi = 300)
ggsave("figures/chd_vs_age.pdf", width = 7, height = 5)
```


f. Write a 3--4 sentence caption summarizing the main findings as if this were
   going into a manuscript.

First, the CHD rates increase by age group for males and females. Overall, the CHD rates for males are higher than they are for females. Lastly, it appears that the different in CHD rates between those with and without hypertension, is different for males and females. For females, having hypertension is a bigger risk factor for CHD than for males (especially for the older age groups). 
