---
title: "Day 4 Lab: Data Visualization"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(this.path)
library(nycflights13)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```


**Exercise 1 — Distributions.**

Using the Framingham data:

a. Make a histogram of BMI. Experiment with at least two `binwidth` values and
   pick the one that reveals the most useful structure. Add a vertical line at
   BMI = 30 (obesity threshold) using `geom_vline()` and label it with
   `annotate()`.
   
   
   Here are histograms with binwidths equal to 1/2, 1 and 2. Personally, I like the version with 1 the best.
   
```{r}
ggplot(framingham, aes(x = BMI)) +
  geom_histogram(binwidth = 1/2, fill = "steelblue", color = "white") +
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```

```{r}
ggplot(framingham, aes(x = BMI)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```

```{r}
ggplot(framingham, aes(x = BMI)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white")+
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```

Next i'll add the verticle line at 30:

```{r}
ggplot(framingham, aes(x = BMI)) +
  geom_histogram(fill = "steelblue", color = "white") +
  geom_vline(xintercept = 30, color = "red",
             linewidth = 1, linetype = "dashed") +
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```

Finally, i'll add the annotation labeling the line.

```{r}
ggplot(framingham, aes(x = BMI)) +
  geom_histogram(fill = "steelblue", color = "white") +
  geom_vline(xintercept = 30, color = "red",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = 31, y = 600,
           label = "Obesity threshold",
           color = "red", hjust = 0, size = 4) +
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```
   

b. Overlay density plots of BMI separately for diabetic and non-diabetic
   participants using `geom_density()` with `fill` mapped to `diabetes` and
   `alpha = 0.4`. Use `scale_fill_brewer(palette = "Set1")`. What do you
   notice about the distributions?

```{r}
ggplot(framingham, aes(x = BMI, fill = diabetes)) +
  geom_density(alpha = 0.4) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Body Mass Index (BMI)", y = "Count",
       title = "Distribution of BMI")
```

c. Make a bar chart showing the count of participants at each education level.
   Tilt the x-axis labels so they don't overlap (`angle = 20, hjust = 1` inside
   `element_text()`). Then make a second version showing the **percentage**
   at each level rather than the count (compute percentages first with `dplyr`,
   then use `geom_col()`).

```{r}
framingham |>
  filter(!is.na(education)) |>
  ggplot(aes(x = education)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Education Level", y = "Percent (%)",
       title = "Education Distribution") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

framingham |>
  filter(!is.na(education)) |>
  count(education) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = education, y = pct)) +
  geom_col(fill = "steelblue") +
  labs(x = "Education Level", y = "Percent (%)",
       title = "Education Distribution (%)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```


---

**Exercise 3 — Scatterplots and Smoothers.**

a. Make a scatterplot of age vs. total cholesterol (`totChol`). With over 4,000
   points there will be overplotting --- use `alpha` to address it. Add a loess
   smoother (`method = "loess"`) in red with `se = TRUE`. Describe the overall
   trend.
   
```{r}
ggplot(framingham, aes(x = age, y = totChol)) +
  geom_point(alpha = 0.2) +
  # method = "loess" lines can show non-linear trends.
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  labs(x = "Age (years)", y = "Total Cholesterol",
       title = "Total Cholesterol vs. Age with Loess Smoother")
```
   

b. Repeat but color points by `currentSmoker` and add **separate** linear trend
   lines for each group (`method = "lm"`, `se = FALSE`). Does the relationship
   between age and cholesterol differ between smokers and non-smokers?


```{r}
ggplot(framingham, aes(x = age, y = totChol, color = currentSmoker)) +
  geom_point(alpha = 0.2) +
  # method = "loess" lines can show non-linear trends.
  geom_smooth(method = "lm",se = FALSE) +
  labs(x = "Age (years)", y = "Total Cholesterol",
       title = "Total Cholesterol vs. Age with Loess Smoother")
```

c. Facet the plot from (b) by sex (`facet_wrap(~ male)`). Add a complete set
   of labels and apply `theme_minimal()`. Are there any sex differences in the
   smoking–cholesterol relationship?

```{r}
ggplot(framingham, aes(x = age, y = totChol, color = currentSmoker)) +
  geom_point(alpha = 0.2) +
  # method = "loess" lines can show non-linear trends.
  geom_smooth(method = "lm",se = FALSE) +
  facet_wrap(~male) +
  labs(x = "Age (years)", y = "Total Cholesterol",
       title = "Total Cholesterol vs. Age with Loess Smoother") +
  theme_minimal()
```

---

**Exercise 5 — Joining Before Plotting.**

This exercise combines Day 3 joins with Day 4 visualization using `nycflights13`.

a. Join `flights` with `weather` on `origin` and `time_hour` to add visibility
   (`visib`) and temperature (`temp`) to each flight. Call this `flights_wx`.

```{r}
library(nycflights13)

flights_wx <- flights |> 
  left_join(weather |> select(origin, time_hour, visib, temp), 
            )
```


b. Compute the mean arrival delay by rounded visibility level
   (`mutate(visib = round(visib))`), grouped by origin airport. Then plot mean
   delay (y) vs. visibility (x) with a separate colored line per origin airport
   using `geom_line()` and `geom_point()`. Use `scale_color_brewer(palette = "Dark2")`.
   Is there a relationship between visibility and delays? Does it differ by airport?
   
   
   First, i'm going to get the data that we we'll end up plotting.
```{r}
mn_delay_vis_origin = flights_wx |> 
  # Creating rounded visibility variable.
  mutate(visib = round(visib)) |> 
  # Grouping by origin and visibility
  group_by(origin, visib) |> 
  # Getting the average delay
  summarise(
    mean_delay = mean(arr_delay, na.rm = TRUE),
    .groups = "drop"
  )
mn_delay_vis_origin
```

Now, I'm going to plot this data:

```{r}
#| warning: false
# Give ggplot the data along with the x, y, and color variables.
  ggplot(mn_delay_vis_origin, aes(x = visib, y = mean_delay, color = origin)) +
  # Add lines and points
  geom_line() + geom_point() +
  # Fancy colors
  scale_color_brewer(palette = "Dark2") +
  # Nicer labels
  labs(x = "Visibility", y = "Average Delay",
       title = "Average Delay vs. Visibility by Airport") +
  theme_minimal()
```
   
   If you were so inclined, you could do the above all at once:
   
```{r}
#| eval: false
flights_wx |> 
  # Creating rounded visibility variable.
  mutate(visib = round(visib)) |> 
  # Grouping by origin and visibility
  group_by(origin, visib) |> 
  # Getting the average delay
  summarise(
    mean_delay = mean(arr_delay, na.rm = TRUE),
    .groups = "drop"
  ) |> 
# Give ggplot the data along with the x, y, and color variables.
  ggplot(aes(x = visib, y = mean_delay, color = origin)) +
  # Add lines and points
  geom_line() + geom_point() +
  # Fancy colors
  scale_color_brewer(palette = "Dark2") +
  # Nicer labels
  labs(x = "Visibility", y = "Average Delay",
       title = "Average Delay vs. Visibility by Airport") +
  theme_minimal()
```   

c. Make a boxplot of departure delay (`dep_delay`) by `origin` airport, limited
   to delays between --20 and 120 minutes (use `filter()` first, or
   `coord_cartesian()`). Order the boxes by median delay, apply
   `theme_classic()`, and add proper labels.

```{r}
flights_wx |> 
  # filter to delays between 20 and 120
  filter(arr_delay >= -20 & arr_delay <= 120) |> 
  # Giving ggplot the x variable (reorder by increaseing median arrival delay) and y variable
  ggplot(aes(x = reorder(origin, arr_delay, FUN = median), y = arr_delay)) +
  # Adding the boxplot
  geom_boxplot() +
  # Classic Theme
  theme_classic() +
  # Adding labels
  labs(x = "Origin Airport", y = "Arrival Delay",
       title = "Boxplot of Arrival Delay by Origin Airport")
```

