---
title: "Day 13: Multiple Testing and Clustering"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(this.path)
library(SummarizedExperiment)
library(DESeq2)
library(airway)
library(factoextra)
library(cluster)
library(pheatmap)
library(patchwork)
setwd(this.dir())

data(airway)
```

# Overview

Yesterday we produced a DESeq2 results table with one row per gene. Two
columns appeared in that table -- `pvalue` and `padj` -- and we said to always
use `padj`. Today we explain why, and then turn to a related question: once you
have a list of differentially expressed genes, how do you find structure in it?

Two topics:

- **Part 1: Multiple Testing.** What the `padj` column actually contains, why
  raw p-values are not usable for 20,000 simultaneous tests, and how the
  Benjamini-Hochberg procedure fixes this.
- **Part 2: Clustering.** K-means and hierarchical clustering, how to choose
  the number of clusters, and how these methods connect to the heatmaps and
  PCA plots from Days 11--12.

Throughout we use the `airway` dataset so that every result connects directly
to yesterday's analysis.

---

# Part 1: Multiple Testing

## The Problem

Let us start with a concrete demonstration. If we test one gene at $\alpha =
0.05$ and the gene is not differentially expressed, there is a 5% chance of a
false positive. That is an acceptable error rate for one test. But in an RNA-seq
experiment we run thousands of tests simultaneously.

```{r}
# How many tests does DESeq2 actually run on the airway data?
# Re-run the Day 12 DESeq2 analysis quickly
counts     <- assay(airway)
sample_info <- as.data.frame(colData(airway))

dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = sample_info,
  design    = ~ cell + dex
)
dds$dex <- relevel(dds$dex, ref = "untrt")
keep    <- rowSums(counts(dds)) >= 10
dds     <- dds[keep, ]
deseq_dds <- DESeq(dds)
res       <- results(deseq_dds)

# How many genes were tested?
cat("Total genes tested:", sum(!is.na(res$pvalue)), "\n")
```

```{r}
# If we used raw p-values at alpha = 0.05, how many false positives
# would we expect purely by chance if NO gene were truly DE?
n_tested <- sum(!is.na(res$pvalue))
expected_fp <- 0.05 * n_tested
cat("Expected false positives at alpha = 0.05:", round(expected_fp), "\n")
```

With roughly 12,000--15,000 genes being tested, we would expect 600--750 false
positives from raw p-values alone, even in a dataset where nothing is
differentially expressed. The raw p-value column is not a usable basis for
calling significant genes.

## Visualizing the Problem

```{r}
#| fig-width: 8
#| fig-height: 4
res_df <- as.data.frame(res) |>
  rownames_to_column("gene_id") |>
  as_tibble() |>
  filter(!is.na(pvalue), !is.na(padj))

p1 <- ggplot(res_df, aes(x = pvalue)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0.05, color = "firebrick",
             linetype = "dashed", linewidth = 1) +
  labs(title = "Raw p-values",
       subtitle = "Red line = 0.05 threshold",
       x = "p-value", y = "Number of genes") +
  theme_minimal()

p2 <- ggplot(res_df, aes(x = padj)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0.05, color = "firebrick",
             linetype = "dashed", linewidth = 1) +
  labs(title = "Adjusted p-values (BH)",
       subtitle = "Red line = 0.05 threshold",
       x = "Adjusted p-value", y = "Number of genes") +
  theme_minimal()

p1 + p2
```

**Reading the p-value histogram:** a uniform distribution under the null
(the flat region on the right) mixed with a spike near zero (the true
positives) is the expected shape. The flat region tells you roughly how many
null hypotheses are in your data; the spike tells you how many true signals
there are.

## Two Error Rates

The multiple testing literature distinguishes two quantities worth controlling.

**Family-wise error rate (FWER):**
$$\text{FWER} = P(\text{at least one false positive})$$

**False discovery rate (FDR):**
$$\text{FDR} = E\!\left(\frac{\text{false positives}}{\text{all rejections}}\right)$$

The FDR is a proportion, not a probability. Controlling FDR at 5% means that
among all the genes you call significant, you expect at most 5% to be false
positives. This is the right criterion for genomics: you can tolerate a few
false leads in a list of 500 candidate genes, but you do not want half of them
to be noise.

## Bonferroni: Simple but Severe

The Bonferroni correction controls the FWER by dividing $\alpha$ by the number
of tests:

```{r}
# Bonferroni threshold for airway data
alpha    <- 0.05
bonf_threshold <- alpha / n_tested
cat("Bonferroni threshold:", formatC(bonf_threshold, format = "e", digits = 2), "\n")

# How many genes pass?
cat("Genes significant after Bonferroni:",
    sum(res_df$pvalue < bonf_threshold, na.rm = TRUE), "\n")

# Compare to unadjusted
cat("Genes with raw p < 0.05:",
    sum(res_df$pvalue < 0.05, na.rm = TRUE), "\n")
```

Bonferroni demands a raw p-value below roughly $3 \times 10^{-6}$ to call a
gene significant. That is extremely conservative. In a dataset with strong
treatment effects like `airway`, Bonferroni will still find many significant
genes. But in a noisier study with weaker effects it would miss nearly
everything. This is not caution -- it is a different kind of error, called a
Type II error (false negative).

## The Benjamini-Hochberg (BH) Procedure

The BH procedure controls the FDR. The intuition is to compare each sorted
p-value to a threshold that scales with its rank.

```{r}
# Manual BH calculation on a small example
# (to make the procedure concrete before applying to 15,000 genes)
example_pvals <- c(0.0008, 0.009, 0.016, 0.019, 0.196,
                   0.350, 0.541, 0.781, 0.900, 0.993)
M     <- length(example_pvals)
alpha <- 0.05

bh_df <- tibble(
  rank    = seq_along(example_pvals),
  pvalue  = sort(example_pvals),
  threshold = (rank / M) * alpha,
  below   = pvalue <= threshold
)

bh_df
```

```{r}
# The BH rule: reject all hypotheses up to and including
# the highest rank where the condition is met
k_star <- max(bh_df$rank[bh_df$below])
cat("Largest rank k where p_(k) <= (k/M)*alpha:", k_star, "\n")
cat("Reject hypotheses at ranks 1 through", k_star, "\n")
```

```{r}
#| fig-width: 7
#| fig-height: 4
bh_df |>
  ggplot(aes(x = rank, y = pvalue)) +
  geom_point(size = 3, color = "steelblue") +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_abline(slope = alpha / M, intercept = 0,
              color = "firebrick", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = k_star + 0.5, color = "gray40",
             linetype = "dotted", linewidth = 1) +
  annotate("text", x = k_star - 0.3, y = 0.15,
           label = paste("Reject\nall up to\nrank", k_star),
           hjust = 1, size = 3.5) +
  labs(x = "Rank (j)", y = "p-value",
       title = "BH Procedure: Sorted p-values vs. BH Threshold Line",
       subtitle = "Red line: BH threshold (j/M)*alpha. Points below = rejected.") +
  theme_minimal()
```

The key insight: we reject everything with rank up to 4, even though rank 3's
p-value fell above its threshold. The BH rule is about the *last* rank where
the condition is met, not every rank individually.

## BH in Practice: DESeq2 Results

```{r}
# Compute BH-adjusted p-values manually and compare to DESeq2's padj
manual_padj <- p.adjust(res_df$pvalue, method = "BH")

# They should match DESeq2's padj closely
tibble(
  deseq2_padj = res_df$padj,
  manual_padj = manual_padj
) |>
  filter(!is.na(deseq2_padj)) |>
  summarize(
    max_diff  = max(abs(deseq2_padj - manual_padj[!is.na(deseq2_padj)])),
    mean_diff = mean(abs(deseq2_padj - manual_padj[!is.na(deseq2_padj)]))
  )
```

The values are not identical, but this is not always the case. The differences can come from DESeq2's
**independent filtering**: before applying BH, DESeq2 automatically removes
genes with very low mean expression (they would require extreme p-values to
pass any threshold, so including them only hurts power for the genes that
remain). The filtered genes get `padj = NA`, which is why you see `NA` values
in the results table.

```{r}
# Summary: raw vs adjusted
tibble(
  Criterion        = c("Raw p < 0.05", "BH adjusted p < 0.05",
                       "Bonferroni adjusted p < 0.05"),
  `Genes passing`  = c(
    sum(res_df$pvalue < 0.05, na.rm = TRUE),
    sum(res_df$padj   < 0.05, na.rm = TRUE),
    sum(p.adjust(res_df$pvalue, method = "bonferroni") < 0.05, na.rm = TRUE)
  )
)
```

The BH procedure finds far more significant genes than Bonferroni, while still
controlling the proportion of false discoveries. This is why the genomics field
adopted it as the standard.

---

# Part 2: Clustering

A list of 500 differentially expressed genes is hard to interpret one gene at
a time. Clustering helps by grouping genes with similar expression patterns --
and grouping samples with similar overall profiles -- so you can identify
biological themes rather than individual genes.

## Setup: VST Data for Clustering

Clustering requires data on a comparable scale. We use the variance-stabilizing
transformation (VST) from Day 12, restricted to the top 200 most variable genes.

```{r}
vsd <- vst(deseq_dds, blind = FALSE)

# Keep the 200 most variable genes (across samples)
top_var_genes <- order(rowVars(assay(vsd)), decreasing = TRUE)[1:200]
mat           <- assay(vsd)[top_var_genes, ]

# Scale each gene to zero mean and unit variance
mat_scaled <- t(scale(t(mat)))

cat("Matrix dimensions:", nrow(mat_scaled), "genes x",
    ncol(mat_scaled), "samples\n")
```

## K-Means Clustering

### Choosing K

The first step is deciding how many clusters to look for. Two diagnostic tools
are the elbow plot and the silhouette width.

```{r}
#| fig-width: 8
#| fig-height: 4
#| message: false

# Elbow plot: total within-cluster sum of squares vs K
p_elbow <- fviz_nbclust(mat_scaled, kmeans,
                         method  = "wss",
                         k.max   = 10,
                         nstart  = 25) +
  labs(title = "Elbow Method") +
  theme_minimal()

# Silhouette width vs K
p_sil <- fviz_nbclust(mat_scaled, kmeans,
                       method = "silhouette",
                       k.max  = 10,
                       nstart = 25) +
  labs(title = "Average Silhouette Width") +
  theme_minimal()

p_elbow + p_sil
```

**Reading these plots:**

- **Elbow plot:** look for a bend where adding more clusters gives diminishing
  returns. A clear elbow at $K = 3$ or $K = 4$ is common in gene expression
  data.
- **Silhouette plot:** choose the $K$ that maximizes the average silhouette
  width. Values above 0.5 indicate reasonably well-separated clusters; values
  near 0.2 indicate weak structure.

### Fitting K-Means

```{r}
set.seed(2026)
km_fit <- kmeans(mat_scaled, centers = 3, nstart = 25)
table(km_fit$cluster)
```

```{r}
#| fig-width: 6
#| fig-height: 5
fviz_cluster(km_fit,
             data          = mat_scaled,
             geom          = "point",
             ellipse.type  = "norm",
             palette       = c("steelblue","firebrick","darkgreen"),
             ggtheme       = theme_minimal()) +
  labs(title = "K-Means Clustering of Top 200 Variable Genes (K=3)")
```

The cluster plot projects the 200-dimensional data onto the first two principal
components for visualization. The four clusters represent groups of genes with
similar expression profiles across the eight samples.

## Hierarchical Clustering

K-means required us to choose $K = 3$ upfront. Hierarchical clustering builds
a tree that shows all possible groupings, letting us choose the number of
clusters after seeing the structure.

### Clustering Samples

With only 8 samples, hierarchical clustering produces a readable dendrogram
and is the standard way to check whether samples group by treatment.

```{r}
#| fig-width: 7
#| fig-height: 5
# Sample-level clustering (columns of the VST matrix)
sample_dist <- dist(t(mat_scaled), method = "euclidean")
hc_samples  <- hclust(sample_dist, method = "complete")

# Color labels by treatment group
dend_labels <- colData(vsd)$dex[order.dendrogram(as.dendrogram(hc_samples))]
label_colors <- ifelse(dend_labels == "trt", "firebrick", "steelblue")

plot(hc_samples,
     main   = "Hierarchical Clustering of Samples",
     xlab   = "",
     sub    = "",
     labels = paste(colData(vsd)$dex,
                    colData(vsd)$cell, sep = "_"))
```

```{r}
# Cut to 2 clusters and check against treatment labels
sample_clusters <- cutree(hc_samples, k = 2)
table(Cluster = sample_clusters, Treatment = colData(vsd)$dex)
```

If the clustering is working as expected, the two clusters should correspond
closely to the treated and untreated groups. Any sample that clusters with the
wrong group is worth investigating.

### Clustering Genes

```{r}
#| fig-width: 7
#| fig-height: 5
# Gene-level clustering (rows) -- use a smaller set for a readable plot
top200  <- order(rowVars(assay(vsd)), decreasing = TRUE)[1:200]
mat200  <- t(scale(t(assay(vsd)[top200, ])))
gene_dist <- dist(mat200, method = "euclidean")
hc_genes  <- hclust(gene_dist, method = "complete")

plot(hc_genes,
     main   = "Hierarchical Clustering of Top 50 Variable Genes",
     xlab   = "",
     sub    = "",
     labels = FALSE)
abline(h = 12, lty = 2, col = "firebrick")
```

```{r}
# Cut to 3 clusters
gene_clusters_hc <- cutree(hc_genes, k = 3)
table(gene_clusters_hc)
```

The dashed line shows where we would cut the tree to get 4 clusters. Adjusting
the cut height changes the number of clusters: higher cut = fewer, coarser
clusters; lower cut = more, finer clusters.

### Comparing K-Means and Hierarchical Results

```{r}
# Do the two methods agree on the top 200 genes?
comparison <- tibble(
  kmeans_cluster = km_fit$cluster,
  hc_cluster     = gene_clusters_hc
)

table(KMeans = comparison$kmeans_cluster,
      Hierarchical = comparison$hc_cluster)
```

Perfect agreement between two different methods would be unusual. Some genes
that K-means puts in one cluster may land in a different cluster
hierarchically. Where both methods agree, you can be more confident that a
grouping reflects real biological structure rather than an artifact of the
algorithm.


### Visualizing Gene Clusters in a Heatmap

```{r}
#| fig-width: 8
#| fig-height: 7
# Add cluster labels to the row annotation
gene_clusters <- data.frame(
  K_means = factor(paste0("C", km_fit$cluster)),
  Hierarc = factor(paste0("C", gene_clusters_hc)),
  row.names = names(km_fit$cluster)
)

col_anno <- colData(vsd) |>
  as_tibble(rownames = "sample") |>
  select(sample, dex, cell) |>
  column_to_rownames("sample")

pheatmap(mat_scaled,
         annotation_row    = gene_clusters,
         annotation_col    = col_anno,
         cluster_rows      = TRUE,
         cluster_cols      = TRUE,
         show_rownames     = FALSE,
         show_colnames     = FALSE,
         color             = colorRampPalette(
           c("steelblue", "white", "firebrick"))(100),
         main              = "Top 200 Variable Genes: K-Means versus Hierarchical Clusters")
```

Look at whether the gene clusters (row annotation) correspond to blocks of
consistently up- or down-regulated genes across the treated vs. untreated
samples. If K-means is working well, each cluster should show a coherent
expression pattern.


## Silhouette Widths: Evaluating Cluster Quality

```{r}
#| fig-width: 7
#| fig-height: 4
# Silhouette plot for the K-means solution (K=4)
sil_obj <- silhouette(km_fit$cluster, dist(mat_scaled))
fviz_silhouette(sil_obj,
                palette = c("steelblue","firebrick","darkgreen","orange"),
                ggtheme = theme_minimal()) +
  labs(title = "Silhouette Plot: K-Means K=3")
```

**Reading the silhouette plot:**

- Each bar is one gene, colored by its cluster.
- Tall bars (near 1): the gene fits well in its assigned cluster and is far
  from the nearest other cluster.
- Short or negative bars: the gene is ambiguous -- it is almost as close to
  another cluster as to its own.
- Average silhouette widths near 0.1--0.2 are common for high-dimensional
  genomics data. This does not mean the clustering is wrong; it means the
  clusters are not perfectly separated in the full high-dimensional space,
  which is expected.

---

# Summary

**Multiple testing:**

- Testing $M$ genes simultaneously at $\alpha = 0.05$ generates an expected
  $0.05 \times M$ false positives from raw p-values alone.
- Bonferroni controls the FWER by multiplying p-values by $M$. Very
  conservative; rarely appropriate in genomics.
- The Benjamini-Hochberg procedure controls the FDR. If you call all genes
  with `padj < 0.05` significant, at most 5% of those calls are expected to
  be false positives. This is what DESeq2's `padj` column contains. Always
  use it.

**Clustering:**

- K-means partitions genes (or samples) into $K$ groups by minimizing
  within-cluster variance. Choose $K$ with the elbow plot or silhouette width.
  Always scale your data and use `nstart > 1`.
- Hierarchical clustering builds a dendrogram without requiring $K$ upfront.
  Complete linkage is the default. Cut the tree at any height to get any
  number of clusters.
- Clustering is exploratory. Validate findings by checking whether two
  methods agree, whether the clusters show coherent patterns in a heatmap,
  and whether cluster membership correlates with known biology.

Tomorrow: PCA and penalized regression in the genomics context.
