---
title: "Day 13 (PM) Lab: Multiple Testing and Clustering"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This afternoon you will apply the Day 13 morning material to the `fission`
dataset: an RNA-sequencing experiment in fission yeast (*S. pombe*) comparing
a wild-type strain to an *atf21* deletion mutant under oxidative stress.

The experiment has two factors: strain (wild-type vs mutant) and time (0, 15,
30, 60, 120, 180 minutes after stress). Today we use only the **time zero**
samples -- three wild-type and three mutant replicates -- and ask: which genes
differ between the two strains at baseline?

The DESeq2 workflow is identical to what you have already done. The new
material is in interpreting the multiple testing results and clustering the
output.

::: {.callout-note}
Install `fission` if needed: `BiocManager::install("fission")`

Install `factoextra` if needed: `install.packages("factoextra")`
:::

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** Load the data, subset to time zero,
  and run DESeq2 together.
- **Part 2 (on your own, ~90 min):** Four exercises on multiple testing and
  clustering.

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(this.path)
library(SummarizedExperiment)
library(DESeq2)
library(fission)
library(factoextra)
library(cluster)
library(pheatmap)
library(patchwork)
# Some conflicted functions for this .qmd, setting defaults here:
select  <- dplyr::select
filter  <- dplyr::filter
setdiff <- base::setdiff
setwd(this.dir())

data(fission)
fission
```

## Explore the Design

```{r}
colData(fission) |>
  as_tibble()
```

The experiment has `r ncol(fission)` samples: 3 replicates per strain per
time point, across 6 time points. For today we subset to time zero (the
unstressed baseline) so the comparison is straightforward: wild-type vs.
mutant with 3 replicates each.

```{r}
# Subset to time = 0 only
fission_t0 <- fission[, fission$minute == "0"]
colData(fission_t0) |> as_tibble() 
```

## Build and Run DESeq2

```{r}
dds <- DESeqDataSet(fission_t0, design = ~ strain)
dds$strain <- relevel(dds$strain, ref = "wt")

keep <- rowSums(counts(dds)) >= 10
dds  <- dds[keep, ]
cat("Genes retained after filtering:", nrow(dds), "\n")
```

```{r}
deseq_dds <- DESeq(dds)
res        <- results(deseq_dds)
summary(res)
```

## Results Table

```{r}
res_df <- as.data.frame(res) |>
  rownames_to_column("gene_id") |>
  as_tibble() |>
  arrange(padj)

head(res_df, 10)
```

## P-Value Histogram

```{r}
#| fig-width: 8
#| fig-height: 4
p1 <- res_df |>
  filter(!is.na(pvalue)) |>
  ggplot(aes(x = pvalue)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0.05, color = "firebrick",
             linetype = "dashed", linewidth = 1) +
  labs(title = "Raw p-values", x = "p-value", y = "Genes") +
  theme_minimal()

p2 <- res_df |>
  filter(!is.na(padj)) |>
  ggplot(aes(x = padj)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0.05, color = "firebrick",
             linetype = "dashed", linewidth = 1) +
  labs(title = "BH-adjusted p-values", x = "Adjusted p-value",
       y = "Genes") +
  theme_minimal()

p1 + p2
```

The spike near zero in the raw p-value histogram represents the true
differentially expressed genes. The flat right-hand region represents the null
genes. The BH procedure uses this structure to set an adaptive threshold that
controls the false discovery rate.

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document. Title it
`"Day 13 Lab: Multiple Testing and Clustering"`.

**Step 2.** Copy the full setup and demo chunks (through the `res_df`
creation). You need `deseq_dds`, `res`, and `res_df` available before
starting.

**Step 3.** Work through Exercises 1--4. Add 2--3 sentences of interpretation
after each result.

**Step 4.** Render when finished.

---

**Exercise 1 -- Comparing Correction Methods.**

a. How many genes have a raw p-value below 0.05? Below 0.01?

b. Compute Bonferroni-adjusted p-values using
   `p.adjust(res_df$pvalue, method = "bonferroni")`. How many genes are
   significant at the 0.05 level after Bonferroni correction?

c. How many genes have `padj < 0.05` (BH correction)?

d. Summarize the three counts in a tibble with columns `Method` and
   `N_significant`. Describe in one sentence why BH finds more significant
   genes than Bonferroni.

e. Some genes have `padj = NA`. Count them. Then inspect their `baseMean`
   values:
   ```r
   res_df |> filter(is.na(padj)) |> pull(baseMean) |> summary()
   ```
   What do you notice? Why does removing these genes before applying BH
   increase power for the genes that remain?

```{r}
# your code here
```

---

**Exercise 2 -- Exploring the DE Results.**

a. Among genes with `padj < 0.05`, how many are upregulated
   (log2FoldChange > 0) and how many are downregulated? In this experiment,
   which group does upregulated mean higher expression in -- wild-type or
   mutant?

b. Print the top 10 most significant genes sorted by `padj`. Report gene
   ID, log2FoldChange, and padj.

c. For the most significant gene, convert the log2 fold change to a regular
   fold change with `2^(log2FoldChange)`. Interpret this in plain language.

d. Produce a volcano plot: `log2FoldChange` on the x-axis, `-log10(padj)`
   on the y-axis. Color significant genes (padj < 0.05) in red. Add dashed
   lines at LFC = -1, LFC = 1, and `-log10(0.05)`. Describe what the genes
   in the top-right and top-left corners represent.

```{r}
# your code here
```

---

**Exercise 3 -- Hierarchical Clustering of Samples.**

a. Apply the variance-stabilizing transformation:
   ```r
   vsd <- vst(deseq_dds, blind = FALSE)
   ```

b. Take the 300 most variable genes and row-scale the matrix:
   ```r
   top_var <- order(rowVars(assay(vsd)), decreasing = TRUE)[1:300]
   mat     <- t(scale(t(assay(vsd)[top_var, ])))
   ```

c. Compute sample-to-sample Euclidean distances with `dist(t(mat))`.
   Fit hierarchical clustering with complete linkage.

d. Plot the dendrogram, labeling each leaf with strain and sample name:
   ```r
   plot(hc,
        labels = paste(colData(vsd)$strain,
                       colData(vsd)$sample, sep = "_"))
   ```

e. Cut to 2 clusters with `cutree(hc, k = 2)`. Cross-tabulate cluster
   membership against strain. Do the two clusters correspond to wild-type
   and mutant? What does this tell you about the overall effect of the
   mutation on the transcriptome?

```{r}
# your code here
```

---

**Exercise 4 -- K-Means Clustering of Genes.**

a. Using the `mat` matrix from Exercise 3, use the silhouette method to
   help choose K:
   ```r
   fviz_nbclust(mat, kmeans, method = "silhouette",
                k.max = 8, nstart = 25)
   ```
   Which K maximizes average silhouette width?

b. Fit K-means with your chosen K using `set.seed(2026)` and
   `nstart = 25`. How many genes are in each cluster?

c. Produce a heatmap with `pheatmap()` including row annotations for
   K-means cluster and column annotations for strain:
   ```r
   gene_anno <- data.frame(
     Cluster   = paste0("C", km_fit$cluster),
     row.names = names(km_fit$cluster)
   )
   col_anno <- data.frame(
     Strain    = as.character(colData(vsd)$strain),
     row.names = colnames(vsd)
   )
   pheatmap(mat,
            annotation_row = gene_anno,
            annotation_col = col_anno,
            show_rownames  = FALSE,
            show_colnames  = FALSE,
            cluster_rows   = TRUE,
            cluster_cols   = TRUE,
            color = colorRampPalette(
              c("steelblue", "white", "firebrick"))(100))
   ```

d. Pick one cluster and describe its expression pattern: consistently
   higher in the mutant or in the wild-type? Is the pattern consistent
   across all three replicates within each strain?

e. Compute and plot silhouette widths:
   ```r
   fviz_silhouette(silhouette(km_fit$cluster, dist(mat)))
   ```
   Are some clusters better-defined than others? What might a poorly
   defined cluster (low average silhouette width) mean biologically?

```{r}
# your code here
```

---

# Quick Reference Card

## Setup

```r
library(tidyverse)
library(SummarizedExperiment)
library(DESeq2)
library(fission)
library(factoextra)
library(cluster)
library(pheatmap)

data(fission)

# Subset to time zero
fission_t0 <- fission[, fission$minute == "0"]
```

## DESeq2 Workflow

```r
dds <- DESeqDataSet(fission_t0, design = ~ strain)
dds$strain <- relevel(dds$strain, ref = "wt")
keep       <- rowSums(counts(dds)) >= 10
dds        <- dds[keep, ]
deseq_dds  <- DESeq(dds)
res        <- results(deseq_dds)

res_df <- as.data.frame(res) |>
  rownames_to_column("gene_id") |>
  as_tibble() |>
  arrange(padj)
```

## Multiple Testing

```r
# Counts at different thresholds
sum(res_df$pvalue < 0.05, na.rm = TRUE)
sum(res_df$padj   < 0.05, na.rm = TRUE)

# Bonferroni
bonf <- p.adjust(res_df$pvalue, method = "bonferroni")
sum(bonf < 0.05, na.rm = TRUE)

# Genes excluded by independent filtering
sum(is.na(res_df$padj))
res_df |> filter(is.na(padj)) |> pull(baseMean) |> summary()

# Direction
res_df |>
  filter(padj < 0.05) |>
  count(direction = if_else(log2FoldChange > 0, "Up", "Down"))

# Fold change on original scale
2^log2FoldChange
```

## VST and Top Variable Genes

```r
vsd     <- vst(deseq_dds, blind = FALSE)
top_var <- order(rowVars(assay(vsd)), decreasing = TRUE)[1:300]
mat     <- t(scale(t(assay(vsd)[top_var, ])))
```

## Hierarchical Clustering

```r
d  <- dist(t(mat), method = "euclidean")
hc <- hclust(d, method = "complete")
plot(hc, labels = paste(colData(vsd)$strain,
                        colData(vsd)$sample, sep = "_"))
cutree(hc, k = 2)
table(Cluster = cutree(hc, k = 2), Strain = colData(vsd)$strain)
```

## K-Means Clustering

```r
# Choose K
fviz_nbclust(mat, kmeans, method = "silhouette", k.max = 8, nstart = 25)

# Fit
set.seed(2026)
km_fit <- kmeans(mat, centers = 3, nstart = 25)
table(km_fit$cluster)

# Silhouette plot
fviz_silhouette(silhouette(km_fit$cluster, dist(mat)))
```

## Heatmap

```r
gene_anno <- data.frame(
  Cluster   = paste0("C", km_fit$cluster),
  row.names = names(km_fit$cluster)
)
col_anno <- data.frame(
  Strain    = as.character(colData(vsd)$strain),
  row.names = colnames(vsd)
)
pheatmap(mat,
         annotation_row = gene_anno,
         annotation_col = col_anno,
         show_rownames  = FALSE,
         show_colnames  = FALSE,
         cluster_rows   = TRUE,
         cluster_cols   = TRUE,
         color = colorRampPalette(
           c("steelblue", "white", "firebrick"))(100))
```

## Key Terms

| Term | Meaning |
|---|---|
| Raw p-value | Do not use for calling significance across many tests |
| FWER | Probability of any false positive; Bonferroni controls this |
| FDR | Expected proportion of false discoveries; BH controls this |
| `padj` | BH-adjusted p-value from DESeq2 -- always use this |
| Independent filtering | Low-count genes excluded before BH; they get `padj = NA` |
| LFC > 0 | Higher expression in mutant relative to wild-type |
| K-means | Requires K upfront; scale data first; use `nstart > 1` |
| Hierarchical | Dendrogram; no K upfront; cut tree to choose clusters |
| Silhouette width | -1 to +1; higher = gene fits its cluster well |
