---
title: "Day 7: Model Selection and Cross-Validation"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| echo: false
library(tidyverse)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  filter(!is.na(sysBP), !is.na(age), !is.na(BMI), !is.na(totChol),
         !is.na(diabetes), !is.na(cigsPerDay), !is.na(glucose)) |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )

FITNESS <- tibble(
  age      = c(44,40,44,42,38,47,40,43,44,38,44,45,45,47,54,
               49,51,51,48,49,57,54,52,50,51,54,51,57,49,48,52),
  weight   = c(89.47,75.07,85.84,68.15,89.02,77.45,75.98,81.19,
               81.42,81.87,73.03,87.66,66.45,79.15,83.12,81.42,
               69.63,77.91,91.63,73.37,73.37,79.38,76.32,70.87,
               67.25,91.63,73.71,59.08,76.32,61.24,82.78),
  runtime  = c(11.37,10.07,8.65,8.17,9.22,11.63,11.95,10.85,
               13.08,8.63,10.13,14.03,11.12,10.60,10.33,8.95,
               10.95,10.00,10.25,10.08,12.63,11.17,9.63,8.92,
               11.08,12.88,10.47,9.93,9.40,11.50,10.50),
  restpulse = c(62,62,45,40,55,58,70,64,63,48,45,56,51,47,50,
               44,57,48,48,67,58,62,48,48,48,44,59,49,56,52,53),
  runpulse = c(178,185,156,166,178,176,176,162,174,170,168,186,
               176,162,166,180,168,162,162,168,174,156,164,146,
               172,168,186,148,186,170,170),
  maxpulse = c(182,185,168,172,180,176,180,170,176,186,168,192,
               176,164,170,185,172,168,164,168,176,165,166,155,
               172,172,188,155,188,176,172),
  Y        = c(44.609,45.313,54.297,59.571,49.874,44.811,45.681,
               49.091,39.442,60.055,50.541,37.388,44.754,47.273,
               51.855,49.156,40.836,46.672,46.774,50.388,39.407,
               46.080,45.441,54.625,45.118,39.203,45.790,50.545,
               48.730,47.920,47.467)
)
```

# Overview

This document accompanies the Day 7 morning lecture slides. It demonstrates
the practical R tools for model selection and cross-validation using two
datasets:

- **Framingham Heart Study**: comparing models predicting systolic blood
  pressure
- **FITNESS dataset**: stepwise selection with multiple physiological
  predictors

---

# Why $R^2$ Is Not Enough

A natural first instinct is to pick the model with the highest $R^2$. Let's
see why this fails.

```{r}
# Fit models with increasing numbers of predictors
m1 <- lm(sysBP ~ age, data = framingham)
m2 <- lm(sysBP ~ age + BMI, data = framingham)
m3 <- lm(sysBP ~ age + BMI + totChol, data = framingham)
m4 <- lm(sysBP ~ age + BMI + totChol + cigsPerDay, data = framingham)
m5 <- lm(sysBP ~ age + BMI + totChol + cigsPerDay + glucose, data = framingham)
m6 <- lm(sysBP ~ age + BMI + totChol + cigsPerDay + glucose + diabetes, data = framingham)

# R-squared always goes up
tibble(
  predictors = 1:6,
  R_squared  = c(summary(m1)$r.squared, summary(m2)$r.squared,
                 summary(m3)$r.squared, summary(m4)$r.squared,
                 summary(m5)$r.squared, summary(m6)$r.squared)
) |>
  ggplot(aes(x = predictors, y = R_squared)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  scale_x_continuous(breaks = 1:6) +
  labs(x = "Number of predictors", y = expression(R^2),
       title = expression(R^2 ~ "always increases with more predictors"),
       subtitle = "Even if the added predictors are not useful") +
  theme_minimal()
```

$R^2$ goes up with every predictor we add, regardless of whether it
actually improves prediction. We need criteria that **penalize**
unnecessary complexity.

---

# AIC and BIC

## Comparing Models

AIC and BIC add a penalty for the number of parameters. Lower values are
better.

```{r}
AIC(m1, m2, m3, m4, m5, m6)
```

```{r}
BIC(m1, m2, m3, m4, m5, m6)
```

Notice: AIC keeps decreasing through model 6, but BIC levels off earlier.
BIC penalizes complexity more heavily and tends to prefer simpler models.

## Interpreting AIC Differences

```{r}
aic_values <- AIC(m1, m2, m3, m4, m5, m6)
aic_values$delta_AIC <- aic_values$AIC - min(aic_values$AIC)
aic_values
```

A $\Delta$AIC less than 2 means the models are essentially equivalent. A
$\Delta$AIC greater than 10 is strong evidence for the lower-AIC model. When
models are close, prefer the simpler one.

## Visualizing AIC and BIC Together

```{r}
#| fig-width: 7
#| fig-height: 4
# Extract the AIC and BIC values correctly
aic_vals <- c(AIC(m1), AIC(m2), AIC(m3), AIC(m4), AIC(m5), AIC(m6))
bic_vals <- c(BIC(m1), BIC(m2), BIC(m3), BIC(m4), BIC(m5), BIC(m6))

comparison <- tibble(
  predictors = rep(1:6, 2),
  criterion  = rep(c("AIC", "BIC"), each = 6),
  value      = c(aic_vals, bic_vals)
)

ggplot(comparison, aes(x = predictors, y = value, color = criterion)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:6) +
  scale_color_manual(values = c("AIC" = "steelblue", "BIC" = "firebrick")) +
  labs(x = "Number of predictors", y = "Criterion value (lower is better)",
       title = "AIC vs. BIC Across Nested Models",
       color = "Criterion") +
  theme_minimal()
```

---

# Stepwise Selection with `step()`

## FITNESS Dataset: Full Model

We use the FITNESS dataset from Day 6. Start with a model containing all six
predictors.

```{r}
model_full <- lm(Y ~ age + weight + runtime + restpulse + runpulse + maxpulse,
                 data = FITNESS)
summary(model_full)
```

## Stepwise Selection (Both Directions)

`step()` iteratively adds and removes predictors to minimize AIC.

```{r}
model_step_both <- step(model_full, direction = "both", trace = 1)
```

The trace output shows exactly what the algorithm does at each step: which
predictor it considers adding or removing, the resulting AIC, and which
change it selects. The final model is the one where no single add or remove
improves AIC.

```{r}
summary(model_step_both)
```

## Comparing Directions

Forward selection and backward elimination can give different results.

```{r}
# Forward: start from intercept only
model_null <- lm(Y ~ 1, data = FITNESS)
model_step_fwd <- step(model_null,
                       scope = list(lower = model_null, upper = model_full),
                       direction = "forward", trace = 0)

# Backward: start from full model
model_step_bwd <- step(model_full, direction = "backward", trace = 0)

# Compare
tibble(
  direction = c("Forward", "Backward", "Both"),
  formula   = c(deparse(formula(model_step_fwd)),
                deparse(formula(model_step_bwd)),
                deparse(formula(model_step_both))),
  AIC       = c(AIC(model_step_fwd), AIC(model_step_bwd), AIC(model_step_both)),
  R_squared = c(summary(model_step_fwd)$r.squared,
                summary(model_step_bwd)$r.squared,
                summary(model_step_both)$r.squared)
)
```

::: {.callout-note}
If the three directions select the same model, you can be more confident in
the result. If they disagree, it suggests that some predictors are marginal
and the choice may not matter much practically.
:::

## Formally Comparing Nested Models

We can also use the partial F-test (`anova()`) to compare any two nested
models directly.

```{r}
# Does the full model improve significantly over the stepwise model?
anova(model_step_both, model_full)
```

A non-significant p-value means the extra predictors in the full model do not
meaningfully improve fit --- we prefer the simpler stepwise model.

---

# Cross-Validation

## The Idea

AIC estimates prediction error using a formula. Cross-validation estimates it
by **actually predicting held-out data**. We implement 5-fold CV from
scratch, then compare to AIC.

## Setting Up 5-Fold CV

```{r}
set.seed(42)   # for reproducibility

# Assign each observation to a fold
n <- nrow(framingham)
K <- 5
framingham$fold <- sample(rep(1:K, length.out = n))

table(framingham$fold)
```

## Writing a CV Function

```{r}
cv_mse <- function(formula, data, K = 5) {
  mse_per_fold <- numeric(K)

  for (k in 1:K) {
    train <- data[data$fold != k, ]
    test  <- data[data$fold == k, ]

    model <- lm(formula, data = train)
    preds <- predict(model, newdata = test)
    mse_per_fold[k] <- mean((test$sysBP - preds)^2)
  }

  mean(mse_per_fold)
}
```

This function:

1. Loops over the $K$ folds.
2. For each fold, fits the model on the training data (everything except
   that fold).
3. Predicts the held-out fold.
4. Computes the mean squared error (MSE) for that fold.
5. Returns the average MSE across all folds.

## Comparing Models with CV

```{r}
cv_results <- tibble(
  model       = c("age only",
                   "age + BMI",
                   "age + BMI + totChol",
                   "age + BMI + totChol + cigsPerDay",
                   "age + BMI + totChol + cigsPerDay + glucose",
                   "age + BMI + totChol + cigsPerDay + glucose + diabetes"),
  predictors  = 1:6,
  cv_mse      = c(
    cv_mse(sysBP ~ age, framingham),
    cv_mse(sysBP ~ age + BMI, framingham),
    cv_mse(sysBP ~ age + BMI + totChol, framingham),
    cv_mse(sysBP ~ age + BMI + totChol + cigsPerDay, framingham),
    cv_mse(sysBP ~ age + BMI + totChol + cigsPerDay + glucose, framingham),
    cv_mse(sysBP ~ age + BMI + totChol + cigsPerDay + glucose + diabetes, framingham)
  )
)

cv_results
```

## Visualizing CV Error

```{r}
#| fig-width: 7
#| fig-height: 4
ggplot(cv_results, aes(x = predictors, y = cv_mse)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 3) +
  geom_point(data = filter(cv_results, cv_mse == min(cv_mse)),
             color = "firebrick", size = 4) +
  scale_x_continuous(breaks = 1:6) +
  labs(x = "Number of predictors", y = "5-Fold CV Mean Squared Error",
       title = "Cross-Validation Error Across Nested Models",
       subtitle = "Red point marks the model with lowest CV error") +
  theme_minimal()
```

::: {.callout-important}
Compare this plot to the $R^2$ plot from Section 1. $R^2$ always goes up.
CV error goes down and then **levels off or increases** as overfitting
begins. This is why CV is a better guide for model selection than $R^2$.
:::

## AIC vs. CV: Do They Agree?

```{r}
cv_results$AIC <- c(AIC(m1), AIC(m2), AIC(m3), AIC(m4), AIC(m5), AIC(m6))

cv_results |>
  select(model, predictors, cv_mse, AIC) |>
  mutate(
    best_cv  = if_else(cv_mse == min(cv_mse), "<-- best CV", ""),
    best_aic = if_else(AIC == min(AIC), "<-- best AIC", "")
  )
```

In many practical situations, AIC and CV will select the same model or very
similar ones. When they disagree, it is usually on marginal predictors that
don't change the conclusions meaningfully.

---

# Using `rsample` for Cross-Validation

The manual loop above is instructive, but in practice the `rsample` package
makes CV cleaner and less error-prone.

```{r}
library(rsample)

set.seed(42)
folds <- vfold_cv(framingham, v = 5)
folds
```

Each fold object contains a training/test split. We can map over them:

```{r}
cv_rsample <- function(formula, folds) {
  mse_vals <- map_dbl(folds$splits, function(split) {
    train <- analysis(split)
    test  <- assessment(split)
    model <- lm(formula, data = train)
    preds <- predict(model, newdata = test)
    mean((test$sysBP - preds)^2)
  })

  tibble(
    mean_mse = mean(mse_vals),
    sd_mse   = sd(mse_vals)
  )
}

cv_rsample(sysBP ~ age + BMI + totChol, folds)
```

The standard deviation tells you how variable the CV error is across folds.
If it is large relative to the mean, the estimate is noisy and you should
interpret model comparisons cautiously.

---

# Putting It All Together

Here is the complete model selection workflow applied to the Framingham data:

**Step 1: Check variable form** (Day 6 — loess smoothers for each predictor).

**Step 2: Fit a full model** with all candidate predictors.

```{r}
full <- lm(sysBP ~ age + BMI + totChol + cigsPerDay + glucose + diabetes,
           data = framingham)
summary(full)
```

**Step 3: Run stepwise selection** with AIC.

```{r}
selected <- step(full, direction = "both", trace = 0)
summary(selected)
```

**Step 4: Compare with AIC and a formal F-test.**

```{r}
AIC(full, selected)
anova(selected, full)
```

**Step 5: Validate with cross-validation.**

```{r}
set.seed(42)
folds <- vfold_cv(framingham, v = 5)

cv_full     <- cv_rsample(sysBP ~ age + BMI + totChol + cigsPerDay + glucose + diabetes, folds)
cv_selected <- cv_rsample(formula(selected), folds)

bind_rows(
  cv_full     |> mutate(model = "Full"),
  cv_selected |> mutate(model = "Stepwise-selected")
) |>
  select(model, mean_mse, sd_mse)
```

**Step 6: Run diagnostics on the final model** (Day 6 workflow).

```{r}
#| fig-width: 8
#| fig-height: 7
par(mfrow = c(2, 2))
plot(selected)
par(mfrow = c(1, 1))
```

**Step 7: Interpret the final model.**

```{r}
confint(selected)
```

::: {.callout-tip}
This seven-step workflow — check variable form, fit full model, stepwise
selection, AIC comparison, cross-validation, diagnostics, interpretation —
is a template you can apply to any regression analysis.
:::
