---
title: "Day 7 (PM) Lab: Model Selection and Cross-Validation"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This lab puts the Day 7 morning material into practice. You will work through
the full model selection workflow --- from comparing candidate models with
AIC/BIC, to automated stepwise selection, to cross-validation --- using the
Framingham Heart Study dataset.

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** Follow along as the instructor
  builds a model for total cholesterol using the complete model selection
  workflow.
- **Part 2 (on your own, ~90 min):** Six exercises. Create a new `.qmd`
  file, work through each exercise, render when done.

The **Quick Reference Card** in Section 3 has the key functions.

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(rsample)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  filter(!is.na(totChol), !is.na(age), !is.na(BMI), !is.na(sysBP),
         !is.na(diaBP), !is.na(cigsPerDay), !is.na(glucose),
         !is.na(heartRate)) |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```

## Goal: Predict Total Cholesterol

We want to build a model predicting `totChol` from available continuous and
categorical predictors. This is a different outcome from what we used this
morning (`sysBP`), so students see the workflow applied to a fresh problem.

## Step 1: Candidate Models

```{r}
m_small  <- lm(totChol ~ age, data = framingham)
m_medium <- lm(totChol ~ age + BMI + male, data = framingham)
m_large  <- lm(totChol ~ age + BMI + male + sysBP + diaBP + cigsPerDay +
                  glucose + heartRate + currentSmoker, data = framingham)
```

## Step 2: Compare with AIC and BIC

```{r}
AIC(m_small, m_medium, m_large)
BIC(m_small, m_medium, m_large)
```

## Step 3: Stepwise Selection

```{r}
m_step <- step(m_large, direction = "both", trace = 0)
summary(m_step)
AIC(m_large, m_step)
```

## Step 4: Cross-Validate the Finalists

```{r}
set.seed(123)
folds <- vfold_cv(framingham, v = 5)

cv_mse <- function(formula, folds) {
  mse_vals <- map_dbl(folds$splits, function(split) {
    train <- analysis(split)
    test  <- assessment(split)
    model <- lm(formula, data = train)
    preds <- predict(model, newdata = test)
    mean((test$totChol - preds)^2)
  })
  tibble(mean_mse = mean(mse_vals), sd_mse = sd(mse_vals))
}

bind_rows(
  cv_mse(formula(m_small),  folds) |> mutate(model = "Small (age)"),
  cv_mse(formula(m_medium), folds) |> mutate(model = "Medium"),
  cv_mse(formula(m_step),   folds) |> mutate(model = "Stepwise"),
  cv_mse(formula(m_large),  folds) |> mutate(model = "Large (full)")
) |>
  select(model, mean_mse, sd_mse) |>
  arrange(mean_mse)
```

## Step 5: Diagnostics on the Final Model

```{r}
#| fig-width: 8
#| fig-height: 7
par(mfrow = c(2, 2))
plot(m_step)
par(mfrow = c(1, 1))
```

The workflow in five steps: candidate models → AIC/BIC → stepwise → CV →
diagnostics. This is the same template from this morning, applied to a new
outcome.

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document: File → New File → Quarto Document.
Title it `"Day 7 Lab: Model Selection"`, select HTML, click **Create**.

**Step 2.** Replace the template content with this YAML and setup chunk:

````markdown
---
title: "Day 7 Lab: Model Selection and Cross-Validation"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{{r}}
library(tidyverse)
library(rsample)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  filter(!is.na(sysBP), !is.na(age), !is.na(BMI), !is.na(totChol),
         !is.na(diaBP), !is.na(cigsPerDay), !is.na(glucose),
         !is.na(heartRate)) |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```
````

**Step 3.** Work through Exercises 1--6 in order. For each one, add a code
chunk and 2--3 sentences describing what you found.

**Step 4.** Render (`Ctrl+Shift+K` / `Cmd+Shift+K`) when finished.

::: {.callout-tip}
Check the **Quick Reference Card** in Section 3 before asking for help.
The most common issues are: (1) comparing AIC across models fit on different
subsets of the data (this is invalid --- all models must use the same rows),
and (2) forgetting `set.seed()` before cross-validation (results will change
each run).
:::

::: {.callout-note}
Exercises 1--4 cover the core material. Exercise 5 requires combining CV
with model comparison in a single analysis. Exercise 6 is a challenge that
brings in Day 6 material (non-linear terms). Don't worry if you don't
reach it.
:::

---

**Exercise 1 --- AIC and BIC Basics.**

Build four models predicting `sysBP`:

- Model A: `sysBP ~ age`
- Model B: `sysBP ~ age + BMI`
- Model C: `sysBP ~ age + BMI + diaBP`
- Model D: `sysBP ~ age + BMI + diaBP + totChol + cigsPerDay + glucose`

a. Compute the $R^2$ of each model. Does $R^2$ always increase? Create a
   simple plot of $R^2$ vs. the number of predictors.

b. Compute AIC and BIC for all four models using `AIC()` and `BIC()`. Which
   model does AIC prefer? Which does BIC prefer? Do they agree?

c. Compute the $\Delta$AIC for each model relative to the best. Which models
   are within 2 AIC units of the best (i.e., essentially equivalent)?

d. In 2--3 sentences, explain why $R^2$ and AIC can disagree about which
   model is best.

```{r}
# your code here
```

---

**Exercise 2 --- Stepwise Selection.**

a. Fit the full model: `sysBP ~ age + BMI + totChol + cigsPerDay + glucose +
   diaBP + heartRate + male + currentSmoker`. Run `step()` with
   `direction = "both"` and `trace = 1`. Watch the output carefully ---
   which predictors are considered at each step? Which are dropped?

b. Run `step()` again with `direction = "forward"` starting from the
   intercept-only model. Use the `scope` argument to define the search space
   (see the morning .qmd for the syntax). Does forward selection give the
   same result as stepwise?

c. Run `step()` with `direction = "backward"`. Does it agree with the other
   two?

d. Create a summary table comparing the three selected models: their
   formulas, AIC values, and $R^2$. If they differ, which would you choose
   and why?

```{r}
# your code here
```

---

**Exercise 3 --- Formally Comparing Nested Models.**

Using the stepwise-selected model from Exercise 2 as your "reduced" model
and the full model as your "full" model:

a. Use `anova(reduced, full)` to run a partial F-test. Is the full model
   significantly better than the stepwise model?

b. Report the p-value and interpret it in one sentence.

c. Report `AIC()` for both models. Does the AIC comparison agree with the
   F-test? Should it?

d. Based on these comparisons, which model would you report as your final
   model and why?

```{r}
# your code here
```

---

**Exercise 4 --- Cross-Validation from Scratch.**

a. Assign each observation in the Framingham data to one of 5 folds using
   `sample(rep(1:5, length.out = n))`. Remember to use `set.seed()` for
   reproducibility.

b. Write a `for` loop that implements 5-fold CV for a single model. For each
   fold: fit the model on the training data (folds $\neq k$), predict the
   test data (fold $= k$), and compute the MSE. Store the 5 MSE values in a
   vector.

c. Apply your loop to three models:
   - `sysBP ~ age`
   - `sysBP ~ age + BMI + diaBP`
   - The stepwise-selected model from Exercise 2

   Report the mean CV-MSE and standard deviation for each.

d. Which model has the lowest CV error? Is this the same model that AIC
   preferred in Exercise 1? Explain in 1--2 sentences.

e. Create a plot showing the mean CV-MSE for each model with error bars
   ($\pm$ 1 standard deviation). Use `geom_point()` and `geom_errorbar()`.

```{r}
# your code here
```

---

**Exercise 5 --- The Full Workflow.**

Build the best model you can for predicting `diaBP` (diastolic blood
pressure) --- a different outcome from what we have been using. Apply the
complete workflow:

a. **Check variable form:** Use `pivot_longer()` and `facet_wrap()` with
   loess smoothers to inspect the relationship between `diaBP` and each
   candidate continuous predictor (`age`, `BMI`, `sysBP`, `totChol`,
   `cigsPerDay`, `glucose`, `heartRate`). Which predictors show a clear
   relationship?

b. **Fit a full model** using all the predictors you identified plus `male`
   and `currentSmoker` as categorical predictors.

c. **Run stepwise selection** with `step()`. Report the selected model.

d. **Cross-validate** the full model, the stepwise model, and a small model
   of your choosing using 5-fold CV with `rsample::vfold_cv()`. Report
   mean CV-MSE and SD for each.

e. **Run diagnostics** on your final chosen model. Are there any concerns?

f. **Report and interpret** the coefficients and 95% confidence intervals of
   the final model. For each predictor, write one sentence explaining what
   the coefficient means in clinical terms.

```{r}
# your code here
```

---

**Exercise 6 (Challenge) --- Non-Linear Terms and Model Selection.**

This exercise combines Day 6 material (non-linear variable form) with
Day 7 model selection.

a. Recall from the Day 6 lab that `glucose` may have a non-linear
   relationship with blood pressure. Fit two competing models for `sysBP`:
   - `m_linear`: `sysBP ~ age + BMI + diaBP + glucose`
   - `m_spline`: `sysBP ~ age + BMI + diaBP + splines::ns(glucose, df = 3)`

   Compare them with AIC. Does the spline term for glucose improve the
   model?

b. Now add `heartRate` to both models and repeat the AIC comparison. Does the
   answer change once `heartRate` is in the model?

c. Cross-validate both the linear and spline versions (with `heartRate`)
   using 5-fold CV. Report mean CV-MSE and SD for each. Does CV agree with
   AIC?

d. Apply stepwise selection to the spline model. Does `step()` keep the
   spline term? Note: `step()` treats `ns(glucose, df = 3)` as a single
   block --- it either keeps all the spline terms or drops them all.

e. Run diagnostics on the better-fitting model. Compare the Residuals vs.
   Fitted plot to the linear version. Does allowing non-linearity in glucose
   reduce any patterning in the residuals?

f. Based on everything above, write a 3--4 sentence paragraph summarizing
   your final model choice and justification --- as if this were the methods
   section of a short paper.

```{r}
# your code here
```

---

# Quick Reference Card

## Information Criteria

| Function | What it does |
|---|---|
| `AIC(m1, m2, m3)` | Compare models by AIC --- lower is better |
| `BIC(m1, m2, m3)` | Compare models by BIC --- lower is better, stronger penalty |

**Interpreting $\Delta$AIC:** $<2$ = equivalent; $2$--$10$ = some evidence; $>10$ = strong evidence.

**Key rule:** AIC/BIC can only compare models fit on the **same data** (same rows, same outcome).

## Stepwise Selection

```r
# Stepwise (both directions) --- recommended default
step(full_model, direction = "both", trace = 0)

# Forward selection --- start from intercept
model_null <- lm(y ~ 1, data = df)
step(model_null,
     scope = list(lower = model_null, upper = full_model),
     direction = "forward", trace = 0)

# Backward elimination --- start from full model
step(full_model, direction = "backward", trace = 0)
```

## Comparing Nested Models

```r
# Partial F-test: does the full model improve over the reduced?
anova(model_reduced, model_full)
```

## Cross-Validation (Manual)

```r
set.seed(42)
n <- nrow(df)
df$fold <- sample(rep(1:5, length.out = n))

mse_per_fold <- numeric(5)
for (k in 1:5) {
  train <- df[df$fold != k, ]
  test  <- df[df$fold == k, ]
  model <- lm(y ~ x1 + x2, data = train)
  preds <- predict(model, newdata = test)
  mse_per_fold[k] <- mean((test$y - preds)^2)
}
mean(mse_per_fold)  # CV-MSE
sd(mse_per_fold)    # SD across folds
```

## Cross-Validation (with rsample)

```r
library(rsample)
set.seed(42)
folds <- vfold_cv(df, v = 5)

mse_vals <- map_dbl(folds$splits, function(split) {
  train <- analysis(split)
  test  <- assessment(split)
  model <- lm(y ~ x1 + x2, data = train)
  preds <- predict(model, newdata = test)
  mean((test$y - preds)^2)
})

mean(mse_vals)  # CV-MSE
sd(mse_vals)    # SD across folds
```

## Error Bar Plot for CV Results

```r
results |>
  ggplot(aes(x = model, y = mean_mse)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mse - sd_mse,
                    ymax = mean_mse + sd_mse), width = 0.2)
```

## Useful Reminders

- Always use `set.seed()` before any CV so results are reproducible
- `AIC()` and `BIC()` require models fit on the **same rows** --- filter NAs first
- `step()` with `trace = 0` suppresses output; `trace = 1` shows the algorithm
- When AIC and CV agree, you can be confident; when they disagree, trust CV
- After model selection, always run diagnostics on the final model (Day 6 workflow)
