---
title: "Day 3 Lab: Importing, Tidying, and Joining"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(readxl)
library(nycflights13)
library(this.path)
this.dir <- this.dir()
setwd(this.dir)
```

**Exercise 1 — Reading and Cleaning.**
Read `framingham.csv` from the `data/` folder using `read_csv()`. The dataset
uses blank cells for missing values, which `read_csv()` handles automatically.
After reading in:

a. Use `str()` to check column types. Which variables are stored as `<dbl>` but
   should be factors?

Option 1: setting the working directory and just using `framingham.csv`.

```{r}
setwd('/Users/mclainfamily/Library/CloudStorage/OneDrive-UniversityofSouthCarolina/Teaching/Summer Workshop/INBRE-Bios-Summer-26/data')
framingham <- read_csv("framingham.csv")
```


Option 2: not setting the working directory and using the full file path.

```{r}
framingham <- read_csv("/Users/mclainfamily/Library/CloudStorage/OneDrive-UniversityofSouthCarolina/Teaching/Summer Workshop/INBRE-Bios-Summer-26/data/framingham.csv")
head(framingham)
```

Option 3: setting the working directory and using a partial file path. 

```{r}
this.dir <- this.dir()
setwd(this.dir)
framingham <- read_csv("../../data/framingham.csv")
head(framingham)
str(framingham)
```   

::: {.callout-note}
For option 3 to work, you need to have your lab saved in `INBRE-Bios-Summer-26-main/week1-foundations/day03-tidyverse/`
:::


b. Convert at least four binary or categorical variables to factors with
   meaningful labels using `mutate()` and `factor()` (you did this on Day 2 ---
   copy that block if you still have it).
   
   
As we discussed in yesterdays AM session, many of the categorical variables are coded as numeric (e.g., male, currentSmoker, education)

```{r}
framingham <- framingham |>
  mutate(
    male            = factor(male,            levels = c(0, 1),
                                              labels = c("Female", "Male")),
    currentSmoker   = factor(currentSmoker,   levels = c(0, 1),
                                              labels = c("Non-smoker", "Smoker")),
    BPMeds          = factor(BPMeds,          levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentStroke = factor(prevalentStroke, levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentHyp    = factor(prevalentHyp,    levels = c(0, 1),
                                              labels = c("No", "Yes")),
    diabetes        = factor(diabetes,        levels = c(0, 1),
                                              labels = c("No", "Yes")),
    TenYearCHD      = factor(TenYearCHD,      levels = c(0, 1),
                                              labels = c("No", "Yes")),
    education       = factor(education,       levels = 1:4,
                                              labels = c("Some HS", "HS Diploma",
                                                         "Some College",
                                                         "College Degree"))
  )
```


   
c. Write the cleaned dataset to a new file called `framingham_clean.csv` using
   `write_csv()`.

Writing is easy. This file will be saved in your working directory.

```{r}
getwd()
write_csv(framingham,"framingham_clean.csv")
```



---

**Exercise 4 — Left Joins.**
Using `flights` and the other `nycflights13` tables:

a. Add the full airline name (from `airlines`) to `flights`. Then compute the
   mean departure delay by airline name. Which airline has the worst average
   departure delay?

```{r}
flights |>
left_join(airlines) |> 
  group_by(name) |> 
  summarise(
    mn_dep_delay = mean(dep_delay, na.rm = TRUE)
  ) |> 
  arrange(desc(mn_dep_delay))
```

b. Add the plane's number of seats (from `planes`) to `flights` using
   `left_join()` with `by = "tailnum"`. What proportion of flights are missing
   seat information (i.e., had no matching record in `planes`)?


```{r}
flights |>
left_join(planes, by = "tailnum") |> 
  mutate(
    miss_seat = is.na(seats)
  ) |> 
  summarise(
    mn_miss_seat = mean(miss_seat, na.rm = TRUE)
  ) 
```

c. Add the destination airport's full name and time zone (from `airports`) to
   `flights`. Which time zone has the most flight destinations from NYC?

```{r}
flights |>
left_join(airports, by = c("dest" = "faa")) |> 
  group_by(tzone) |> 
  summarise(
    total = n()
  ) |> 
  arrange(
    desc(total)
  )

```

---

**Exercise 5 — Joins and Grouped Summaries.**
This exercise combines joining with the `group_by()` / `summarize()` workflow
from Day 2.

a. Join `flights` with `weather` (using `origin` and `time_hour` as the
   compound key) to add temperature (`temp`) and visibility (`visib`) to each
   flight.

```{r}
flights |>
left_join(weather |> select(origin, time_hour, temp, visib)) 
```

b. Compute the mean arrival delay by visibility level. Round visibility to the
   nearest integer first using `mutate(visib = round(visib))`. Sort from worst
   to best delay. Is there a relationship between visibility and delays?

```{r}
flights |>
left_join(weather |> select(origin, time_hour, temp, visib)) |> 
  mutate(visib = round(visib)) |> 
  group_by(visib) |> 
  summarise(
    mn_delay = mean(arr_delay, na.rm = TRUE)
  ) |> 
  arrange(
    desc(mn_delay)
  )
```

Yes, there definetly appears to be a relationship between visibility and the delay. Lower visibility the higher the delay.

c. Repeat (b) but compute separately for each origin airport. Use
   `group_by(origin, visib)`. Describe in 1--2 sentences what you find.
   
```{r}
flights |>
left_join(weather |> select(origin, time_hour, temp, visib)) |> 
  mutate(visib = round(visib)) |> 
  group_by(origin,visib) |> 
  summarise(
    mn_delay = mean(arr_delay, na.rm = TRUE)
  ) |> 
  arrange(
    desc(mn_delay)
  )
```

Visibility equals 1 has the largest delay for all airports. LGA seems to have less (lower) delays for visibility 0 and 2 than EWR and JFK.
