---
title: "Day 14 (PM) Lab: Principal Component Analysis"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This afternoon you will practice PCA on a new dataset: chemical measurements
of 178 wines from three cultivars grown in the same region of Italy. Each wine
has 13 continuous measurements including alcohol content, acidity, phenolic
compounds, color intensity, and other chemical properties.

The dataset has no missing values, the variables are on different scales (so
scaling is clearly necessary), and the three cultivars give you a natural
grouping to validate the PCA results against -- without using that grouping
during the PCA itself. It is a clean, well-studied dataset that makes PCA
results easy to check.

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** Load the data and run through the
  setup together.
- **Part 2 (on your own, ~90 min):** Five exercises taking you through a
  complete PCA workflow.

::: {.callout-note}
The wine dataset is in the `pgirmess` package or can be loaded directly.
We load it from a built-in copy below -- no installation needed.
:::

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(this.path)
library(factoextra)
library(patchwork)
setwd(this.dir())

select <- dplyr::select
filter <- dplyr::filter
```

## Load the Wine Data

```{r}
# The wine dataset is available in the rattle package,
# or we can read it directly from the UCI repository
# Here we use the version bundled with FactoMineR
data(wine, package = "FactoMineR")

# The FactoMineR version has a slightly different structure
# Let's inspect it
head(wine)
dim(wine)
```

::: {.callout-note}
If `FactoMineR::wine` has a different structure on your machine, use this
alternative which loads the UCI wine data directly:

```r
url <- paste0("https://archive.ics.uci.edu/ml/",
              "machine-learning-databases/wine/wine.data")
wine_raw <- read.csv(url, header = FALSE)
names(wine_raw) <- c("Cultivar", "Alcohol", "MalicAcid", "Ash",
  "AlcalinityAsh", "Magnesium", "TotalPhenols", "Flavanoids",
  "NonflavanoidPhenols", "Proanthocyanins", "ColorIntensity",
  "Hue", "OD280_OD315", "Proline")
wine_raw$Cultivar <- factor(wine_raw$Cultivar)
```
:::

For the exercises we will use a clean version with clear variable names:

```{r}
# Build a clean tibble with the 13 chemical variables + cultivar label
# Using the UCI wine data structure (adjust column names if your load differs)
wine_clean <- wine |>
  as_tibble() |>
  rename_with(tolower)

# Check the structure
glimpse(wine_clean)
```

```{r}
# Summary statistics for the chemical variables
wine_clean |>
  select(where(is.numeric)) |>
  summary()
```

The variables are on very different scales: `proline` ranges in the hundreds
while `nonflavano_phenols` is near zero. **Scaling is essential** before PCA.

## Step 1: Scale and Run PCA

```{r}
# Extract the numeric variables (drop the cultivar label for PCA)
X <- wine_clean |>
  select(where(is.numeric)) |>
  as.matrix()

# Standardize
X_s <- scale(X)

# Run PCA
pca_wine <- prcomp(X_s, center = FALSE, scale. = FALSE)
```

## Step 2: Quick Scree Plot

```{r}
fviz_eig(pca_wine, addlabels = TRUE,
         barfill = "steelblue", barcolor = "gray40") +
  labs(title = "Scree Plot: Wine PCA") +
  theme_minimal()
```

The first two PCs together explain roughly 55% of the total variance, and
the curve shows a clear elbow after PC2 or PC3 -- much sharper than the NHANES
data from the morning, where variance was spread more evenly.

## Step 3: Score Plot

```{r}
#| fig-width: 6
#| fig-height: 5
fviz_pca_ind(pca_wine,
             geom.ind   = "point",
             habillage  = wine_clean$label,
             addEllipses = TRUE,
             repel      = TRUE) +
  labs(title = "PCA Score Plot: Wine Cultivars") +
  theme_minimal()
```

The three cultivars separate cleanly along PC1 and PC2 even though we did not
use the cultivar labels during PCA. This confirms that the chemical variables
genuinely capture cultivar differences -- exactly what you want to see.

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document. Title it
`"Day 14 Lab: Principal Component Analysis"`.

**Step 2.** Copy the setup and data-loading chunks from Part 1. You need
`pca_wine`, `X_s`, and `wine_clean` available.

**Step 3.** Work through Exercises 1--5. Add 2--3 sentences of interpretation
after each result.

**Step 4.** Render when finished.

::: {.callout-note}
Exercises 1--3 cover the core PCA output. Exercise 4 explores the loadings
in depth. Exercise 5 is a challenge that uses the PC scores in a regression
model.
:::

---

**Exercise 1 -- Variance Explained.**

a. Compute the proportion of variance explained by each PC. The formula is:
   ```r
   var_exp <- pca_wine$sdev^2 / sum(pca_wine$sdev^2)
   ```
   Print a table with the PC number, individual variance explained (%), and
   cumulative variance explained (%).

b. How many PCs are needed to explain at least 70% of the total variance?
   At least 90%?

c. Apply Kaiser's rule: keep PCs with eigenvalue greater than 1. The
   eigenvalues are `pca_wine$sdev^2`. How many PCs would Kaiser's rule
   suggest keeping?

d. Make a scree plot using `fviz_eig()`. Where does the "elbow" appear?
   Does it agree with the answers from parts b and c? Which criterion
   would you use, and why?

```{r}
# your code here
```

---

**Exercise 2 -- Score Plot and Cultivar Separation.**

a. Produce a score plot (observations in PC space) using `fviz_pca_ind()`.
   Color points by cultivar (`habillage = wine_clean$label`) and add
   confidence ellipses (`addEllipses = TRUE`).

b. Do the three cultivars separate in the first two PCs? Which cultivar
   is most distinct from the other two?

c. PC1 runs left to right on the plot. Look at where each cultivar falls
   along PC1. Which cultivar has the highest PC1 scores (furthest right)?
   Which has the lowest (furthest left)?

d. Now make a second score plot using PC1 and PC3 instead of PC1 and PC2.
   Pass `axes = c(1, 3)` to `fviz_pca_ind()`. Does adding PC3 improve or
   worsen the separation between cultivars?

```{r}
# your code here
```

---

**Exercise 3 -- Loadings.**

a. Extract the loadings matrix from `pca_wine$rotation`. Print the loadings
   for the first three PCs, rounded to 3 decimal places.

b. For PC1: which three variables have the largest absolute loadings? What
   do these variables have in common chemically? Based on this, what
   biological property might PC1 be capturing?

c. For PC2: which three variables have the largest absolute loadings? Are
   these the same variables as PC1 or different? What might PC2 be
   capturing?

d. Make a loading bar plot for PC1 and PC2 (see the Quick Reference Card
   for the ggplot2 code). Does the pattern of loadings match your
   interpretation from parts b and c?

e. Produce a **biplot** using `fviz_pca_biplot()`. Color observations by
   cultivar. Describe: which variables have arrows pointing in the same
   direction as the cultivar that has the highest PC1 scores?

```{r}
# your code here
```

---

**Exercise 4 -- Effect of Scaling.**

This exercise demonstrates why scaling matters.

a. Run PCA **without** scaling:
   ```r
   pca_unscaled <- prcomp(X, center = TRUE, scale. = FALSE)
   ```
   Make a scree plot. What percentage of variance does PC1 explain in the
   unscaled PCA? Compare this to the scaled version from Exercise 1.

b. Look at which variable has the largest variance in the raw data:
   ```r
   apply(X, 2, var) |> sort(decreasing = TRUE)
   ```
   Which variable dominates? Does this match which variables have the
   largest loadings on PC1 in the unscaled PCA?

c. Make score plots for both the scaled and unscaled PCA, colored by
   cultivar. Do the cultivars still separate in the unscaled version?

d. Write 2--3 sentences explaining why you should (almost) always scale
   before PCA, using what you observed in parts a--c as evidence.

```{r}
# your code here
```

---

**Exercise 5 (Challenge) -- PC Scores in Logistic Regression.**

The three wine cultivars can be predicted from the chemical measurements.
This exercise uses the PC scores from the scaled PCA as predictors in a
multinomial logistic regression -- a direct application of PCR.

a. Extract the scores for the first 5 PCs:
   ```r
   scores_df <- as.data.frame(pca_wine$x[, 1:5]) |>
     as_tibble() |>
     mutate(cultivar = wine_clean$label)
   ```

b. Split the data into training (70%) and test (30%) sets using
   `set.seed(2026)` and `sample()`. Fit a multinomial logistic regression
   on the training set using `nnet::multinom()`:
   ```r
   library(nnet)
   fit <- multinom(cultivar ~ PC1 + PC2 + PC3 + PC4 + PC5,
                   data = train_scores, trace = FALSE)
   ```

c. Predict cultivar on the test set and build a confusion matrix:
   ```r
   pred <- predict(fit, newdata = test_scores)
   table(Predicted = pred, Actual = test_scores$cultivar)
   ```
   What is the overall classification accuracy?

d. Now fit a second model using only PC1 and PC2. Compare the classification
   accuracy of the 2-PC model to the 5-PC model. How much accuracy is lost
   by using only the first two components?

e. **Discussion.** In the morning notes we warned that PCA is unsupervised --
   it finds the directions of maximum variance in $X$ without looking at $Y$.
   In this wine example, the PCA worked well for classification anyway. Why
   do you think that is? Can you think of a situation where the first two PCs
   would explain most of the variance in $X$ but fail to separate the groups
   in $Y$?

```{r}
# your code here
```

---

# Quick Reference Card

## Setup

```r
library(tidyverse)
library(factoextra)
library(patchwork)
select <- dplyr::select
filter <- dplyr::filter
```

## Loading the Wine Data

```r
data(wine, package = "FactoMineR")

# Or from UCI directly (no package needed):
url <- paste0("https://archive.ics.uci.edu/ml/",
              "machine-learning-databases/wine/wine.data")
wine_raw <- read.csv(url, header = FALSE)
names(wine_raw) <- c("Cultivar", "Alcohol", "MalicAcid", "Ash",
  "AlcalinityAsh", "Magnesium", "TotalPhenols", "Flavanoids",
  "NonflavanoidPhenols", "Proanthocyanins", "ColorIntensity",
  "Hue", "OD280_OD315", "Proline")
wine_raw$Cultivar <- factor(wine_raw$Cultivar)
```

## Running PCA

```r
# Extract numeric variables and scale
X   <- wine_clean |> select(where(is.numeric)) |> as.matrix()
X_s <- scale(X)

# Fit PCA (on already-scaled data)
pca_res <- prcomp(X_s, center = FALSE, scale. = FALSE)

# Fit PCA with scaling inside prcomp
pca_res <- prcomp(X, center = TRUE, scale. = TRUE)
```

## Variance Explained

```r
# Proportion of variance per PC
var_exp <- pca_res$sdev^2 / sum(pca_res$sdev^2)

# Cumulative
cumsum(var_exp)

# Table
tibble(PC = seq_along(var_exp),
       Individual  = round(var_exp * 100, 1),
       Cumulative  = round(cumsum(var_exp) * 100, 1))

# Kaiser's rule: eigenvalue > 1
pca_res$sdev^2

# Scree plot
fviz_eig(pca_res, addlabels = TRUE,
         barfill = "steelblue", barcolor = "gray40")
```

## Score Plots

```r
# Color by group
fviz_pca_ind(pca_res,
             geom.ind    = "point",
             habillage   = group_vector,
             addEllipses = TRUE,
             repel       = TRUE)

# Use different PCs (e.g., PC1 vs PC3)
fviz_pca_ind(pca_res, axes = c(1, 3),
             habillage = group_vector)

# Extract scores manually
scores <- as.data.frame(pca_res$x)
```

## Loadings

```r
# Raw loadings matrix
pca_res$rotation[, 1:3] |> round(3)

# Largest contributors to PC1
sort(abs(pca_res$rotation[, 1]), decreasing = TRUE)

# Loading bar plot
as.data.frame(pca_res$rotation) |>
  rownames_to_column("variable") |>
  pivot_longer(-variable, names_to = "PC", values_to = "loading") |>
  filter(PC %in% c("PC1","PC2")) |>
  ggplot(aes(x = reorder(variable, loading), y = loading,
             fill = loading > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "steelblue",
                                "FALSE" = "firebrick")) +
  facet_wrap(~ PC, scales = "free_y") +
  coord_flip() +
  labs(x = NULL, y = "Loading weight") +
  theme_minimal()

# Biplot
fviz_pca_biplot(pca_res,
                geom.ind    = "point",
                habillage   = group_vector,
                addEllipses = TRUE,
                repel       = TRUE)
```

## Predicting New Data

```r
# Apply same rotation to new (scaled) data
X_new_s <- scale(X_new,
                 center = attr(X_s, "scaled:center"),
                 scale  = attr(X_s, "scaled:scale"))
new_scores <- predict(pca_res, newdata = X_new_s)
```

## Key Concepts

| Term | Meaning |
|---|---|
| Loadings | Weights each variable contributes to a PC. Large = influential. |
| Scores | Each observation's value on each PC. Used in regression/plots. |
| Eigenvalue | Variance explained by a PC. Kaiser's rule: keep eigenvalue > 1. |
| Scree plot | Eigenvalue vs. PC number. Look for the elbow. |
| Biplot | Shows both observations (points) and variables (arrows). |
| Global method | Transformation is linear and invertible. Can use in regression. |
| Local method | t-SNE, UMAP. For visualization only. Cannot use in regression. |
