---
title: "Day 14 (PM) Lab: Principal Component Analysis -- Solutions"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(this.path)
library(factoextra)
library(patchwork)
setwd(this.dir())

select <- dplyr::select
filter <- dplyr::filter
```

## Load the Wine Data

```{r}
url <- paste0("https://archive.ics.uci.edu/ml/",
              "machine-learning-databases/wine/wine.data")
wine_raw <- read.csv(url, header = FALSE)
names(wine_raw) <- c("Cultivar", "Alcohol", "MalicAcid", "Ash",
  "AlcalinityAsh", "Magnesium", "TotalPhenols", "Flavanoids",
  "NonflavanoidPhenols", "Proanthocyanins", "ColorIntensity",
  "Hue", "OD280_OD315", "Proline")
wine_raw$Cultivar <- factor(wine_raw$Cultivar)

wine <- wine_raw

head(wine)
dim(wine)
```

For the exercises we will use a clean version with clear variable names:

```{r}
# Build a clean tibble with the 13 chemical variables + cultivar label
# Using the UCI wine data structure (adjust column names if your load differs)
wine_clean <- wine |>
  as_tibble() |>
  rename_with(tolower)

# Check the structure
glimpse(wine_clean)
```

```{r}
# Summary statistics for the chemical variables
wine_clean |>
  select(where(is.numeric)) |>
 summary()
```

The variables are on very different scales:  **Scaling is essential** before PCA.

## Step 1: Scale and Run PCA

```{r}
# Extract the numeric variables (drop the cultivar label for PCA)
X <- wine_clean |>
  select(where(is.numeric)) |>
  as.matrix()

# Standardize
X_s <- scale(X)

# Run PCA
pca_wine <- prcomp(X_s, center = FALSE, scale. = FALSE)
```

## Step 2: Quick Scree Plot

```{r}
fviz_eig(pca_wine, addlabels = TRUE,
         barfill = "steelblue", barcolor = "gray40") +
  labs(title = "Scree Plot: Wine PCA") +
  theme_minimal()
```

The first two PCs together explain roughly 55% of the total variance, and
the curve shows a clear elbow after PC2 or PC3 -- much sharper than the NHANES
data from the morning, where variance was spread more evenly.

## Step 3: Score Plot

```{r}
#| fig-width: 6
#| fig-height: 5
fviz_pca_ind(pca_wine,
             geom.ind   = "point",
             habillage  = wine_clean$cultivar,
             addEllipses = TRUE,
             repel      = TRUE) +
  labs(title = "PCA Score Plot: Wine Cultivars") +
  theme_minimal()
```

---

**Exercise 1 -- Variance Explained.**

a. Compute the proportion of variance explained by each PC. The formula is:
   ```r
   var_exp <- pca_wine$sdev^2 / sum(pca_wine$sdev^2)
   ```
   Print a table with the PC number, individual variance explained (%), and
   cumulative variance explained (%).
   

```{r}
# Proportion of variance explained by each PC
var_explained <- pca_wine$sdev^2 / sum(pca_wine$sdev^2)
cum_var       <- cumsum(var_explained)

var_df <- tibble(
  PC         = seq_along(var_explained),
  Individual = var_explained * 100,
  Cumulative = cum_var * 100
)

var_df |>
  mutate(across(c(Individual, Cumulative), ~ round(., 1))) |>
  print(n = 10)
```  

b. How many PCs are needed to explain at least 70% of the total variance?
   At least 90%?
   
   If we want to explain 70% of the variance we need to use 4 PCs, if we want to explain 90% we need to use 8 PCs.

c. Apply Kaiser's rule: keep PCs with eigenvalue greater than 1. The
   eigenvalues are `pca_wine$sdev^2`. How many PCs would Kaiser's rule
   suggest keeping?

```{r}
pca_wine$sdev^2
```

Kaiser's rule suggests using 3 PCs. 

d. Make a scree plot using `fviz_eig()`. Where does the "elbow" appear?
   Does it agree with the answers from parts b and c? Which criterion
   would you use, and why?


```{r}
#| fig-width: 8
#| fig-height: 4
# Proportion of variance explained by each PC
var_explained <- pca_wine$sdev^2 / sum(pca_wine$sdev^2)
cum_var       <- cumsum(var_explained)

var_df <- tibble(
  PC         = seq_along(var_explained),
  Individual = var_explained * 100,
  Cumulative = cum_var * 100
)

p1 <- fviz_eig(pca_wine, addlabels = TRUE,
               barcolor = "gray40", barfill = "steelblue") +
  labs(title = "Variance Explained per PC",
       x = "Principal Component", y = "% Variance") +
  theme_minimal()

p2 <- var_df |>
  ggplot(aes(x = PC, y = Cumulative)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_hline(yintercept = 80, linetype = "dashed",
             color = "firebrick", linewidth = 0.8) +
  annotate("text", x = 8, y = 91.5,
           label = "80% threshold", color = "firebrick", size = 3.5) +
  labs(title = "Cumulative Variance Explained",
       x = "Number of PCs", y = "Cumulative % Variance") +
  theme_minimal()

p1 + p2
```

From the Elbow method, I would say around 2 or 3 PCs seems to be a good choice.

---

**Exercise 2 -- Score Plot and Cultivar Separation.**

a. Produce a score plot (observations in PC space) using `fviz_pca_ind()`.
   Color points by cultivar (`habillage = wine_clean$label`) and add
   confidence ellipses (`addEllipses = TRUE`).


```{r}
#| fig-width: 7
#| fig-height: 6
#| message: false
fviz_pca_ind(
  pca_wine,
  geom.ind   = "point",
  habillage  = wine_clean$cultivar,
  addEllipses = TRUE,
  repel      = TRUE,
  col.var    = "black",
  alpha.ind  = 0.3,
  pointsize  = 1
) +
  labs(title  = "PCA Biplot: Wine PCA",
       color  = "Cultivar",
       fill   = "Cultivar") +
  theme_minimal()
```

b. Do the three cultivars separate in the first two PCs? Which cultivar
   is most distinct from the other two?
   
   Yes, the PCAs clearly separate the wines by cultivar. This confirms that the chemical variables genuinely capture cultivar differences – exactly what you want to see. Region 3, appears to be the most different.

c. PC1 runs left to right on the plot. Look at where each cultivar falls
   along PC1. Which cultivar has the highest PC1 scores (furthest right)?
   Which has the lowest (furthest left)?
   
   Wines from region 3 have large positive loadings on PC1, while wines from region 1 have larger negative loadings (on average). Wines from region 2 have PC1 loading that center around 0 with large spread.

d. Now make a second score plot using PC1 and PC3 instead of PC1 and PC2.
   Pass `axes = c(1, 3)` to `fviz_pca_ind()`. Does adding PC3 improve or
   worsen the separation between cultivars?

```{r}
#| fig-width: 7
#| fig-height: 6
#| message: false
fviz_pca_ind(
  pca_wine,
  geom.ind   = "point",
  habillage  = wine_clean$cultivar,
  addEllipses = TRUE,
  repel      = TRUE,
  col.var    = "black",
  alpha.ind  = 0.3,
  axes = c(1,3),
  pointsize  = 1
) +
  labs(title  = "PCA Biplot: Wine PCA",
       color  = "Cultivar",
       fill   = "Cultivar") +
  theme_minimal()
```

PC3 has basically no relationship with region. The regional differences were much more clear when visualizing as a function of PC1 and PC2.

---

**Exercise 3 -- Loadings.**

a. Extract the loadings matrix from `pca_wine$rotation`. Print the loadings
   for the first three PCs, rounded to 3 decimal places.

```{r}
# Numerical loadings for the first three PCs
pca_wine$rotation[, 1:3] |> round(3)
```

b. For PC1: which three variables have the largest absolute loadings? What
   do these variables have in common chemically? Based on this, what
   biological property might PC1 be capturing?
   
   Flavanoids, totalphenols, and od280_od315 have the largest loadings. These variables are all tied to phenolic content in wine. In plain terms, this axis is mostly capturing how phenolic-rich the wine is. Phenolics are linked to things like bitterness, astringency, antioxidant capacity, and overall “structure” of the wine.

c. For PC2: which three variables have the largest absolute loadings? Are
   these the same variables as PC1 or different? What might PC2 be
   capturing?
   
   For PC2, colorintensity, alcohol, and proline all have large loadings. These are all often related to grape ripeness, fermentation/extraction, and wine concentration/body. So PC2 is probably capturing something like grape maturation and fermentation outcomes rather than a single chemical class.     

d. Make a loading bar plot for PC1 and PC2 (see the Quick Reference Card
   for the ggplot2 code). Does the pattern of loadings match your
   interpretation from parts b and c?

```{r}
#| fig-width: 9
#| fig-height: 5
# Plot loadings for PC1 and PC2
loadings_df <- as.data.frame(pca_wine$rotation) |>
  rownames_to_column("variable") |>
  as_tibble() |>
  pivot_longer(cols = -variable, names_to = "PC", values_to = "loading") |>
  filter(PC %in% c("PC1", "PC2"))

loadings_df |>
  ggplot(aes(x = reorder(variable, loading), y = loading,
             fill = loading > 0)) +
  geom_col(show.legend = FALSE, color = "white") +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "firebrick")) +
  facet_wrap(~ PC, scales = "free_y") +
  coord_flip() +
  labs(x = NULL, y = "Loading weight",
       title = "PCA Loadings: PC1 and PC2",
       subtitle = "Blue = positive contribution; red = negative contribution") +
  theme_minimal()
```

e. Produce a **biplot** using `fviz_pca_biplot()`. Color observations by
   cultivar. Describe: which variables have arrows pointing in the same
   direction as the cultivar that has the highest PC1 scores?


```{r}
#| fig-width: 7
#| fig-height: 6
#| message: false
fviz_pca_biplot(
  pca_wine,
  geom.ind   = "point",
  habillage  = wine_clean$cultivar,
  addEllipses = TRUE,
  repel      = TRUE,
  col.var    = "black",
  alpha.ind  = 0.3,
  pointsize  = 1
) +
  labs(title  = "PCA Biplot: Wine dataset",
       color  = "Cultivar",
       fill   = "Cultivar") +
  theme_minimal()
```


