---
title: "Day 14: Dimension Reduction and Principal Component Regression"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(this.path)
library(factoextra)
library(pls)
library(patchwork)
setwd(this.dir())

# Resolve namespace conflicts
select  <- dplyr::select
filter  <- dplyr::filter
```

# Overview

This week we have been working with genomics data where $p \gg n$: far more
variables than observations. Today we step back and look at the general problem
of dimension reduction -- what to do when your predictors are high-dimensional,
correlated, or both -- and how to use the results in a regression model.

We use the **NHANES** dataset: a large cross-sectional survey of U.S. adults
with measurements on BMI, blood pressure, cholesterol, pulse, sleep, income,
and alcohol use. These variables are correlated in biologically meaningful ways,
which makes them a good teaching example for both PCA and PCR.

Three examples:

- **Example 1:** PCA on NHANES health indicators -- fitting, scree plot,
  loadings, and biplot.
- **Example 2:** Using PCA scores for visualization -- what the components
  represent and how samples distribute across them.
- **Example 3:** Principal Component Regression -- using PC scores to predict
  systolic blood pressure while eliminating collinearity.

---

# The Data

```{r}
library(NHANES)
library(janitor)

# Adult participants only, complete cases on our variables of interest
vars_keep <- c("Age", "BMI", "bp_sys_ave", "bp_dia_ave", "tot_chol",
               "direct_chol", "pulse", "sleep_hrs_night",
               "hh_income_mid", "alcohol_day")

nhanes <- NHANES |>
  clean_names() |>
  filter(age >= 20) |>
  select(id, gender, race1, all_of(tolower(vars_keep))) |>
  drop_na()

cat("Observations:", nrow(nhanes), "\n")
cat("Variables:   ", length(vars_keep), "\n")
```

```{r}
# Quick look at the variables
nhanes |>
  select(-id, -gender, -race1) |>
  summary()
```

Several variables are right-skewed (household income, alcohol). We
log-transform these before standardizing.

```{r}
to_log <- c("tot_chol", "direct_chol", "sleep_hrs_night",
            "hh_income_mid", "alcohol_day")

nhanes_tr <- nhanes |>
  mutate(across(all_of(to_log), ~ log(. + 1)))

# Numeric matrix for PCA (no IDs or categorical variables)
X <- nhanes_tr |>
  select(age, bmi, bp_sys_ave, bp_dia_ave, tot_chol,
         direct_chol, pulse, sleep_hrs_night,
         hh_income_mid, alcohol_day) |>
  as.matrix()

# Standardize: PCA is sensitive to scale
X_s <- scale(X)
```

::: {.callout-important}
**Always scale before PCA** when variables are measured in different units or
on different scales. Without scaling, variables with large variance (e.g.,
household income) will dominate the first component and variables with small
variance (e.g., sleep hours) will barely contribute.
:::

---

# Example 1: Fitting PCA

## Running PCA

```{r}
pca_res <- prcomp(X_s, center = FALSE, scale. = FALSE)
# center = FALSE and scale. = FALSE because X_s is already centered and scaled
```

## Scree Plot and Variance Explained

```{r}
#| fig-width: 8
#| fig-height: 4
# Proportion of variance explained by each PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2)
cum_var       <- cumsum(var_explained)

var_df <- tibble(
  PC         = seq_along(var_explained),
  Individual = var_explained * 100,
  Cumulative = cum_var * 100
)

p1 <- fviz_eig(pca_res, addlabels = TRUE,
               barcolor = "gray40", barfill = "steelblue") +
  labs(title = "Variance Explained per PC",
       x = "Principal Component", y = "% Variance") +
  theme_minimal()

p2 <- var_df |>
  ggplot(aes(x = PC, y = Cumulative)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_hline(yintercept = 90, linetype = "dashed",
             color = "firebrick", linewidth = 0.8) +
  annotate("text", x = 8, y = 91.5,
           label = "90% threshold", color = "firebrick", size = 3.5) +
  labs(title = "Cumulative Variance Explained",
       x = "Number of PCs", y = "Cumulative % Variance") +
  theme_minimal()

p1 + p2
```

```{r}
var_df |>
  mutate(across(c(Individual, Cumulative), ~ round(., 1))) |>
  print(n = 10)
```

**Reading the scree plot:**

- PC1 explains about 18% of the total variance, PC2 about 17%, and so on.
- The curve has no sharp elbow -- variance is spread fairly evenly across
  components, which is typical when variables are only moderately correlated.
- To explain at least 80% of the variance, we need about 7 PCs. To explain
  90%, we need 8.

For a dataset where all variables are biologically related health indicators,
this diffuse scree plot makes sense: there is no single dominant dimension.

## Loadings

The loadings tell us what each PC represents in terms of the original variables.

```{r}
#| fig-width: 9
#| fig-height: 5
# Plot loadings for PC1 and PC2
loadings_df <- as.data.frame(pca_res$rotation) |>
  rownames_to_column("variable") |>
  as_tibble() |>
  pivot_longer(cols = -variable, names_to = "PC", values_to = "loading") |>
  filter(PC %in% c("PC1", "PC2"))

loadings_df |>
  ggplot(aes(x = reorder(variable, loading), y = loading,
             fill = loading > 0)) +
  geom_col(show.legend = FALSE, color = "white") +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "firebrick")) +
  facet_wrap(~ PC, scales = "free_y") +
  coord_flip() +
  labs(x = NULL, y = "Loading weight",
       title = "PCA Loadings: PC1 and PC2",
       subtitle = "Blue = positive contribution; red = negative contribution") +
  theme_minimal()
```

```{r}
# Numerical loadings for the first three PCs
pca_res$rotation[, 1:3] |> round(3)
```

**Interpreting PC1:**

Age, systolic BP, total cholesterol, and diastolic BP all have large negative
loadings on PC1. Alcohol use has a large positive loading. A high PC1 score
(more positive) corresponds to younger participants who drink more; a low PC1
score corresponds to older participants with higher blood pressure and
cholesterol. PC1 could be described as a "cardiovascular aging" axis, running
from younger/higher-alcohol at one end to older/higher-BP at the other.

**Interpreting PC2:**

BMI, systolic BP, and diastolic BP have the largest (positive) loadings on PC2. HDL
cholesterol (`direct_chol`) has a large negative loading (higher HDL is
associated with lower BMI, a known relationship). PC2 looks like a "metabolic
syndrome" axis.

::: {.callout-tip}
Interpreting PCs is part art, part science. The signs of the loadings are
arbitrary (flip all signs and you get an equivalent component), so only the
pattern of which variables load together -- and their relative magnitudes --
carries information. A loading of -0.54 on age and -0.40 on systolic BP tells
you those two variables co-vary in the same direction within this component,
not that one is "better" than the other.
:::

---

# Example 2: Visualizing PCA Scores

## Biplot

The biplot shows both the observations (points) and the variables (arrows) in
the same plot.

```{r}
#| fig-width: 7
#| fig-height: 6
#| message: false
fviz_pca_biplot(
  pca_res,
  geom.ind   = "point",
  habillage  = nhanes_tr$gender,
  addEllipses = TRUE,
  repel      = TRUE,
  col.var    = "black",
  alpha.ind  = 0.3,
  pointsize  = 1
) +
  labs(title  = "PCA Biplot: NHANES Health Indicators",
       color  = "Gender",
       fill   = "Gender") +
  theme_minimal()
```

**Reading the biplot:**

- Each point is one participant. Points close together have similar health
  profiles across the ten variables.
- Each arrow is one variable, pointing in the direction of its maximum
  contribution to the two displayed PCs.
- Arrows pointing in the same direction are positively correlated in the data
  (age and systolic BP; BMI and diastolic BP). Arrows pointing in opposite
  directions are negatively correlated (age and alcohol use).
- The two gender groups show modest separation: male participants (on average
  higher alcohol use, different BP profiles) tend to cluster somewhat
  differently from female participants.

## Score Plot Colored by Age Group

```{r}
#| fig-width: 7
#| fig-height: 5
scores <- as.data.frame(pca_res$x) |>
  as_tibble() |>
  bind_cols(nhanes_tr |> select(gender, age))

scores |>
  mutate(age_group = cut(age, breaks = c(20, 40, 60, 80),
                         labels = c("20-40", "40-60", "60-80"),
                         include.lowest = TRUE)) |>
  ggplot(aes(x = PC1, y = PC2, color = age_group)) +
  geom_point(alpha = 0.3, size = 0.8) +
  stat_ellipse(linewidth = 1) +
  scale_color_manual(values = c("20-40"  = "steelblue",
                                 "40-60"  = "darkgreen",
                                 "60-80"  = "firebrick")) +
  labs(title = "PCA Scores Colored by Age Group",
       x = "PC1 (cardiovascular aging)",
       y = "PC2 (metabolic syndrome)",
       color = "Age group") +
  theme_minimal()
```

Older participants cluster on the left side of PC1 (consistent with our
loading interpretation: lower PC1 score = older age with higher BP).
PC2 does not show a strong age gradient, as expected since it primarily
captures BMI and metabolic risk rather than age per se.

---

# Example 3: Principal Component Regression

## The Collinearity Problem

We want to predict **systolic blood pressure** from the other health
indicators. Let us first check the collinearity among predictors.

```{r}
# Exclude the outcome from the predictor matrix
X_pred <- nhanes_tr |>
  select(age, bmi, bp_dia_ave, tot_chol, direct_chol,
         pulse, sleep_hrs_night, hh_income_mid, alcohol_day)

# Correlation matrix
cor_mat <- cor(X_pred)
round(cor_mat, 2)
```

```{r}
#| fig-width: 6
#| fig-height: 5
# Visualize correlations
corrplot::corrplot(cor_mat, method = "color", type = "upper",
                  tl.cex = 0.8, addCoef.col = "black",
                  number.cex = 0.65,
                  col = colorRampPalette(c("firebrick","white","steelblue"))(100))
```

Diastolic BP and age are both correlated with the outcome and with each other.
Several other moderate correlations exist. Fitting a plain OLS regression with
all these predictors will give unstable coefficient estimates.

## Standard OLS: The Instability

```{r}
nhanes_model <- nhanes_tr |>
  select(bp_sys_ave, age, bmi, bp_dia_ave, tot_chol, direct_chol,
         pulse, sleep_hrs_night, hh_income_mid, alcohol_day)

ols_fit <- lm(bp_sys_ave ~ ., data = nhanes_model)
summary(ols_fit)
```

The OLS model fits, but notice the large standard errors on some coefficients. Overall, this doesn't appear to have to many issues though. 

## PCR: The Fix

PCR replaces the correlated predictors with uncorrelated PC scores, then
regresses the outcome on those scores.

```{r}
# Step 1: PCA on the predictors (not including the outcome)
X_pred_s <- scale(as.matrix(X_pred))
pca_pred  <- prcomp(X_pred_s, center = FALSE, scale. = FALSE)

# How much variance do the first r PCs explain?
var_pred <- pca_pred$sdev^2 / sum(pca_pred$sdev^2)
cumsum(var_pred) |> round(3)
```

```{r}
# Keep 5 PCs (explains ~75% of predictor variance -- a reasonable choice here)
# In practice, choose r by cross-validation
r      <- 5
scores <- as.data.frame(pca_pred$x[, 1:r])
names(scores) <- paste0("PC", 1:r)

# Step 2: Regress outcome on PC scores
Y <- nhanes_tr$bp_sys_ave
pcr_fit <- lm(Y ~ ., data = scores)
summary(pcr_fit)
```

The PC scores are uncorrelated by construction, so the standard errors are
well-behaved and the coefficients are interpretable. Each significant PC
represents a dimension of the predictor space that is independently associated
with systolic BP.

## Choosing r by Cross-Validation

Choosing $r = 5$ above was somewhat arbitrary. The `pls` package provides
`pcr()` with built-in cross-validation.

```{r}
#| fig-width: 7
#| fig-height: 4
# pcr() from the pls package handles the whole workflow
pcr_cv <- pcr(bp_sys_ave ~ age + bmi + bp_dia_ave + tot_chol +
                            direct_chol + pulse + sleep_hrs_night +
                            hh_income_mid + alcohol_day,
              data      = as.data.frame(nhanes_tr),
              scale     = TRUE,
              validation = "CV")

# Plot cross-validated RMSE vs number of components
validationplot(pcr_cv, val.type = "RMSEP",
               main = "PCR: Cross-Validated RMSE vs Number of Components")
```

```{r}
# Find the r with the lowest CV RMSEP
cv_rmse <- RMSEP(pcr_cv)$val[1, 1, ]  # intercept-only to full model
best_r   <- which.min(cv_rmse) - 1    # subtract 1 for the intercept row
cat("Optimal number of components:", best_r, "\n")
cat("CV RMSEP at optimal r:        ", round(min(cv_rmse), 3), "\n")
```

The optimal number of components according to cross-validation is
typically 6--8 for this dataset, depending on the random seed. The key
observation is that adding more components beyond this point does not
meaningfully improve prediction -- confirming that we can reduce 9
predictors to a handful of PCs without losing predictive ability.

## Comparing OLS and PCR Predictions

```{r}
#| fig-width: 8
#| fig-height: 4
# Predictions on training data (for illustration; honest comparison uses test data)
pred_ols <- fitted(ols_fit)
pred_pcr <- as.vector(predict(pcr_cv, ncomp = best_r))

tibble(
  Method    = c("OLS (all predictors)", paste0("PCR (", best_r, " components)")),
  R_squared = c(summary(ols_fit)$r.squared,
                cor(pred_pcr, Y)^2)
) |>
  mutate(R_squared = round(R_squared, 4))
```

The R-squared values are similar, confirming that PCR achieves comparable
predictive performance with fewer effective parameters. The key advantages of
PCR are in settings where OLS fails entirely: when $p > n$ (more predictors
than observations) or when collinearity is severe enough to make OLS
coefficients meaningless.

## Back-Transforming to Original Coefficients

One limitation of PCR is that the regression coefficients are on the PC scale,
not the original variable scale. We can back-transform them.

```{r}
# Extract coefficients on original variable scale
beta_pc <- coef(pcr_fit)[-1]  # drop intercept
rotation <- pca_pred$rotation[, 1:r]
scale_sd  <- attr(X_pred_s, "scaled:scale")
scale_mn  <- attr(X_pred_s, "scaled:center")

# Original-scale coefficients
beta_orig <- (rotation %*% beta_pc) / scale_sd
round(beta_orig, 4)
```

These are the implied regression coefficients on the original (log-transformed,
but not standardized) variables. The signs and magnitudes should be more
interpretable than the raw OLS output when collinearity is high.

---

# Connecting Back to Genomics

The PCA you ran on Days 11--13 was doing the same thing as `prcomp()` above,
just on a much larger matrix (genes instead of health variables).

```{r}
# The plotPCA() call in DESeq2 is equivalent to:
# pca_res <- prcomp(t(vst_matrix), scale. = FALSE)
# scores  <- pca_res$x[, 1:2]
# The samples are the "observations" and the genes are the "variables"

# In the Framingham data you could do:
# pca_chd <- prcomp(framingham[, continuous_vars], scale. = TRUE)
# Then use pca_chd$x[, 1:3] as predictors in a logistic regression

cat("The same three lines of code work whether you have 10 variables or 20,000.\n")
```

The only difference in genomics is that the matrix is transposed (samples in
rows rather than genes), and the VST transformation stabilizes the count
variance before passing it to `prcomp()`. The rest of the workflow is
identical.

---

# Summary

**Dimension reduction:**

- When predictors are high-dimensional or correlated, create a smaller set
  of variables that summarize the originals.
- Global methods (PCA) preserve overall variance structure and can be used
  in regression. Local methods (t-SNE, UMAP) preserve neighborhood structure
  and are for visualization only.

**PCA:**

- Creates uncorrelated components ordered by variance explained.
- Always scale your variables first. Use a scree plot or cumulative variance
  to choose the number of components.
- Loadings tell you what each component represents. Scores are used
  downstream in regression, clustering, or visualization.

**PCR:**

- Replace correlated predictors with uncorrelated PCA scores, then regress
  $Y$ on the scores.
- Eliminates collinearity completely. Works even when $p > n$.
- Choose the number of PCs by cross-validation, not by how much variance
  they explain in $X$.
- Core limitation: PCA is unsupervised. Components explaining the most
  variance in $X$ are not guaranteed to be the most predictive of $Y$.
