---
title: "Day 11: Genomics Data -- Structure, Formats, and Quality Assessment"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(this.path)
library(SummarizedExperiment)

# uncomment the lines below to install packages required for 'airway' and its analysis:
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("airway")
# BiocManager::install("DESeq2")


library(airway)
setwd(this.dir())
```

# Overview

This week we shift from modeling a handful of variables on a few thousand
participants to analyzing **tens of thousands of variables simultaneously**.
The biological data are different, but the statistical questions are familiar:
is this variable associated with the outcome? How do we adjust for confounders?
How do we avoid false positives when testing thousands of hypotheses at once?

Today's session has one goal: understand the **structure** of genomics data
well enough to work with it in R. Everything else this week builds on this
foundation.

We will work with the `airway` dataset throughout: an RNA-sequencing experiment
measuring gene expression in human airway smooth muscle cells treated with
dexamethasone (a glucocorticoid), compared with untreated controls. This
dataset is small enough to explore interactively and clean enough to illustrate
the core ideas without distraction.

Three examples:

- **Example 1:** The count matrix -- what it is, what it looks like, and why
  standard regression does not apply directly.
- **Example 2:** The `SummarizedExperiment` object -- how genomics data is
  organized in R, and how to access its components.
- **Example 3:** Quality assessment -- library sizes, zero-count genes, count
  distributions, and PCA.

---

# The Statistical Setting

## Why This Week Is Different

In everything we have done so far, the number of observations $n$ was much
larger than the number of variables $p$. In the Framingham data, 4,240
participants and fewer than 20 predictors. Standard regression works because
there is far more data than parameters to estimate.

RNA-sequencing data flips this relationship:

```{r}
data(airway)
dim(airway)
```

The `airway` dataset has `r nrow(airway)` genes (rows) and only
`r ncol(airway)` samples (columns). In a typical study you might have
$p \approx 20{,}000$ genes and $n \approx 6$ to $50$ samples. We are firmly in
the regime $p \gg n$.

This creates real statistical problems:

- You cannot fit a single regression with 20,000 predictors when you have 8
  observations. The system is massively underdetermined.
- We instead test each gene **separately** (one gene at a time), which creates
  a different problem: testing 20,000 hypotheses simultaneously means many false
  positives by chance alone. Multiple testing correction (Day 13) is essential.
- The data are **counts**, not continuous measurements. Standard linear models
  and t-tests applied directly to raw counts give incorrect results. We need
  count-appropriate models (Day 12).

## The Upstream Pipeline

We will not run the bioinformatics pipeline that converts raw sequencing output
into analysis-ready data. That pipeline requires specialized software (STAR,
HISAT2, featureCounts, Salmon) and computing infrastructure. What matters for us
is knowing what each stage produces and where our work begins.

| Stage | Output | What it contains |
|---|---|---|
| Sequencing | FASTQ | Raw DNA sequences + quality scores |
| Alignment | BAM | Reads mapped to positions on the reference genome |
| Counting | Count matrix | Reads per gene per sample (integers) |
| **Our starting point** | **Count matrix** | **What we analyze** |

The `airway` dataset is already at the "count matrix" stage. Someone else ran
the pipeline. Our job starts here.

---

# Example 1: The Count Matrix

## Loading and Inspecting the Data

```{r}
data(airway)
airway
```

The printed summary tells us the object type (`RangedSummarizedExperiment`),
dimensions, and what components are stored. We will unpack each component in
Example 2. For now, extract the count matrix:

```{r}
counts <- assay(airway)
dim(counts)
counts[1:6, 1:4]
```

Each row is a gene (identified by an Ensembl gene ID such as
`ENSG00000000003`). Each column is a sample (identified by an SRA run
accession such as `SRR1039508`). Each value is the number of sequencing reads
that mapped to that gene in that sample.

## Statistical Properties of Count Data

Counts are non-negative integers. That immediately rules out several tools:

**Why not use a standard t-test or linear regression on raw counts?**

```{r}
#| fig-width: 8
#| fig-height: 4

# Distribution of counts for the first sample
counts_s1 <- counts[, 1]

p1 <- tibble(count = counts_s1) |>
  filter(count > 0) |>
  ggplot(aes(x = count)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Raw counts (one sample, non-zero genes only)",
       x = "Count", y = "Number of genes") +
  theme_minimal()

p2 <- tibble(count = counts_s1) |>
  filter(count > 0) |>
  mutate(log2count = log2(count + 1)) |>
  ggplot(aes(x = log2count)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  labs(title = "log2(count + 1) -- same sample",
       x = "log2(count + 1)", y = "Number of genes") +
  theme_minimal()

library(patchwork)
p1 + p2
```

The raw count distribution is extremely right-skewed: a handful of highly
expressed genes have counts in the hundreds of thousands, while most genes have
counts in the single digits or zero. The log-transformed distribution is much
more symmetric. This shape is characteristic of RNA-seq data and rules out
Gaussian (normal) models for the raw counts.

Three additional properties make standard methods inappropriate:

1. **Counts are integers.** Residuals from a linear model fit to count data
   cannot be normally distributed.
2. **Variance is not constant.** Genes with higher mean expression have higher
   variance. Homoscedasticity fails.
3. **Many zeros.** Genes with very low expression may have zero counts in some
   or all samples. Zeros are real observations, not missing data.

The appropriate model is the **negative binomial distribution**, which
generalizes the Poisson to allow variance to exceed the mean. DESeq2
(Day 12) fits this model.

## Zero-Count Genes

```{r}
# How many genes have zero counts across all samples?
n_all_zero <- sum(rowSums(counts) == 0)
n_low      <- sum(rowSums(counts) < 10)

cat("Genes with zero total counts:", n_all_zero, "\n")
cat("Genes with fewer than 10 total counts:", n_low, "\n")
cat("Total genes:", nrow(counts), "\n")
```

A meaningful fraction of genes have very low or zero counts across all samples.
These genes carry no information about differential expression and add noise to
the multiple testing correction. Standard practice is to filter them out before
modeling. We will do this in the PM lab and again on Day 12.

## Sequencing Depth

Each sample is sequenced to a certain **library size** (total number of reads).
If one sample has twice as many reads as another, its counts will tend to be
twice as large even if the underlying biology is identical. Raw counts are not
directly comparable across samples.

```{r}
#| fig-width: 7
#| fig-height: 4
library_sizes <- colSums(counts)

tibble(
  sample    = names(library_sizes),
  lib_size  = library_sizes,
  treatment = colData(airway)$dex
) |>
  ggplot(aes(x = sample, y = lib_size / 1e6, fill = treatment)) +
  geom_col(color = "white") +
  scale_fill_manual(values = c("untrt" = "steelblue", "trt" = "firebrick")) +
  labs(x = "Sample", y = "Total reads (millions)",
       fill = "Treatment",
       title = "Library Sizes by Sample",
       subtitle = "Samples must be normalized before comparison") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Library sizes vary across samples even within the same experiment. Before
comparing gene expression across samples, we must adjust for these differences.
DESeq2 handles this through **size factor estimation** -- a single scaling
number per sample that accounts for its total sequencing depth.

---

# Example 2: The SummarizedExperiment Object

## Three Components, Always Linked

A genomics dataset is not just a count matrix. It has three linked components:

1. **The assay (count matrix):** genes x samples, integer counts.
2. **Sample metadata (colData):** one row per sample -- treatment group, cell
   line, batch, patient ID, etc.
3. **Gene metadata (rowData):** one row per gene -- gene name, chromosome,
   genomic coordinates, gene type.

These three components must stay **synchronized**. If you reorder samples, the
sample metadata must reorder too. If you filter genes, the gene metadata must
filter too. Bioconductor's `SummarizedExperiment` enforces this by design.

```{r}
# The three accessor functions
assay(airway)[1:4, 1:4]       # count matrix
colData(airway)               # sample metadata
rowData(airway)[1:4, ]        # gene metadata (first 4 rows)
```

## The Sample Metadata: Your Study Design

```{r}
colData(airway) |> as_tibble()
```

The `colData` is your **study design matrix**. Before fitting any model, you
need to understand every column:

- `SampleName`: the sample identifier.
- `cell`: the cell line (four different cell lines in this experiment).
- `dex`: treatment group -- `trt` (dexamethasone-treated) or `untrt`
  (untreated control). **This is the primary variable of interest.**
- `albut`: whether the sample was also treated with albuterol (not the focus
  here).
- `Run`, `avgLength`, `Experiment`, `Sample`, `BioProject`: SRA metadata.

::: {.callout-important}
**Always inspect `colData` before modeling.** Know which column encodes
treatment vs. control. Verify that the sample order in the count matrix matches
the sample order in the metadata. A mismatch here means every downstream
result is wrong.
:::

## Verifying the Alignment

One of the most important sanity checks in genomics analysis: confirm that the
column names of the count matrix match the row names of `colData`.

```{r}
# Column names of the count matrix
colnames(assay(airway))

# Row names of the sample metadata
rownames(colData(airway))

# Are they identical and in the same order?
identical(colnames(assay(airway)), rownames(colData(airway)))
```

If this returns `FALSE`, something is wrong with the data object and no analysis
should proceed until it is resolved.

## Subsetting a SummarizedExperiment

One of the key benefits of `SummarizedExperiment` is that subsetting keeps all
three components synchronized automatically.

```{r}
# Select only the treated samples
airway_trt <- airway[, airway$dex == "trt"]
dim(airway_trt)
colData(airway_trt)$dex
```

Compare this with what would happen if the count matrix and metadata were stored
as separate objects: you would have to manually subset both in the same order,
and a mistake would silently corrupt the analysis.

## The Gene Metadata

```{r}
airway_rowdata <- rowData(airway) |>
  as_tibble() 
airway_rowdata |>
  select(gene_id, gene_name, gene_biotype) |>
  head(10)
```

```{r}
# What gene types are present?
table_biotype = table(airway_rowdata$gene_biotype) 
head(table_biotype[order(table_biotype, decreasing = TRUE)], 10)
```

Most genes in the annotation are protein-coding, but there are also long
non-coding RNAs, pseudogenes, and other categories. In a standard differential
expression analysis, we often filter to protein-coding genes, though this is
not always necessary.

---

# Example 3: Quality Assessment

Quality assessment in genomics plays the same role as residual diagnostic plots
in regression: it is a required first step before any modeling. Problems
discovered at this stage -- outlier samples, batch effects, misannotated
metadata -- are far easier to fix now than after a full analysis.

## Library Sizes Revisited

We already plotted library sizes above. The summary statistics:

```{r}
tibble(
  sample   = colnames(airway),
  lib_size = colSums(assay(airway)),
  treatment = colData(airway)$dex,
  cell_line = colData(airway)$cell
) |>
  arrange(lib_size)
```

No sample has an extreme outlier library size (e.g., ten times smaller than
the others), which would be a red flag suggesting a failed sequencing run.
The variation we see is normal and will be handled by normalization.

## Count Distributions Across Samples

After confirming that library sizes are reasonable, check that the count
distributions look similar in shape across samples.

```{r}
#| fig-width: 9
#| fig-height: 4
#| message: false
assay(airway) |>
  as_tibble(rownames = "gene") |>
  pivot_longer(-gene, names_to = "sample", values_to = "count") |>
  filter(count > 0) |>
  mutate(log2count = log2(count + 1)) |>
  left_join(
    colData(airway) |> as_tibble(rownames = "sample") |>
      select(sample, dex),
    by = "sample"
  ) |>
  ggplot(aes(x = log2count, color = dex, group = sample)) +
  geom_density(linewidth = 0.8, alpha = 0.8) +
  scale_color_manual(values = c("untrt" = "steelblue", "trt" = "firebrick")) +
  labs(x = "log2(count + 1)", y = "Density",
       color = "Treatment",
       title = "Count Distributions Across Samples",
       subtitle = "Curves should overlap closely -- large departures suggest a problem") +
  theme_minimal()
```

The density curves should overlap closely. Curves that are shifted or shaped
differently from the others may indicate a problematic sample.

## Filtering Low-Count Genes

```{r}
# Keep only genes with at least 10 total counts across all samples
keep <- rowSums(assay(airway)) >= 10
airway_filtered <- airway[keep, ]

cat("Genes before filtering:", nrow(airway), "\n")
cat("Genes after filtering: ", nrow(airway_filtered), "\n")
cat("Genes removed:         ", nrow(airway) - nrow(airway_filtered), "\n")
```

Note that subsetting `airway[keep, ]` keeps the count matrix, sample metadata,
and gene metadata all synchronized automatically.

## Sample-to-Sample Distances

A useful summary of the data is the pairwise distance between samples: how
similar is each sample to every other? If the experiment is working, samples
from the same treatment group should be closer to each other than to samples
from the other group.

```{r}
#| fig-width: 6
#| fig-height: 5
library(DESeq2)

# Variance-stabilizing transformation: puts counts on a scale
# appropriate for distance calculations and PCA
dds <- DESeqDataSet(airway_filtered, design = ~ dex)
vsd <- vst(dds, blind = TRUE)

# Sample-to-sample distances on the transformed data
sample_dists <- dist(t(assay(vsd)))
dist_matrix  <- as.matrix(sample_dists)

# Label rows and columns with treatment group and cell line
colnames(dist_matrix) <- rownames(dist_matrix) <-
  paste(colData(airway_filtered)$dex,
        colData(airway_filtered)$cell, sep = "_")

# Heatmap using base R (or you can use pheatmap)
heatmap(dist_matrix,
        symm     = TRUE,
        col      = colorRampPalette(c("steelblue", "white"))(50),
        main     = "Sample-to-Sample Distances",
        margins  = c(10, 10))
```

Samples in the same treatment group should show small distances (darker shading
if using blue-to-white), forming visible blocks along the diagonal. Unexpected
clustering -- for example, one treated sample clustering with all the untreated
samples -- would be a sign of a mislabeled or failed sample.

## PCA Plot

Principal Component Analysis (PCA) reduces the 20,000-dimensional count matrix
to two dimensions that capture the most variation. It is the fastest way to
see whether samples cluster as expected.

```{r}
#| fig-width: 6
#| fig-height: 5
pca_data <- plotPCA(vsd, intgroup = c("dex", "cell"), returnData = TRUE)
pct_var  <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2, color = dex, shape = cell)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("untrt" = "steelblue", "trt" = "firebrick")) +
  labs(x = paste0("PC1: ", pct_var[1], "% variance"),
       y = paste0("PC2: ", pct_var[2], "% variance"),
       color = "Treatment", shape = "Cell line",
       title = "PCA: Airway Dataset") +
  theme_minimal()
```

**Reading the PCA plot:**

- **PC1** (horizontal axis) captures the most variation in the dataset. In a
  well-designed experiment, PC1 often separates the treatment groups (like it does here).
- **PC2** (vertical axis) captures the next largest source of variation. Here,
  it often reflects cell-line differences.
- If samples from the same treatment group are scattered widely, or if the
  groups do not separate at all, something may be wrong with the experiment,
  the data, or the metadata.

In the `airway` data, treated and untreated samples separate clearly along PC1,
and the four cell lines are visible along PC2. This is a clean, well-behaved
dataset -- a good baseline for learning the methods before working with messier
real-world data.

## Batch Effects

A **batch effect** is systematic technical variation introduced by when or how
samples were processed. Common sources include different sequencing runs,
different lab technicians, different reagent lots, and different processing
dates.

```{r}
# Check whether any technical variable is confounded with treatment
colData(airway) |>
  as_tibble() |>
  select(SampleName, cell, dex, Run) |>
  arrange(dex, cell)
```

In the `airway` data, each cell line contributes one treated and one untreated
sample, so cell line is **balanced** across treatment groups. This is good
experimental design: if cell line had been confounded with treatment (e.g., all
treated samples from one cell line, all untreated from another), it would be
impossible to separate the dexamethasone effect from the cell-line effect.

::: {.callout-warning}
**Batch effects are a design problem first, a statistical problem second.**
Statistical methods (e.g., including batch as a covariate in the model) can
adjust for batch effects if the batch variable is known and balanced across
treatment groups. If batches are confounded with treatment, no statistical
method can separate them. This is a problem that must be caught at the study
design stage, not the analysis stage.
:::

---

# Summary

Five things to carry into Days 12--13:

1. **Genomics data is high-dimensional ($p \gg n$).** We cannot fit a single
   regression. We test genes one at a time and correct for multiple testing.

2. **The count matrix is our starting point.** Rows are genes, columns are
   samples, values are non-negative integers. The upstream bioinformatics
   pipeline produces this; we work with what it produces.

3. **Counts require special treatment.** They are right-skewed, non-normal,
   and have variance that scales with the mean. Normalization for sequencing
   depth is required. The negative binomial model is appropriate. DESeq2
   handles all of this.

4. **The `SummarizedExperiment` keeps everything synchronized.** Always use
   `assay()`, `colData()`, and `rowData()` to access the components. Always
   inspect `colData` before modeling.

5. **Quality assessment before modeling.** Check library sizes, count
   distributions, sample-to-sample distances, and PCA. Identify and investigate
   outlier samples and potential batch effects before fitting any model.

This afternoon you will repeat these steps on a dataset you download directly
from GEO using the `GEOquery` package.
