---
title: "Day 11 (PM) Lab: Genomics Data Exploration -- Solutions"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(this.path)
library(SummarizedExperiment)
library(DESeq2)
library(pasilla)
library(patchwork)
# Some conflicted functions for this .qmd, setting defaults here:
select  <- dplyr::select
filter  <- dplyr::filter
setwd(this.dir())

# Load and align data (same as instructor demo)
counts_file <- system.file("extdata", "pasilla_gene_counts.tsv",
                           package = "pasilla")
meta_file   <- system.file("extdata", "pasilla_sample_annotation.csv",
                           package = "pasilla")

counts <- read.table(counts_file, header = TRUE, row.names = 1) |>
  as.matrix()
meta <- read.csv(meta_file, row.names = 1)
rownames(meta) <- sub("fb", "", rownames(meta))
meta <- meta[colnames(counts), ]

identical(colnames(counts), rownames(meta))

se <- SummarizedExperiment(
  assays  = list(counts = counts),
  colData = meta
)
```

---

# Exercise 1 -- Inspect the Count Matrix and Sample Metadata

## Part a: First 6 rows of the count matrix

```{r}
assay(se)[1:6, ]
```

Row names are Drosophila gene IDs (e.g., `FBgn0000003`). Column names are
sample identifiers. Each value is the number of sequencing reads that mapped to
that gene in that sample -- a non-negative integer.

## Part b: Dimensions

```{r}
cat("Genes:  ", nrow(se), "\n")
cat("Samples:", ncol(se), "\n")
```

The dataset has 14,599 genes and 7 samples.

## Part c: Samples per condition

```{r}
table(colData(se)$condition)
```

There are 3 treated and 4 untreated samples. The experiment is slightly
unbalanced (3 vs. 4), but not severely so. This is common in RNA-seq
experiments where one sample fails quality control or is lost during
preparation.

## Part d: Library type balance

```{r}
table(colData(se)$condition, 
      colData(se)$type)
```

Two of the three treated samples are paired-end, while two of the four untreated samples
are paired-end. So, there are some differences in the proportion of paired-end samples per treatment. On Day 12, we will run a model that will include `type` as a covariate in the design formula.

---

# Exercise 2 -- Count Distribution and Low-Count Filtering

## Part a: Raw and log2-transformed count histograms

```{r}
#| fig-width: 9
#| fig-height: 4
counts_s1 <- assay(se)[, 1]

p1 <- tibble(count = counts_s1) |>
  filter(count > 0) |>
  ggplot(aes(x = count)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Raw counts (non-zero only)",
       x = "Count", y = "Genes") +
  theme_minimal()

p2 <- tibble(count = counts_s1) |>
  filter(count > 0) |>
  mutate(log2count = log2(count + 1)) |>
  ggplot(aes(x = log2count)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  labs(title = "log2(count + 1)",
       x = "log2(count + 1)", y = "Genes") +
  theme_minimal()

p1 + p2
```

The raw count distribution is extremely right-skewed: the vast majority of
genes have very low counts and a small number of highly expressed genes have
counts in the tens of thousands. After log2 transformation, the distribution
is much more symmetric and roughly bell-shaped. This demonstrates why
log-transformation (or the variance-stabilizing transformation used for PCA)
is necessary before any distance-based analysis -- the raw scale is dominated
by a small number of highly expressed genes.

## Part b: Zero-count genes

```{r}
n_zero  <- sum(rowSums(assay(se)) == 0)
n_total <- nrow(se)

cat("Genes with zero total counts:", n_zero, "\n")
cat("Fraction of all genes:       ", round(n_zero / n_total * 100, 1), "%\n")
```

About 2,200 genes (roughly 15% of all genes in the annotation) have zero
counts across all seven samples. These genes carry no information about
differential expression and should be removed before modeling.

## Part c: Filtering low-count genes

```{r}
keep       <- rowSums(assay(se)) >= 10
se_filtered <- se[keep, ]

cat("Genes before filtering:", nrow(se),          "\n")
cat("Genes after filtering: ", nrow(se_filtered), "\n")
cat("Genes removed:         ", nrow(se) - nrow(se_filtered), "\n")
```

Note that subsetting `se[keep, ]` automatically keeps the count matrix, column
metadata, and row metadata synchronized -- one of the key benefits of the
`SummarizedExperiment` structure.

## Part d: Density plot of log2 counts across samples

```{r}
#| fig-width: 8
#| fig-height: 4
#| message: false
assay(se_filtered) |>
  as_tibble(rownames = "gene") |>
  pivot_longer(-gene, names_to = "sample", values_to = "count") |>
  mutate(log2count = log2(count + 1)) |>
  left_join(
    colData(se_filtered) |>
      as_tibble(rownames = "sample") |>
      select(sample, condition),
    by = "sample"
  ) |>
  ggplot(aes(x = log2count, color = condition, group = sample)) +
  geom_density(linewidth = 0.8) +
  scale_color_manual(values = c("untreated" = "steelblue",
                                "treated"   = "firebrick")) +
  labs(x = "log2(count + 1)", y = "Density",
       color = "Condition",
       title = "Count Distributions Across Samples (Filtered)") +
  theme_minimal()
```

The seven density curves overlap closely in shape and position. No single
sample is obviously shifted or shaped differently from the others, which is
a good sign that there are no grossly outlying samples or failed libraries.
The small differences between curves are expected and will be handled by
DESeq2's size factor normalization in tomorrow's analysis.

---

# Exercise 3 -- PCA

## Part a and b: VST and PCA plot

```{r}
#| fig-width: 6
#| fig-height: 5
# Build DESeqDataSet
dds <- DESeqDataSet(se_filtered, design = ~ condition )
dds$condition <- relevel(dds$condition, ref = "untreated")

# Variance-stabilizing transformation
vsd <- vst(dds, blind = TRUE)

# Get PCA data
pca_data <- plotPCA(vsd, intgroup = c("condition", "type"),
                    returnData = TRUE)
pct_var  <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2,
                     color = condition, shape = type)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("untreated" = "steelblue",
                                "treated"   = "firebrick")) +
  labs(x = paste0("PC1: ", pct_var[1], "% variance"),
       y = paste0("PC2: ", pct_var[2], "% variance"),
       color = "Condition", shape = "Library type",
       title = "PCA: pasilla Dataset (VST)") +
  theme_minimal()
```

## Part c: Which PC separates treatment groups?

PC1 separates the treated and untreated samples, capturing approximately
56% of the total variance. The treated
samples (red) cluster on one side of PC1 and the untreated samples (blue)
cluster on the other, confirming that the treatment has a strong and consistent
effect on global gene expression.

## Part d: Does library type drive any separation?

Looking at the shapes (circle = single-read, triangle = paired-end), there is a clear difference between library type along PC2. Here, single-read samples have large positive scores on PC2, while paired-end samples have large negative scores on PC2. This is worrisome, as it indicates that a non-biological element is having a notable impact on the data.

When technical variables dominate the PCA plot instead of the biological
treatment, it indicates a quality problem: samples are clustering
by batch or protocol rather than by biology, making the downstream differential
expression analysis unreliable or uninterpretable. Here, we at least have condition have a clear impact (along with library type).
