---
title: "Day 11 (PM) Lab: Genomics Data Exploration"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This afternoon you will practice the same quality assessment workflow from the
morning lecture on a new dataset. The goal is repetition with a different
example -- the same questions, the same tools, a different experiment.

We will use the `pasilla` dataset: an RNA-seq experiment in *Drosophila
melanogaster* (fruit fly) cells comparing samples where the *pasilla* gene was
knocked down by RNAi to untreated controls. The *pasilla* gene encodes an RNA-
binding protein involved in splicing regulation.

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** Follow along as we load the dataset
  and run through the setup together.
- **Part 2 (on your own, ~70 min):** Three exercises following the quality
  assessment checklist from the morning.

::: {.callout-note}
If `pasilla` is not installed, run:
```r
BiocManager::install("pasilla")
```
:::

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(this.path)
library(SummarizedExperiment)
library(DESeq2)
library(pasilla)
setwd(this.dir())
```

## Loading the Dataset

The `pasilla` package ships with the count matrix and sample metadata as
separate files. We load them and assemble a `SummarizedExperiment` the same
way DESeq2 expects.

```{r}
# Path to the count matrix bundled with the pasilla package
counts_file <- system.file("extdata", "pasilla_gene_counts.tsv",
                           package = "pasilla")

# Path to the sample metadata
meta_file <- system.file("extdata", "pasilla_sample_annotation.csv",
                         package = "pasilla")

# Read both files
counts <- read.table(counts_file, header = TRUE, row.names = 1) |>
  as.matrix()

meta <- read.csv(meta_file, row.names = 1)
```

```{r}
# Check dimensions
dim(counts)
dim(meta)

# The metadata row names must match the count matrix column names
# The pasilla metadata uses slightly different names -- align them
rownames(meta) <- sub("fb", "", rownames(meta))
meta <- meta[colnames(counts), ]

# Confirm alignment
identical(colnames(counts), rownames(meta))
```

```{r}
# Inspect the metadata
meta
```

The `condition` column encodes our treatment variable: `treated` (pasilla
knockdown) vs. `untreated` (control). The `type` column tells us whether each
library was single-read or paired-end sequencing -- a potential technical
variable to keep in mind.

```{r}
# Build the SummarizedExperiment
se <- SummarizedExperiment(
  assays  = list(counts = counts),
  colData = meta
)

se
```

```{r}
# Library sizes
tibble(
  sample    = colnames(se),
  lib_size  = colSums(assay(se)),
  condition = colData(se)$condition
) |>
  ggplot(aes(x = sample, y = lib_size / 1e6, fill = condition)) +
  geom_col(color = "white") +
  scale_fill_manual(values = c("untreated" = "steelblue",
                               "treated"   = "firebrick")) +
  labs(x = "Sample", y = "Total reads (millions)",
       fill = "Condition",
       title = "Library Sizes: pasilla Dataset") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document. Title it
`"Day 11 Lab: Genomics Data Exploration"`.

**Step 2.** Copy the setup chunk from Part 1 into your document (including the
data loading and `SummarizedExperiment` construction steps). You need the `se`
object to be available before starting the exercises.

**Step 3.** Work through Exercises 1--3. Add 2--3 sentences of interpretation
after each result.

**Step 4.** Render when finished.

---

**Exercise 1 -- Inspect the Count Matrix and Sample Metadata.**

a. Extract the count matrix using `assay(se)` and print the first 6 rows and
   all columns. What do the row names represent? What do the column names
   represent?

b. How many genes are in the dataset? How many samples?

c. Look at the sample metadata with `colData(se)`. How many samples are in
   each condition (`treated` vs. `untreated`)? Is the experiment balanced?

d. The `type` column records whether each sample was sequenced as single-read
   or paired-end. Check whether library type is balanced
   across the two conditions. Is there a potential confound here?

```{r}
# your code here
```

---

**Exercise 2 -- Count Distribution and Low-Count Filtering.**

a. For the first sample, plot a histogram of raw counts (non-zero values only)
   and a histogram of `log2(count + 1)`. Place them side by side using
   `library(patchwork)`. Describe the shape of each.

b. How many genes have zero total counts across all samples? What fraction of
   all genes does this represent?

c. How many genes remain after filtering to those with at least 10 total counts?
   Filter the `se` object using `se[keep, ]` and confirm the new dimensions.

d. Make a density plot of `log2(count + 1)` values, one curve per sample,
   colored by condition. Use the filtered dataset. Do the curves overlap
   closely?

```{r}
# your code here
```

---

**Exercise 3 -- PCA.**

a. Build a `DESeqDataSet` from the filtered `se` object and apply the
   variance-stabilizing transformation:
   ```r
   dds <- DESeqDataSet(se_filtered, design = ~ condition)
   vsd <- vst(dds, blind = TRUE)
   ```

b. Produce a PCA plot using `plotPCA()` with `returnData = TRUE` and plot it
   manually with `ggplot2` so you can control the colors. Color points by
   `condition` and use a different shape for `type` (single-read vs.
   paired-end).

c. Which principal component separates the treatment groups? How much of the
   total variance does it capture?

d. Does library type (`type`) appear to drive any separation in the PCA? What
   would it mean if technical variables dominated the plot instead of the
   biological treatment?

```{r}
# your code here
```
