---
title: "Day 11 (PM) Lab: Genomics Data -- Exploration and Quality Assessment"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This afternoon you will apply every concept from the morning lecture to a
**real dataset you download yourself from GEO**. The goal is to go from a
GEO accession number to a fully quality-assessed dataset, ready for
differential expression analysis tomorrow.

The dataset is a mouse RNA-seq experiment examining the response to
lipopolysaccharide (LPS), a component of bacterial cell walls that triggers
a strong immune response. Mice were either injected with LPS or left
untreated, and RNA was extracted from brain tissue. The study asks: which
genes change their expression in the brain during acute bacterial-driven
inflammation?

**GEO accession: GSE96870**

**How this lab works:**

- **Part 1 (instructor demo, ~25 min):** Follow along as we download and
  inspect the dataset together.
- **Part 2 (on your own, ~90 min):** Six exercises working through the full
  quality assessment pipeline. Create a new `.qmd` file, work through each
  exercise, render when done.

The **Quick Reference Card** at the end has the key functions for both parts.

::: {.callout-note}
This lab uses Bioconductor packages that are installed differently from CRAN
packages. If any package is missing, install with:
```r
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("GEOquery", "SummarizedExperiment", "DESeq2"))
```
:::

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(this.path)
library(GEOquery)
library(SummarizedExperiment)
library(DESeq2)
library(patchwork)
setwd(this.dir())
```

## Step 1: Download the Dataset from GEO

`getGEO()` downloads the study metadata and expression data associated with
a GSE accession number. The `GSEMatrix = TRUE` argument retrieves the processed
expression matrix (what we want) rather than the raw data.

```{r}
#| cache: true
gse <- getGEO("GSE96870", GSEMatrix = TRUE)
# Download the supplementary count matrix
# GSE96870 deposits counts as a supplementary file on GEO
getGEOSuppFiles("GSE96870", makeDirectory = TRUE)

# The downloaded file lands in a GSE96870/ subfolder
# Check what was downloaded
list.files("GSE96870/")
```

`getGEO()` returns a list. For most studies, the first element is the object
we need:

```{r}
class(gse)
length(gse)
gse_obj <- gse[[1]]
class(gse_obj)
```

The object is an `ExpressionSet`, an older Bioconductor container. We will
convert it to a `SummarizedExperiment` shortly. First, let us look at what
we have.

Read the count matrix (adjust filename if different)
```{r}
counts_raw <- read_tsv("GSE96870/GSE96870_counts_brain.tsv.gz",
                       show_col_types = FALSE)

counts_raw[1:5, 1:5]
```


## Step 2: Inspect the Sample Metadata

```{r}
# Sample metadata lives in pData() for ExpressionSet objects
# (equivalent to colData() for SummarizedExperiment)
pd <- pData(gse_obj)
dim(pd)
names(pd)
```

GEO metadata columns have machine-generated names. The useful columns are
usually those with `characteristics_ch1` in the name:

```{r}
pd |>
  select(title, starts_with("characteristics")) |>
  head(10)
```

This tells us the experimental groups and any relevant covariates. Identify
which column encodes the treatment variable before doing anything else.

## Step 3: Inspect the Expression Matrix

```{r}
# The expression matrix: genes x samples
expr <- exprs(gse_obj)
dim(expr)
expr[1:5, 1:4]
```

Note the row names: depending on how the submitter prepared the data, these
may be probe IDs, gene symbols, or Ensembl IDs. Check what you are working
with before proceeding.

```{r}
head(rownames(expr))
```

## Step 4: Check the Count Distribution

```{r}
#| fig-width: 5
#| fig-height: 4
tibble(value = as.vector(expr[, 1])) |>
  filter(value > 0) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Expression values: first sample",
       x = "Value", y = "Number of genes") +
  theme_minimal()
```

Depending on whether the data were submitted as raw counts, normalized counts,
or log-transformed values, the distribution will look different. Raw counts are
right-skewed integers. Log-transformed values are more symmetric but may contain
non-integers or negative values. **Always check** before modeling.

## Step 5: Library Sizes

```{r}
#| fig-width: 7
#| fig-height: 4
tibble(
  sample   = colnames(expr),
  lib_size = colSums(expr)
) |>
  ggplot(aes(x = sample, y = lib_size / 1e6)) +
  geom_col(fill = "steelblue", color = "white") +
  labs(x = "Sample", y = "Total counts (millions)",
       title = "Library Sizes: GSE96870") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Library sizes should be roughly comparable across samples. A sample with a
dramatically lower total count than the others is a warning sign.

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document. Title it
`"Day 11 Lab: Genomics Data Exploration"`.

**Step 2.** Replace the template content with this YAML and setup chunk:

````markdown
---
title: "Day 11 Lab: Genomics Data Exploration"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{{r}}
library(tidyverse)
library(this.path)
library(GEOquery)
library(SummarizedExperiment)
library(DESeq2)
library(patchwork)
setwd(this.dir())
```
````

**Step 3.** Work through Exercises 1--6. For each one, write the code, run it,
and add 2--3 sentences describing what you found.

**Step 4.** Render when finished.

::: {.callout-tip}
The download in Exercise 1 will take 30--60 seconds the first time. After that,
`getGEO()` caches the file locally so subsequent runs are instant. If the
download fails, try adding `destdir = "."` to `getGEO()` to save it to your
working directory.
:::

::: {.callout-note}
Exercises 1--3 cover data acquisition and inspection. Exercises 4--5 cover
quality assessment. Exercise 6 is a challenge that takes the dataset all the
way to PCA.
:::

---

**Exercise 1 -- Download and Inspect the Data.**

a. Download GSE96870 using `getGEO("GSE96870", GSEMatrix = TRUE)`. Assign the
   result to `gse`, then extract the first element as `gse_obj`.

b. Use `dim()` and `class()` on `gse_obj` to confirm what type of object it is
   and how many genes and samples it contains.

c. Use `pData(gse_obj)` to extract the sample metadata. How many samples are
   there? What columns are present?

d. Identify the column that encodes the treatment variable (LPS vs. control).
   It will be one of the `characteristics_ch1.*` columns. Print its unique
   values with `unique()`.

e. **Discussion.** What would happen if you started a differential expression
   analysis without first verifying which column encodes treatment? What could
   go wrong?

```{r}
# your code here
```

---

**Exercise 2 -- The Expression Matrix.**

a. Extract the expression matrix using `exprs(gse_obj)`. What are its
   dimensions? What do the rows represent? What do the columns represent?

b. Print the first 6 row names with `head(rownames(...))`. Are the row names
   gene symbols, Ensembl IDs, or probe IDs? How can you tell?

c. Print the first 6 column names. Do they match the row names of `pData()`
   you found in Exercise 1?

d. Confirm the alignment between the expression matrix and the sample metadata:
   ```r
   identical(colnames(exprs(gse_obj)), rownames(pData(gse_obj)))
   ```
   This should return `TRUE`. If it does not, do not proceed until you
   understand why.

e. What is the range of values in the expression matrix? Use `range()`.
   Based on the range, are these raw counts, normalized counts, or
   log-transformed values?

```{r}
# your code here
```

---

**Exercise 3 -- Understanding the Sample Metadata.**

a. Using `pData(gse_obj)`, create a clean summary table with one row per
   sample showing: the sample title, the treatment group, and any other
   biological variables (e.g., sex, age, tissue). Use `select()` to keep
   only the relevant columns.

b. How many samples are in each treatment group? Use `count()`.

c. Is the experiment balanced? That is, are all treatment groups represented
   by the same number of samples?

d. Check whether there are any potential batch variables in the metadata
   (e.g., different sequencing runs, different processing dates). If so, is
   the batch variable balanced across treatment groups? Use a table or
   `count()` to check.

e. Based on your inspection of the metadata, write a one-paragraph description
   of the study design as you understand it: how many groups, how many samples
   per group, what biological variables are recorded, and whether any
   confounders are apparent.

```{r}
# your code here
```

---

**Exercise 4 -- Count Distribution and Filtering.**

a. Plot the distribution of expression values for the first sample using
   `geom_histogram()`. Label the x-axis clearly. Based on the shape, are these
   raw counts, normalized counts, or log-transformed values?

b. Plot the log2-transformed values for the same sample
   (`log2(expr[, 1] + 1)` if counts; `expr[, 1]` if already log-transformed).
   How does the shape change?

c. How many genes have zero total counts across all samples? Use
   `rowSums(exprs(gse_obj)) == 0`. What fraction of all genes does this
   represent?

d. How many genes have fewer than 10 total counts? Fewer than 1?

e. Filter to keep only genes with at least 10 total counts. How many genes
   remain? Save this filtered expression matrix as `expr_filtered`.

```{r}
# your code here
```

---

**Exercise 5 -- Library Sizes and Sample Distributions.**

a. Compute the library size (total counts) for each sample using `colSums()`.
   Report the minimum, maximum, and coefficient of variation
   (standard deviation divided by mean, times 100).

b. Make a bar chart of library sizes, colored by treatment group. Do library
   sizes appear balanced across groups, or does one group tend to have more
   reads?

c. Make a density plot of log2-transformed expression values, with one curve
   per sample, colored by treatment group. Use `pivot_longer()` to reshape the
   matrix first. The curves should overlap closely -- large departures suggest
   a sample-level problem.

d. Based on the library sizes and count distributions, does any sample look
   like an outlier? What would you do if you found one?

```{r}
# your code here
```

---

**Exercise 6 (Challenge) -- Sample-to-Sample Distances and PCA.**

To run this exercise you need to build a `DESeqDataSet` from the expression
matrix and sample metadata. This requires raw integer counts. If the expression
values in GSE96870 are already normalized or log-transformed (check your answer
from Exercise 2e), use `round()` to convert them to integers as an
approximation for the purposes of this exercise.

a. Build a `DESeqDataSet` object:
   ```r
   # Create a SummarizedExperiment first
   # Replace "your_treatment_column" with the actual column name from pData
   col_data <- pData(gse_obj) |>
     select(treatment = your_treatment_column) |>
     mutate(treatment = factor(treatment))

   se <- SummarizedExperiment(
     assays  = list(counts = round(exprs(gse_obj))),
     colData = col_data
   )

   dds <- DESeqDataSet(se, design = ~ treatment)
   ```

b. Apply the variance-stabilizing transformation:
   ```r
   keep <- rowSums(counts(dds)) >= 10
   dds  <- dds[keep, ]
   vsd  <- vst(dds, blind = TRUE)
   ```

c. Compute and visualize sample-to-sample distances as a heatmap. Label rows
   and columns with the treatment group. Do samples cluster by treatment?

d. Produce a PCA plot using `plotPCA()` with `returnData = TRUE` (see the
   Quick Reference Card). Color points by treatment group. Which principal
   component separates the groups? How much of the total variance does it
   explain?

e. **Discussion.** In the morning lecture, we said PC1 often separates
   treatment groups in a well-designed experiment. Is that true here? What
   does PC2 appear to capture? If you saw a dataset where treatment groups did
   not separate on any principal component, what would that suggest?

```{r}
# your code here
```

---

# Quick Reference Card

## Package Installation (if needed)

```r
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("GEOquery", "SummarizedExperiment", "DESeq2", "airway"))
```

## Downloading from GEO

```r
library(GEOquery)

# Download a dataset by accession number
gse      <- getGEO("GSE96870", GSEMatrix = TRUE)
gse_obj  <- gse[[1]]          # extract the first (usually only) element

# If the download fails, try saving locally first
gse      <- getGEO("GSE96870", GSEMatrix = TRUE, destdir = ".")
```

## ExpressionSet Accessors (GEO download format)

```r
exprs(gse_obj)       # expression matrix (genes x samples)
pData(gse_obj)       # sample metadata (one row per sample)
fData(gse_obj)       # feature/gene metadata (one row per gene)
dim(gse_obj)         # dimensions
sampleNames(gse_obj) # sample IDs
featureNames(gse_obj)# gene/probe IDs
```

## SummarizedExperiment Accessors

```r
assay(se)            # count matrix
colData(se)          # sample metadata
rowData(se)          # gene metadata
dim(se)
colnames(se)
rownames(se)

# Subsetting (keeps all three components synchronized)
se_treated <- se[, se$treatment == "LPS"]
se_filtered <- se[rowSums(assay(se)) >= 10, ]
```

## Count Matrix Operations

```r
counts <- exprs(gse_obj)       # or assay(se)

# Library sizes
colSums(counts)

# Zero and low-count genes
sum(rowSums(counts) == 0)
sum(rowSums(counts) < 10)

# Filter to genes with at least 10 total counts
keep           <- rowSums(counts) >= 10
counts_filtered <- counts[keep, ]
```

## Reshaping for ggplot2

```r
# Wide matrix to long tibble (for density plots, etc.)
counts |>
  as_tibble(rownames = "gene") |>
  pivot_longer(-gene, names_to = "sample", values_to = "count") |>
  mutate(log2count = log2(count + 1))
```

## DESeq2: VST and PCA

```r
library(DESeq2)

# Build DESeqDataSet from a SummarizedExperiment
dds <- DESeqDataSet(se, design = ~ treatment)

# Filter low-count genes (required before vst)
keep <- rowSums(counts(dds)) >= 10
dds  <- dds[keep, ]

# Variance-stabilizing transformation (blind = TRUE for QC)
vsd <- vst(dds, blind = TRUE)

# PCA using plotPCA (returnData = TRUE for manual ggplot)
pca_data <- plotPCA(vsd, intgroup = "treatment", returnData = TRUE)
pct_var  <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2, color = treatment)) +
  geom_point(size = 4) +
  labs(x = paste0("PC1: ", pct_var[1], "% variance"),
       y = paste0("PC2: ", pct_var[2], "% variance")) +
  theme_minimal()
```

## Sample-to-Sample Distance Heatmap

```r
# Compute pairwise distances on VST-transformed data
sample_dists <- dist(t(assay(vsd)))
dist_matrix  <- as.matrix(sample_dists)

# Label with treatment group
rownames(dist_matrix) <- colnames(dist_matrix) <-
  colData(vsd)$treatment

heatmap(dist_matrix,
        symm    = TRUE,
        col     = colorRampPalette(c("steelblue", "white"))(50),
        margins = c(10, 10))
```

## Common Mistakes

| Mistake | What happens | How to avoid |
|---|---|---|
| Using `scale_color_manual()` with combined group labels from `plotPCA()` | Warning, no colors applied | Use `returnData = TRUE` and map color manually in `ggplot()` |
| Calling `vst()` on a raw `SummarizedExperiment` | Error about `design<-` | Convert to `DESeqDataSet` first with `DESeqDataSet(se, design = ~ group)` |
| Calling `vst()` before filtering zero-count genes | Error about zero-variance genes | Filter with `rowSums(counts(dds)) >= 10` before calling `vst()` |
| Not checking `identical(colnames(expr), rownames(pData()))` | Silent metadata misalignment | Always check before any analysis |
