---
title: "Day 12 (PM) Lab: Differential Expression Analysis -- Solutions"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(this.path)
library(SummarizedExperiment)
library(DESeq2)
library(pasilla)
library(pheatmap)
library(patchwork)
setwd(this.dir())

select <- dplyr::select
filter <- dplyr::filter
count <- dplyr::count

# ── Data loading (same as instructor demo) ─────────────────────────────────
counts_file <- system.file("extdata", "pasilla_gene_counts.tsv",
                           package = "pasilla")
meta_file   <- system.file("extdata", "pasilla_sample_annotation.csv",
                           package = "pasilla")

counts <- read.table(counts_file, header = TRUE, row.names = 1) |>
  as.matrix()
meta <- read.csv(meta_file, row.names = 1)
rownames(meta) <- sub("fb", "", rownames(meta))
meta <- meta[colnames(counts), ]

# ── DESeq2 model (same as instructor demo) ─────────────────────────────────
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = meta,
  design    = ~ type + condition
)
dds$condition <- relevel(dds$condition, ref = "untreated")
dds$type      <- relevel(dds$type,      ref = "single-read")
keep          <- rowSums(counts(dds)) >= 10
dds           <- dds[keep, ]
deseq_dds     <- DESeq(dds)
res           <- results(deseq_dds, name = "condition_treated_vs_untreated")

# ── Results tibble used throughout ────────────────────────────────────────
res_df <- as.data.frame(res) |>
  rownames_to_column("gene_id") |>
  as_tibble() |>
  arrange(padj)
```

---

# Exercise 1 -- Explore the Results Table

## Part a: Structure of the results table

```{r}
cat("Rows:", nrow(res_df), "\n")
head(res_df)
```

The table has one row per gene tested. Each row contains the gene ID, the
average normalized expression across all samples (`baseMean`), the estimated
log2 fold change between treated and untreated (`log2FoldChange`), its standard
error (`lfcSE`), a Wald test statistic (`stat`), the raw p-value, and the
Benjamini-Hochberg adjusted p-value (`padj`). There is one row for every gene
that passed the low-count filter, roughly 10,000--12,000 genes.

## Part b: Significant genes

```{r}
cat("padj < 0.05:", res_df |> filter(padj < 0.05) |> nrow(), "\n")
cat("padj < 0.01:", res_df |> filter(padj < 0.01) |> nrow(), "\n")
```

Several hundred genes pass each threshold. Moving from 0.05 to 0.01 roughly
halves the number of significant genes, as the stricter threshold accepts fewer
false discoveries.

## Part c: Direction of change

```{r}
res_df |>
  filter(padj < 0.05) |>
  mutate(direction = if_else(log2FoldChange > 0,
                                    "Upregulated",
                                    "Downregulated")) |>
  count(direction)

```

Upregulated means higher expression in the `treated` (pasilla knockdown)
samples compared with the `untreated` controls. Since pasilla is an
RNA-binding protein involved in splicing, many of the upregulated genes are
likely those normally suppressed by pasilla activity -- when pasilla is knocked
down, their expression rises. The roughly equal number of up and downregulated
genes is consistent with a transcription factor or splicing regulator that
activates some genes and represses others.

## Part d: Most significant gene

```{r}
res_df |>
  slice_min(padj, n = 1) |>
  select(gene_id, log2FoldChange, padj)
```

The most significant gene has a large negative log2 fold change and a
near-zero adjusted p-value, indicating both a biologically meaningful and
statistically robust change in expression after pasilla knockdown.

## Part e: Top 10 genes

```{r}
res_df |>
  slice_min(padj, n = 10) |>
  select(gene_id, log2FoldChange, padj) |>
  mutate(log2FoldChange = round(log2FoldChange, 3),
         padj           = formatC(padj, format = "e", digits = 2))
```

---

# Exercise 2 -- Size Factors and Dispersion

## Part a: Size factors

```{r}
sf <- sizeFactors(deseq_dds)
sort(sf)
```

```{r}
cat("Sample with largest size factor: ", names(which.max(sf)), round(max(sf), 3), "\n")
cat("Sample with smallest size factor:", names(which.min(sf)), round(min(sf), 3), "\n")
```

A large size factor means the sample was sequenced to a greater depth than the
typical sample in this dataset: it has proportionally more reads and would
appear to have higher counts for every gene if we compared raw counts directly.
DESeq2 divides each sample's raw counts by its size factor before testing, so
that differences in count magnitude reflect biology rather than sequencing
depth.

## Part b: Dispersion plot

```{r}
#| fig-width: 6
#| fig-height: 5
plotDispEsts(deseq_dds,
             main = "Dispersion Estimates: pasilla",
             cex  = 0.4)
```

The black dots (per-gene dispersion estimates from the data alone) are
scattered widely around the fitted red trend line, especially at low expression
levels where small counts make per-gene estimates very noisy. The blue dots
(final shrunken estimates) are pulled toward the red trend, reducing the
influence of extreme per-gene estimates. A small number of black open circles
appear above the trend -- these are genes with unusually high per-gene
dispersion that DESeq2 flags as potential outliers and does not shrink toward
the trend, because their extreme dispersion may represent genuine biological
variability rather than estimation noise.

## Part c: Dispersion for three genes

```{r}
# Creating a tibble with the results (IMPORTANT: do not sort by padj.)
res_df <- as.data.frame(res) |>
  rownames_to_column("gene_id") |>
  as_tibble() 

# This gives the dispersions for each row of the DESeqDataSet.
disp_vals <- dispersions(deseq_dds)

# I'm going to add the dispersion to the results.
# This is why it was important that we didn't res_df by sort by padj.
res_df <- res_df |> 
  mutate(
    disp_val = disp_vals)

res_df |> 
  slice_max(baseMean, n=1) |> 
  mutate(Category = "Highest_mean") |> 
  bind_rows(
    res_df |> 
      filter(baseMean > 0) |> 
  slice_min(baseMean, n=1) |> 
  mutate(Category = "Lowest non-zero mean") 
  ) |> 
  bind_rows(
    res_df |> 
  slice_min(padj, n=1) |> 
  mutate(Category = "Top DE gene")
  ) |> 
  arrange(disp_val) |> 
  select(gene_id, Category, baseMean, log2FoldChange, padj, disp_val)

```

The highly expressed gene (highest mean) has a small dispersion estimate (its counts are
consistent across samples relative to its mean). The lowly expressed gene has
a much larger dispersion estimate, reflecting the greater proportional
variability that is typical of low-count genes. The top DE gene has a moderate
mean expression and a dispersion consistent with that expression level -- it
is not extreme in either direction.

---

# Exercise 3 -- Volcano Plot

## Parts a and b: Volcano plot with quadrant lines

```{r}
#| fig-width: 7
#| fig-height: 5
volcano_df <- res_df |>
  filter(!is.na(padj)) |>
  mutate(significant = padj < 0.05)

ggplot(volcano_df, aes(x = log2FoldChange,
                       y = -log10(padj),
                       color = significant)) +
  geom_point(alpha = 0.5, size = 0.9) +
  scale_color_manual(values = c("FALSE" = "gray70", "TRUE" = "firebrick")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed",
             color = "gray40", linewidth = 0.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             color = "gray40", linewidth = 0.8) +
  labs(x = "log2 Fold Change",
       y = expression(-log[10](padj)),
       color = "padj < 0.05",
       title = "Volcano Plot: pasilla Knockdown vs. Control") +
  theme_minimal()
```

## Part c: Gene counts per quadrant

```{r}
volcano_df |>
  filter(padj < 0.05) |>
  mutate(
    quadrant = case_when(
      log2FoldChange >  1 ~ "Upper right (sig + up > 2x)",
      log2FoldChange < -1 ~ "Upper left  (sig + down > 2x)",
      TRUE                ~ "Upper center (sig, small LFC)"
    )
  ) |>
  count(quadrant)
```

## Part d: Interpretation of quadrants

Genes in the top-right corner are both statistically significant and strongly
upregulated after pasilla knockdown -- these are the most compelling candidates
for follow-up, being both reliable (low false discovery probability) and
biologically meaningful in magnitude. Genes in the bottom-center have small
absolute log2 fold changes and low significance: the expression change is
negligible or the evidence is weak, and these genes would not be prioritized.

---

# Exercise 4 -- Heatmap of Top DE Genes

## Parts a--d: VST, subset, scale, and heatmap

```{r}
#| fig-width: 7
#| fig-height: 7
vsd <- vst(deseq_dds, blind = FALSE)

# Top 25 most significant genes
top25_ids <- res_df |>
  filter(!is.na(padj)) |>
  slice_min(padj, n = 25) |>
  pull(gene_id)

mat        <- assay(vsd)[top25_ids, ]
mat_scaled <- t(scale(t(mat)))

col_anno <- colData(vsd) |>
  as_tibble(rownames = "sample") |>
  select(sample, condition, type) |>
  mutate(condition = as.character(condition),
         type      = as.character(type)) |>
  column_to_rownames("sample")

pheatmap(mat_scaled,
         annotation_col  = col_anno,
         cluster_rows    = TRUE,
         cluster_cols    = TRUE,
         show_rownames   = TRUE,
         show_colnames   = FALSE,
         color           = colorRampPalette(
           c("steelblue", "white", "firebrick"))(100),
         main            = "Top 25 DE Genes (VST, row-scaled)")
```

The treated and untreated samples cluster separately along the columns,
confirming that the top DE genes genuinely separate the two experimental
groups. A two-block pattern is visible in the rows: one block of genes is
consistently higher (red) in treated samples and lower (blue) in untreated,
and a second block runs in the opposite direction. This is the expected pattern
for a strong, consistent treatment effect.

## Part e: Library type effect

Looking at the `type` column annotation, the two paired-end untreated sample
and the three paired-end treated samples do not cluster separately from the
single-read untreated samples within each condition. This confirms what we
saw in the Day 11 PCA: library type is not the dominant source of variation in
this dataset. Nevertheless, because it is partially confounded with condition
(all treated samples happen to be paired-end), including it in the design
formula is the correct modeling choice. Failing to do so would leave a potential
technical confound unaddressed, even if the practical impact appears small here.

---

# Exercise 5 -- Model Comparison: With vs. Without Library Type

## Part a: Fit the unadjusted model

```{r}
dds_simple <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = meta,
  design    = ~ condition
)
dds_simple$condition <- relevel(dds_simple$condition, ref = "untreated")
keep_simple          <- rowSums(counts(dds_simple)) >= 10
dds_simple           <- dds_simple[keep_simple, ]
dds_simple           <- DESeq(dds_simple)
res_simple           <- results(dds_simple,
                                name = "condition_treated_vs_untreated")

res_simple_df <- as.data.frame(res_simple) |>
  rownames_to_column("gene_id") |>
  as_tibble() |>
  arrange(padj)
```

## Part b: Number of significant genes

```{r}
tibble(
  Model        = c("Adjusted (type + condition)", "Unadjusted (condition only)"),
  Sig_padj_05  = c(
    res_df        |> filter(padj < 0.05) |> nrow(),
    res_simple_df |> filter(padj < 0.05) |> nrow()
  )
)
```

Adjusting for library type typically increases the number of significant genes.
By including `type` in the model, we explain some of the residual variability
in the data, which tightens the estimates and increases power for detecting true
treatment effects.

## Part c: LFC comparison for top 20 genes

```{r}
#| fig-width: 6
#| fig-height: 5
top20_ids <- res_df |>
  filter(!is.na(padj)) |>
  slice_min(padj, n = 20) |>
  pull(gene_id)

lfc_compare <- tibble(
  gene_id   = top20_ids,
  LFC_adj   = res_df$log2FoldChange[match(top20_ids, res_df$gene_id)],
  LFC_simple = res_simple_df$log2FoldChange[
    match(top20_ids, res_simple_df$gene_id)]
)

ggplot(lfc_compare, aes(x = LFC_adj, y = LFC_simple)) +
  geom_point(color = "steelblue", size = 3) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "firebrick") +
  labs(x = "LFC: adjusted model (type + condition)",
       y = "LFC: unadjusted model (condition only)",
       title = "LFC Comparison: Top 20 DE Genes") +
  theme_minimal()
```

The LFC estimates are generally similar between the two models for the top
genes, with points falling close to the diagonal. This is consistent with our
observation that library type is not the dominant driver of expression
differences in this dataset. Larger departures from the diagonal would indicate
that unadjusted estimates were confounded by library type.

## Part d: MA plot comparison

```{r}
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 2))
plotMA(res,        ylim = c(-5, 5), alpha = 0.05, colSig = "firebrick",
       main = "Adjusted (type + condition)")
plotMA(res_simple, ylim = c(-5, 5), alpha = 0.05, colSig = "firebrick",
       main = "Unadjusted (condition only)")
par(mfrow = c(1, 1))
```

The two MA plots look broadly similar, but the adjusted model typically has
slightly more red (significant) points, particularly among genes with moderate
to high expression. The unadjusted model shows somewhat wider scatter in the
non-significant genes, reflecting the extra unexplained variance from not
modeling library type.

## Part e: Discussion

Adjusting for library type changes the results because library type (single-read
vs. paired-end) introduces systematic technical differences in which reads are
captured and counted. In this experiment, all three treated samples happen to be
paired-end while most untreated samples are single-read. If we fit a model with
only `condition`, the model cannot distinguish between expression differences due
to the knockdown and expression differences due to the sequencing protocol. The
library-type effect gets absorbed into the condition coefficient, potentially
biasing the treatment estimates.

This is the same confounding logic from Day 8's logistic regression: when a
technical variable is correlated with the exposure of interest, failing to
adjust for it produces biased estimates of the effect you care about. Including
`type` in the design formula tells DESeq2 to estimate the library-type effect
separately and then test for the treatment effect after removing it -- the
genomics equivalent of an adjusted odds ratio.

In this particular dataset, the confounding appears modest (the LFC estimates
are similar between models), probably because the biological treatment effect is
large enough to dominate any library-type artifact. But the correct approach is
to adjust regardless, because we cannot know in advance how large the
confounding will be.
