---
title: "Day 12: Differential Expression Analysis with DESeq2"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(this.path)
library(SummarizedExperiment)
library(DESeq2)
library(airway)
library(patchwork)
setwd(this.dir())

data(airway)
```

# Overview

Yesterday we learned how genomics data is structured: count matrices,
`SummarizedExperiment` objects, and quality assessment. Today we answer the
scientific question that motivated the experiment:

> **Which genes change their expression between dexamethasone-treated and
> untreated airway smooth muscle cells?**

This is called **differential expression (DE) analysis**. The core idea is
simple: for each gene, test whether the mean count differs between groups.
The challenge is doing this correctly for 20,000 genes simultaneously, with
only a handful of replicates per group, on count data that violates the
assumptions of a standard t-test.

We will use DESeq2, the current standard tool for RNA-seq differential
expression. Three examples:

- **Example 1:** Building the DESeq2 model -- data input, filtering, and
  setup.
- **Example 2:** Running the analysis and interpreting the results table.
- **Example 3:** Visualization -- MA plots, volcano plots, and heatmaps.

We also show how to adjust for a covariate (cell line) in the design formula,
connecting this directly to the confounding framework from Days 8--9.

---

# Why Not a t-Test?

Before fitting anything, it is worth understanding why standard tools do not
work on raw RNA-seq counts.

```{r}
counts <- assay(airway)

# Pull counts for one gene across all 8 samples
gene1 <- counts["ENSG00000103196", ]
gene1
```

We have 4 treated and 4 untreated observations for this gene. A t-test would
assume these 8 numbers are approximately normal with equal variance. Let us
look at the count distribution across all genes in one sample to see whether
that assumption holds:

```{r}
#| fig-width: 8
#| fig-height: 4
p1 <- tibble(count = counts[, 1]) |>
  filter(count > 0) |>
  ggplot(aes(x = count)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Count distribution (log x-axis)",
       x = "Count (log scale)", y = "Genes") +
  theme_minimal()

# Mean-variance relationship across genes
gene_stats <- tibble(
  mean_count = rowMeans(counts),
  var_count  = apply(counts, 1, var)
) |>
  filter(mean_count > 0, var_count > 0)

p2 <- gene_stats |>
  slice_sample(n = 3000) |> #Randomly sample 3000 rows (out of >33K)
  ggplot(aes(x = mean_count, y = var_count)) +
  geom_point(alpha = 0.15, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0,
              color = "firebrick", linetype = "dashed", linewidth = 1) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Mean vs. variance (log-log scale)",
       subtitle = "Red line = Poisson (variance = mean)",
       x = "Mean count", y = "Variance") +
  theme_minimal()

p1 + p2
```

Two problems are immediately visible:

1. **Right-skewed counts.** Even on a log scale, the distribution is far from
   normal.
2. **Variance grows with the mean.** The variance is consistently above the
   red dashed line (which would be exact Poisson). This is called
   **overdispersion**, and it is the rule rather than the exception in
   RNA-seq data.

The appropriate model is the **negative binomial distribution**:
$\text{Var}(Y) = \mu + \alpha\mu^2$, where $\alpha > 0$ is the dispersion
parameter. DESeq2 estimates $\alpha$ separately for each gene and uses it
when testing for differential expression.

---

# Example 1: Building the DESeq2 Model

## Verifying Data Alignment

Before building the model, confirm that the count matrix and sample metadata
are aligned. This is the same check we did in the Day 11 lab.

```{r}
sample_info <- as.data.frame(colData(airway))
sample_info |> select(SampleName, cell, dex)
```

```{r}
identical(colnames(counts), rownames(sample_info))
```

## Building the DESeqDataSet

The `DESeqDataSet` bundles the count matrix, sample metadata, and experimental
design into one object. The design formula tells DESeq2 which variable to test.

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = sample_info,
  design    = ~ dex        # compare treated vs. untreated
)
dds
```

## Filtering Low-Count Genes

Genes with very few counts across all samples carry no information about
differential expression. Including them wastes computational time and slightly
increases the multiple testing burden. We keep only genes with at least 10
total reads across all samples.

```{r}
keep <- rowSums(counts(dds)) >= 10
dds  <- dds[keep, ]

cat("Genes before filtering:", nrow(counts), "\n")
cat("Genes after filtering: ", nrow(dds),    "\n")
cat("Genes removed:         ", nrow(counts) - nrow(dds), "\n")
```

## Setting the Reference Level

DESeq2 calculates fold changes as treated divided by the reference. We want
the reference to be the untreated group so that a positive fold change means
the gene went up after treatment.

```{r}
dds$dex <- relevel(dds$dex, ref = "untrt")
levels(dds$dex)
```

---

# Example 2: Running the Analysis and Interpreting Results

## Running DESeq2

A single call to `DESeq()` performs the entire workflow:

1. Estimates size factors (normalizes for sequencing depth).
2. Estimates dispersion for every gene (borrowing strength across genes).
3. Fits the negative binomial model for every gene.
4. Runs a Wald test for the treatment effect.

```{r}
deseq_dds <- DESeq(dds)
```

## Inspecting Size Factors

Size factors capture the sequencing depth of each sample. Samples with more
reads get a larger size factor; DESeq2 divides raw counts by the size factor
before comparing.

```{r}
sizeFactors(deseq_dds)
```

Samples range from roughly 0.7 to 1.3. If one sample had a size factor of 5
(five times the typical depth), that would be a sign of a badly outlying
library and worth investigating.

## The Dispersion Plot

DESeq2 estimates a dispersion parameter for every gene. With only 8 samples,
per-gene estimates are noisy. DESeq2 stabilizes them by shrinking each
estimate toward a trend fitted across all genes -- borrowing strength in the
same spirit as partial pooling in mixed models.

```{r}
#| fig-width: 6
#| fig-height: 5
plotDispEsts(deseq_dds,
             main = "Dispersion Estimates",
             cex  = 0.4)
```

**Reading this plot:**

- **Black dots:** per-gene dispersion estimates from the data alone. Noisy,
  especially at low expression.
- **Red curve:** the fitted dispersion trend across all genes.
- **Blue dots:** the final shrunken estimates used in testing. They are pulled
  toward the red curve, reducing the influence of extreme per-gene estimates.

Genes with very high per-gene estimates (outliers far above the trend) are
shown as open black circles -- DESeq2 flags these and does not shrink them,
because extreme dispersion may be genuine biology rather than noise.

## Extracting Results

```{r}
res <- results(deseq_dds)
res
```

```{r}
summary(res)
```

The summary tells us how many genes were tested, how many passed the adjusted
p-value threshold of 0.05 (the default), and how many had very low counts
(excluded from testing).

**What each column means:**

- `baseMean`: average normalized count across all 8 samples. Higher values
  mean the gene is more abundantly expressed overall.
- `log2FoldChange`: the estimated change in expression on a log2 scale.
  LFC = 1 means treated cells have twice the expression; LFC = -1 means half.
- `lfcSE`: the standard error of the LFC estimate. Wide SE means the estimate
  is uncertain.
- `stat`: the Wald test statistic (LFC divided by lfcSE).
- `pvalue`: raw p-value. Do not use this for calling significant genes.
- `padj`: Benjamini-Hochberg adjusted p-value. **Use this.**

::: {.callout-important}
**Always use `padj`, not `pvalue`.** With 20,000 tests, a raw p-value
threshold of 0.05 gives roughly 1,000 false positives even if no gene is
truly differentially expressed. The adjusted p-value controls the false
discovery rate: if we call all genes with `padj < 0.05` significant, we
expect at most 5% of those calls to be wrong.
:::

## Top Differentially Expressed Genes

```{r}
res_df <- as.data.frame(res) |>
  rownames_to_column("gene_id") |>
  as_tibble() |>
  arrange(padj)

head(res_df)
```

The most significant genes have adjusted p-values near zero and large absolute
fold changes. Let us add gene symbols from the `rowData` to make these
interpretable:

```{r}
gene_info <- as.data.frame(rowData(airway)) |>
  select(gene_id, gene_name, gene_biotype)

res_annotated <- res_df |>
  left_join(gene_info, by = "gene_id") |>
  select(gene_name, gene_id, gene_biotype, baseMean, log2FoldChange, padj) |>
  filter(!is.na(padj)) |>
  arrange(padj)

head(res_annotated, 10)
```

```{r}
# How many significant at padj < 0.05?
res_annotated |>
  filter(padj < 0.05) |>
  summarize(
    n_sig        = n(),
    n_up         = sum(log2FoldChange > 0),
    n_down       = sum(log2FoldChange < 0)
  )
```

## Interpreting the Top Hits

```{r}
# Top 5 upregulated and top 5 downregulated
res_annotated |>
  filter(padj < 0.05) |>
  slice_max(log2FoldChange,  n = 5) |>
  bind_rows(
    res_annotated |>
      filter(padj < 0.05) |>
      slice_min(log2FoldChange, n = 5)
  ) |>
  select(gene_name, log2FoldChange, padj) |>
  mutate(direction = if_else(log2FoldChange > 0, "Up", "Down"),
         log2FoldChange = round(log2FoldChange, 2),
         padj = formatC(padj, format = "e", digits = 2))
```

Dexamethasone is a glucocorticoid -- an anti-inflammatory steroid. Many of the
top upregulated genes are known glucocorticoid-responsive genes, and the top
downregulated genes are often involved in pro-inflammatory signaling. This
biological coherence is a good sign that the analysis is working correctly.

---

# Example 3: Visualization

## MA Plot

The MA plot shows the relationship between a gene's mean expression level and
its log2 fold change. It is the standard first visualization for a DE analysis.

```{r}
#| fig-width: 6
#| fig-height: 5
plotMA(res, ylim = c(-6, 8),
       main = "MA Plot: Dexamethasone Treatment",
       colSig = "firebrick",
       alpha = 0.05)
abline(h = c(-1, 1), lty = 2, col = "gray40")
```

**Reading the MA plot:**

- Each point is one gene. The x-axis is mean expression (log scale); the
  y-axis is log2 fold change.
- Red points are significant (padj < 0.05); grey points are not.
- The dashed lines at LFC = +1 and -1 mark a two-fold change in either
  direction.
- Genes at the far left (low mean expression) have noisy LFC estimates --
  the cloud of non-significant points fans out widely at low counts. This is
  expected and is why filtering low-count genes matters.

## Volcano Plot

The volcano plot combines fold change (effect size) on the x-axis with
statistical significance on the y-axis. Genes in the top corners are the most
interesting: large effect and high confidence.

```{r}
#| fig-width: 7
#| fig-height: 5
volcano_df <- res_annotated |>
  filter(!is.na(padj)) |>
  mutate(
    significant = padj < 0.05,
    label_gene  = if_else(padj < 1e-20 | abs(log2FoldChange) > 4,
                          gene_name, NA_character_)
  )

ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(data = filter(volcano_df, !significant),
             color = "gray70", alpha = 0.4, size = 0.8) +
  geom_point(data = filter(volcano_df, significant),
             color = "firebrick", alpha = 0.6, size = 1) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed",
             color = "gray40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             color = "gray40") +
  ggrepel::geom_text_repel(aes(label = label_gene),
                           size = 3, max.overlaps = 15,
                           na.rm = TRUE) +
  labs(x = "log2 Fold Change",
       y = expression(-log[10](padj)),
       title = "Volcano Plot: Dexamethasone vs. Untreated",
       subtitle = "Red = significant (padj < 0.05)") +
  theme_minimal()
```

::: {.callout-tip}
`ggrepel::geom_text_repel()` labels selected points without overlapping text.
If `ggrepel` is not installed, run `install.packages("ggrepel")` or simply
remove the label layer.
:::

## Heatmap of Top DE Genes

A heatmap of the most significant genes shows the expression pattern across all
samples simultaneously, making it easy to see whether the groups separate
cleanly.

```{r}
#| fig-width: 7
#| fig-height: 6
library(pheatmap)

# Variance-stabilizing transformation for visualization
vsd <- vst(deseq_dds, blind = FALSE)

# Take the 30 most significant genes
top_genes <- res_annotated |>
  filter(!is.na(padj)) |>
  slice_min(padj, n = 30) |>
  pull(gene_id)

# Extract VST counts for those genes
mat <- assay(vsd)[top_genes, ]
rownames(mat) <- res_annotated$gene_name[match(top_genes,
                                               res_annotated$gene_id)]

# Scale each gene (row) to zero mean and unit variance
mat_scaled <- t(scale(t(mat)))

# Column annotation showing treatment and cell line
col_anno <- colData(vsd) |>
  as_tibble(rownames = "sample") |>
  select(sample, dex, cell) |>
  column_to_rownames("sample")

pheatmap(mat_scaled,
         annotation_col  = col_anno,
         cluster_rows    = TRUE,
         cluster_cols    = TRUE,
         show_rownames   = TRUE,
         show_colnames   = FALSE,
         color           = colorRampPalette(c("steelblue","white","firebrick"))(100),
         main            = "Top 30 DE Genes (VST, row-scaled)")
```

Two things to verify from this plot:

1. **Treated samples cluster together and untreated samples cluster
   together.** If a treated sample clusters with the untreated group, it may
   be mislabeled or an outlier.
2. **Upregulated and downregulated genes form distinct blocks.** A clear
   two-block pattern (one block high in treated, one block high in untreated)
   is the expected result of a strong, consistent treatment effect.

---

# Adjusting for Cell Line

The `airway` experiment used four different cell lines. Cell lines can differ
in their baseline expression levels and potentially in how they respond to
treatment. If we do not account for this, cell-line variation inflates the
residual variance and reduces our power to detect treatment effects.

This is exactly the confounding problem we solved with multiple logistic
regression on Day 8. The solution here is the same: include the potential
confounder in the model.

```{r}
# Model 1: treatment only
dds_simple <- DESeqDataSetFromMatrix(
  countData = counts(dds),
  colData   = colData(dds),
  design    = ~ dex
)
dds_simple$dex <- relevel(dds_simple$dex, ref = "untrt")
dds_simple <- DESeq(dds_simple)
res_simple  <- results(dds_simple)

# Model 2: adjust for cell line, then test treatment
dds_cell <- DESeqDataSetFromMatrix(
  countData = counts(dds),
  colData   = colData(dds),
  design    = ~ cell + dex
)
dds_cell$dex <- relevel(dds_cell$dex, ref = "untrt")
dds_cell <- DESeq(dds_cell)
res_cell  <- results(dds_cell)
```

```{r}
# Compare the number of significant genes
tibble(
  Model      = c("Treatment only", "Adjusted for cell line"),
  Sig_genes  = c(
    sum(results(dds_simple)$padj < 0.05, na.rm = TRUE),
    sum(results(dds_cell)$padj   < 0.05, na.rm = TRUE)
  )
)
```

Adjusting for cell line typically increases the number of significant genes
because we have removed a source of unwanted variation from the residual.
This is the statistical gain from good experimental design and proper
modeling -- the same principle applies here as in any regression analysis.

::: {.callout-important}
**The adjusted model is almost always preferable when the design includes a
blocking variable.** If the four cell lines had not been included in the
design, their variation would have been absorbed into the residual, making
the test for dexamethasone less powerful. Including `cell` in the design
formula is the genomics equivalent of adjusting for a confounder in
logistic regression.
:::

```{r}
gene_info <- as.data.frame(rowData(airway)) |>
  select(gene_id, gene_name, gene_biotype)

res_cell_df <- as.data.frame(res_cell) |>
  rownames_to_column("gene_id") |>
  as_tibble() |>
  arrange(padj)

res_cell_annotated <- res_cell_df |>
  left_join(gene_info, by = "gene_id") |>
  select(gene_name, gene_id, gene_biotype, baseMean, log2FoldChange, padj) |>
  filter(!is.na(padj)) |>
  arrange(padj)

# New analysis
head(res_cell_annotated, 10)

# Old Analysis
head(res_annotated, 10)
```


```{r}
#| fig-width: 7
#| fig-height: 5
volcano_df <- res_cell_annotated |>
  filter(!is.na(padj)) |>
  mutate(
    significant = padj < 0.05,
    label_gene  = if_else(padj < 1e-20 | abs(log2FoldChange) > 4,
                          gene_name, NA_character_)
  )

ggplot(volcano_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(data = filter(volcano_df, !significant),
             color = "gray70", alpha = 0.4, size = 0.8) +
  geom_point(data = filter(volcano_df, significant),
             color = "firebrick", alpha = 0.6, size = 1) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed",
             color = "gray40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             color = "gray40") +
  ggrepel::geom_text_repel(aes(label = label_gene),
                           size = 3, max.overlaps = 15,
                           na.rm = TRUE) +
  labs(x = "log2 Fold Change",
       y = expression(-log[10](padj)),
       title = "Volcano Plot: Dexamethasone vs. Untreated",
       subtitle = "Red = significant (padj < 0.05)") +
  theme_minimal()
```

---

# Summary

The DESeq2 workflow in five steps:

1. **Build the `DESeqDataSet`** with counts, sample metadata, and design
   formula. Set the reference level for your grouping variable.

2. **Filter low-count genes.** Keep genes with at least 10 total reads
   across all samples.

3. **Run `DESeq()`.** One call handles normalization, dispersion estimation,
   model fitting, and testing for all genes.

4. **Extract results with `results()`.** Report `padj`, not `pvalue`.
   Positive LFC means upregulated in the treatment group; negative means
   downregulated.

5. **Visualize.** MA plot for an overview of all genes. Volcano plot to
   identify top candidates. Heatmap of top genes to check that the biological
   pattern is coherent.

If the experiment has a blocking variable (cell line, batch, patient), include
it in the design formula. The statistical logic is identical to adjusting for
a confounder in regression.

Tomorrow: **multiple testing in depth**. Why the Benjamini-Hochberg procedure
controls the FDR, how to interpret FDR vs. FWER, and what to do when
thousands of genes are significant or none are.
