---
title: "Day 12 (PM) Lab: Differential Expression Analysis"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This afternoon you will run a complete differential expression analysis on the
`pasilla` dataset from Day 11's PM lab. You already know what this dataset
looks like. Today you go from the count matrix to a ranked list of
differentially expressed genes.

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** We build the DESeq2 object together
  and run the model.
- **Part 2 (on your own, ~90 min):** Five exercises taking you through the
  full DE workflow.

The **Quick Reference Card** at the end has the key functions.

::: {.callout-note}
If `pasilla` is not installed: `BiocManager::install("pasilla")`
:::

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(this.path)
library(SummarizedExperiment)
library(DESeq2)
library(pasilla)
library(pheatmap)
setwd(this.dir())
```

## Load the pasilla Data

```{r}
counts_file <- system.file("extdata", "pasilla_gene_counts.tsv",
                           package = "pasilla")
meta_file   <- system.file("extdata", "pasilla_sample_annotation.csv",
                           package = "pasilla")

counts <- read.table(counts_file, header = TRUE, row.names = 1) |>
  as.matrix()

meta <- read.csv(meta_file, row.names = 1)
rownames(meta) <- sub("fb", "", rownames(meta))
meta <- meta[colnames(counts), ]

identical(colnames(counts), rownames(meta))
```

```{r}
# Reminder of the study design
as_tibble(meta) |> dplyr::select(condition, type)
```

Two conditions: `treated` (pasilla knockdown) vs. `untreated` (control).
Seven samples total: 3 treated and 4 untreated. Library type (single-read
vs. paired-end) is partially confounded with condition -- we will account
for this in the model.

## Build the DESeqDataSet and Filter

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = meta,
  design    = ~ type + condition   # adjust for library type, test condition
)

# Set reference levels
dds$condition <- relevel(dds$condition, ref = "untreated")
dds$type      <- relevel(dds$type,      ref = "single-read")

# Filter low-count genes
keep <- rowSums(counts(dds)) >= 10
dds  <- dds[keep, ]

cat("Genes retained after filtering:", nrow(dds), "\n")
```

## Run DESeq2

```{r}
deseq_dds <- DESeq(dds)
res        <- results(deseq_dds, name = "condition_treated_vs_untreated")
summary(res)
```

## Quick MA Plot

```{r}
#| fig-width: 6
#| fig-height: 5
plotMA(res, ylim = c(-5, 5),
       main  = "MA Plot: pasilla knockdown vs. control",
       alpha = 0.05,
       colSig = "firebrick")
```

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document. Title it
`"Day 12 Lab: Differential Expression"`.

**Step 2.** Copy the full setup and data-loading chunk from Part 1. You need
`deseq_dds` and `res` to be available.

**Step 3.** Work through Exercises 1--5. Add 2--3 sentences of interpretation
after each result.

**Step 4.** Render when finished.

---

**Exercise 1 -- Explore the Results Table.**

a. Convert `res` to a tibble using `as.data.frame(res) |> rownames_to_column("gene_id") |> as_tibble()`. How many rows are there? What does each row represent?

b. How many genes have `padj < 0.05`? How many have `padj < 0.01`? Use
   `filter()` and `nrow()`.

c. Of the significant genes (padj < 0.05), how many are upregulated
   (log2FoldChange > 0) and how many are downregulated? What does upregulated
   mean biologically in this experiment?

d. What is the gene with the smallest adjusted p-value? What is its
   log2FoldChange? Is it upregulated or downregulated after pasilla knockdown?

e. Print the top 10 most significant genes sorted by `padj`. For each, report
   the gene ID, log2FoldChange, and padj.

```{r}
# your code here
```

---

**Exercise 2 -- Size Factors and Dispersion.**

a. Extract and print the size factors using `sizeFactors(deseq_dds)`. Which
   sample has the largest size factor? Which has the smallest? What does a
   large size factor tell you about that sample's sequencing depth?

b. Produce the dispersion plot using `plotDispEsts(deseq_dds)`. Describe what
   you see: where are the per-gene estimates (black dots) relative to the
   fitted trend (red line)? What happened to genes with very high per-gene
   dispersion estimates?

c. Compare the per-gene dispersion estimates with the shrunken estimates for
   three genes: the gene with the highest mean expression, the gene with the
   lowest non-zero mean expression, and the most significant DE gene. You can
   extract dispersion estimates with `dispersions(deseq_dds)` and mean
   expression with `rowMeans(counts(deseq_dds, normalized = TRUE))`.

```{r}
# your code here
```

---

**Exercise 3 -- Volcano Plot.**

a. Starting from the results tibble you created in Exercise 1, create a
   volcano plot with `log2FoldChange` on the x-axis and `-log10(padj)` on the
   y-axis. Color significant genes (padj < 0.05) in red and non-significant
   genes in gray.

b. Add vertical dashed lines at LFC = -1 and LFC = 1 using `geom_vline()`, and
   a horizontal dashed line at `-log10(0.05)` using `geom_hline()`. These lines
   define the four "interesting" quadrants: genes that are both statistically
   significant and biologically meaningful in magnitude.

c. How many genes fall in each of the four quadrants defined by the LFC
   threshold (+/-1) and the significance threshold (padj < 0.05)?

d. Describe what the genes in the top-right corner of the plot have in common.
   What about the genes in the bottom-center?

```{r}
# your code here
```

---

**Exercise 4 -- Heatmap of Top DE Genes.**

a. Apply the variance-stabilizing transformation to `deseq_dds`:
   ```r
   vsd <- vst(deseq_dds, blind = FALSE)
   ```

b. Extract the gene IDs for the 25 most significant genes from your results
   tibble (sorted by padj). Use these to subset the VST matrix:
   `assay(vsd)[top_gene_ids, ]`.

c. Scale the rows of the matrix to zero mean and unit variance using
   `t(scale(t(mat)))`. This makes it easier to see relative differences
   across samples.

d. Produce a heatmap using `pheatmap()`. Include a column annotation showing
   `condition` and `type`. Do treated and untreated samples cluster separately?
   Do you see a two-block pattern in the rows (one group of upregulated genes,
   one group of downregulated)?

e. Based on the heatmap, does library type (`type`) appear to have an effect on
   expression patterns beyond the treatment effect? Does adjusting for it in
   the design formula seem important?

```{r}
# your code here
```

---

**Exercise 5 (Challenge) -- Model Comparison: With vs. Without Library Type.**

In the demo we used `design = ~ type + condition` to adjust for library type.
Here you will compare that model to one that ignores library type entirely.

a. Fit a second DESeq2 model using `design = ~ condition` only (no `type`
   adjustment). Call it `dds_simple` and `res_simple`.

b. Compare the number of significant genes (padj < 0.05) between the two
   models. Does adjusting for library type increase or decrease the number of
   significant genes?

c. For the top 20 genes from the adjusted model, compare their log2FoldChange
   values and padj values between the two models. Create a scatter plot of
   the log2FoldChange from each model on the x and y axes. How different are
   the estimates?

d. Does the MA plot look different between the two models? Use `plotMA()` on
   both result objects side by side using `par(mfrow = c(1, 2))`.

e. **Discussion.** Why might adjusting for library type change the results? In
   this experiment, library type (single-read vs. paired-end) is partially
   confounded with condition. Based on what you see, was it important to adjust
   for it? Connect this back to the confounding discussion from Day 8.

```{r}
# your code here
```

---

# Quick Reference Card

## Setup

```r
library(tidyverse)
library(SummarizedExperiment)
library(DESeq2)
library(pasilla)
library(pheatmap)
```

## Loading pasilla Data

```r
counts_file <- system.file("extdata", "pasilla_gene_counts.tsv",
                           package = "pasilla")
meta_file   <- system.file("extdata", "pasilla_sample_annotation.csv",
                           package = "pasilla")

counts <- read.table(counts_file, header = TRUE, row.names = 1) |>
  as.matrix()
meta <- read.csv(meta_file, row.names = 1)
rownames(meta) <- sub("fb", "", rownames(meta))
meta <- meta[colnames(counts), ]
```

## Building and Running DESeq2

```r
# Build DESeqDataSet
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = meta,
  design    = ~ type + condition      # adjust for type, test condition
)

# Set reference levels
dds$condition <- relevel(dds$condition, ref = "untreated")
dds$type      <- relevel(dds$type,      ref = "single-read")

# Filter low-count genes
keep <- rowSums(counts(dds)) >= 10
dds  <- dds[keep, ]

# Run DESeq2 (normalization + dispersion + testing)
deseq_dds <- DESeq(dds)

# Extract results
res <- results(deseq_dds, name = "condition_treated_vs_untreated")
summary(res)
```

## Working with Results

```r
# Convert to tibble
res_df <- as.data.frame(res) |>
  rownames_to_column("gene_id") |>
  as_tibble() |>
  arrange(padj)

# Count significant genes
res_df |> filter(padj < 0.05) |> nrow()

# Top 10
res_df |> slice_min(padj, n = 10)

# Split by direction
res_df |>
  filter(padj < 0.05) |>
  count(direction = if_else(log2FoldChange > 0, "Up", "Down"))
```

## Key Quantities

| Column | Meaning |
|---|---|
| `baseMean` | Average normalized count across all samples |
| `log2FoldChange` | log2(treated / untreated); positive = upregulated |
| `lfcSE` | Standard error of the LFC estimate |
| `pvalue` | Raw p-value (do not use for calling significance) |
| `padj` | Benjamini-Hochberg adjusted p-value -- use this |

## Visualization

```r
# MA plot
plotMA(res, ylim = c(-5, 5), alpha = 0.05, colSig = "firebrick")

# Dispersion plot
plotDispEsts(deseq_dds)

# Size factors
sizeFactors(deseq_dds)

# Volcano plot
res_df |>
  filter(!is.na(padj)) |>
  mutate(sig = padj < 0.05) |>
  ggplot(aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("FALSE" = "gray70", "TRUE" = "firebrick")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  theme_minimal()

# Variance-stabilizing transformation
vsd <- vst(deseq_dds, blind = FALSE)

# Heatmap of top genes (row-scaled)
mat <- assay(vsd)[top_gene_ids, ]
mat_scaled <- t(scale(t(mat)))
col_anno <- as.data.frame(colData(vsd)[, c("condition", "type")])
pheatmap(mat_scaled,
         annotation_col = col_anno,
         color = colorRampPalette(c("steelblue","white","firebrick"))(100))
```

## Interpreting LFC

| LFC | Meaning |
|---|---|
| LFC = 0 | No change in expression |
| LFC = 1 | Twice the expression in treated vs. untreated |
| LFC = 2 | Four times the expression |
| LFC = -1 | Half the expression in treated vs. untreated |
| LFC = -2 | One quarter the expression |
