---
title: "Day 5 (PM) Lab: Probability, Distributions, and Inference"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This lab accompanies the Day 5 lecture on probability, distributions, and
statistical inference. By the end of this session you will be able to:

- Simulate data from common distributions in R and visualize what they look like
- See the Central Limit Theorem in action through simulation
- Run and correctly interpret a one-sample t-test, two-sample t-test, paired
  t-test, chi-square test, and Wilcoxon rank-sum test
- Produce correct written interpretations of p-values and confidence intervals
- Choose the appropriate test for a given data situation

**How this lab works:**

- **Part 1 (instructor demo, ~30 min):** Distributions and the CLT via
  simulation. Follow along and run the code as you go.
- **Part 2 (on your own, ~80 min):** Six exercises applying the inference
  methods from the lecture to the Framingham Heart Study data. Create a new
  `.qmd` file and work through each exercise.

---

# Part 1: Instructor Demo

## Setup

```{r}
#| echo: false
library(tidyverse)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male            = factor(male,            levels = c(0, 1),
                                              labels = c("Female", "Male")),
    currentSmoker   = factor(currentSmoker,   levels = c(0, 1),
                                              labels = c("Non-smoker", "Smoker")),
    BPMeds          = factor(BPMeds,          levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentStroke = factor(prevalentStroke, levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentHyp    = factor(prevalentHyp,    levels = c(0, 1),
                                              labels = c("No", "Yes")),
    diabetes        = factor(diabetes,        levels = c(0, 1),
                                              labels = c("No", "Yes")),
    TenYearCHD      = factor(TenYearCHD,      levels = c(0, 1),
                                              labels = c("No", "Yes")),
    education       = factor(education,       levels = 1:4,
                                              labels = c("Some HS", "HS Diploma",
                                                         "Some College",
                                                         "College Degree"))
  )
```

```{r}
#| eval: false
library(tidyverse)
library(this.path)
this.dir <- this.dir()
setwd(this.dir)

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male            = factor(male,            levels = c(0, 1),
                                              labels = c("Female", "Male")),
    currentSmoker   = factor(currentSmoker,   levels = c(0, 1),
                                              labels = c("Non-smoker", "Smoker")),
    BPMeds          = factor(BPMeds,          levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentStroke = factor(prevalentStroke, levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentHyp    = factor(prevalentHyp,    levels = c(0, 1),
                                              labels = c("No", "Yes")),
    diabetes        = factor(diabetes,        levels = c(0, 1),
                                              labels = c("No", "Yes")),
    TenYearCHD      = factor(TenYearCHD,      levels = c(0, 1),
                                              labels = c("No", "Yes")),
    education       = factor(education,       levels = 1:4,
                                              labels = c("Some HS", "HS Diploma",
                                                         "Some College",
                                                         "College Degree"))
  )
```

## Common Distributions in R

Every distribution in R has four functions built around it. The prefix tells
you what you get:

| Prefix | What it returns | Example |
|--------|----------------|---------|
| `r` | Random draws from the distribution | `rnorm(n, mean, sd)` |
| `d` | Probability density (or mass) at a value | `dnorm(x, mean, sd)` |
| `p` | Cumulative probability up to a value | `pnorm(q, mean, sd)` |
| `q` | The value at a given cumulative probability | `qnorm(p, mean, sd)` |

We will primarily use `r` (to simulate) and `p` / `q` (to compute
probabilities and critical values).

### The Normal Distribution

```{r}
# Draw 1000 values from N(mean = 0.8, sd = 0.3) -- the crab weight example
set.seed(42)
crab_weights <- rnorm(n = 1000, mean = 0.8, sd = 0.3)

# What fraction weigh more than 1 kg?
mean(crab_weights > 1)

# The exact probability from the distribution
1 - pnorm(1, mean = 0.8, sd = 0.3)
```

```{r}
# Visualize
tibble(weight = crab_weights) |>
  ggplot(aes(x = weight)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.05,
                 fill = "steelblue", color = "white") +
  stat_function(fun = dnorm, args = list(mean = 0.8, sd = 0.3),
                color = "firebrick", linewidth = 1) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray40") +
  annotate("text", x = 1.05, y = 1.1, label = "1 kg threshold",
           hjust = 0, color = "gray40", size = 3.5) +
  labs(x = "Weight (kg)", y = "Density",
       title = "Simulated Blue Crab Weights",
       subtitle = "N(0.8, 0.3²) — red curve is the true density") +
  theme_bw()
```

The 68-95-99.7 rule in numbers:

```{r}
mu <- 0.8; sigma <- 0.3

# Probability within 1, 2, and 3 standard deviations
pnorm(mu + sigma, mu, sigma) - pnorm(mu - sigma, mu, sigma)
pnorm(mu + 2*sigma, mu, sigma) - pnorm(mu - 2*sigma, mu, sigma)
pnorm(mu + 3*sigma, mu, sigma) - pnorm(mu - 3*sigma, mu, sigma)
```

### The Binomial Distribution

```{r}
# 20 patients, each with 30% response probability
# How many respond?
set.seed(42)
rbinom(n = 1, size = 20, prob = 0.3)   # one experiment
rbinom(n = 10, size = 20, prob = 0.3)  # ten experiments

# Simulate 10,000 experiments and look at the distribution
sim_responses <- rbinom(n = 10000, size = 20, prob = 0.3)

tibble(responses = sim_responses) |>
  ggplot(aes(x = responses)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Number of responders (out of 20)",
       y = "Count (out of 10,000 simulations)",
       title = "Binomial(n = 20, p = 0.30)") +
  theme_bw()
```

```{r}
# Exact probability that exactly 8 out of 20 respond
dbinom(8, size = 20, prob = 0.3)

# Probability that 10 or more respond
1 - pbinom(9, size = 20, prob = 0.3)
```

### The Poisson Distribution

```{r}
# Mutations per kilobase, average rate lambda = 2
sim_mutations <- rpois(n = 10000, lambda = 2)

tibble(mutations = sim_mutations) |>
  ggplot(aes(x = mutations)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Mutations per kilobase",
       y = "Count",
       title = "Poisson(λ = 2)") +
  theme_bw()

# Probability of 0 mutations in a kilobase
dpois(0, lambda = 2)

# Probability of 5 or more
1 - ppois(4, lambda = 2)
```

## The Central Limit Theorem in Action

This simulation is the core of the lecture. We will draw samples from a
**skewed** (non-Normal) distribution and watch the sampling distribution of the
mean become Normal as $n$ grows.

```{r}
# Simulate from an exponential distribution -- strongly right-skewed
set.seed(123)
population <- rexp(100000, rate = 0.5)  # mean = 1/rate = 2

tibble(x = population) |>
  ggplot(aes(x = x)) +
  geom_histogram(binwidth = 0.3, fill = "steelblue", color = "white") +
  labs(x = "Value", y = "Count",
       title = "The population distribution (Exponential)",
       subtitle = "Strongly right-skewed — far from Normal") +
  theme_bw()
```

```{r}
# Function: draw B samples of size n, compute their means
sample_means <- function(population, n, B = 5000) {
  replicate(B, mean(sample(population, size = n)))
}

set.seed(123)
means_n5   <- sample_means(population, n = 5)
means_n30  <- sample_means(population, n = 30)
means_n100 <- sample_means(population, n = 100)

# Stack them for plotting
clt_data <- bind_rows(
  tibble(xbar = means_n5,   n = "n = 5"),
  tibble(xbar = means_n30,  n = "n = 30"),
  tibble(xbar = means_n100, n = "n = 100")
) |>
  mutate(n = factor(n, levels = c("n = 5", "n = 30", "n = 100")))

ggplot(clt_data, aes(x = xbar)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "steelblue", color = "white") +
  facet_wrap(~ n, scales = "free") +
  labs(x = "Sample mean", y = "Density",
       title = "Sampling Distribution of the Mean",
       subtitle = "Population is exponential (skewed). As n grows, the mean becomes Normal.") +
  theme_bw()
```

::: {.callout-note}
Notice two things as $n$ increases: (1) the shape becomes bell-shaped even
though the underlying data are skewed; (2) the spread narrows -- the standard
error $\sigma/\sqrt{n}$ shrinks. Both are exactly what the CLT predicts.
:::

Confirm the standard error formula numerically:

```{r}
pop_sd <- sd(population)

# Theoretical SE = sigma / sqrt(n)
pop_sd / sqrt(5)
pop_sd / sqrt(30)
pop_sd / sqrt(100)

# Observed SD of the simulated sample means
sd(means_n5)
sd(means_n30)
sd(means_n100)
```

## A First Hypothesis Test: One-Sample t-Test

The average systolic blood pressure in the general US adult population is
approximately 122 mmHg. Is the Framingham cohort different from this reference?

```{r}
# Always look first
framingham |>
  ggplot(aes(x = sysBP)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 122, linetype = "dashed", color = "firebrick",
             linewidth = 1) +
  annotate("text", x = 125, y = 250, label = "US average (122 mmHg)",
           color = "firebrick", hjust = 0, size = 3.5) +
  labs(x = "Systolic BP (mmHg)", y = "Count",
       title = "Systolic BP in the Framingham Cohort") +
  theme_bw()
```

```{r}
t.test(framingham$sysBP, mu = 122)
```

**Reading the output:**

- `t` is the test statistic: how many standard errors the sample mean is from 122.
- `df` is degrees of freedom ($n - 1$).
- `p-value` -- here essentially 0 -- means the data are extremely unlikely under $H_0: \mu = 122$.
- The 95% CI gives a plausible range for the true population mean.

A complete written interpretation:

> Mean systolic BP in the Framingham cohort was 132.4 mmHg (95% CI: 131.9--132.9),
> significantly higher than the US reference value of 122 mmHg
> ($t_{4238} = 43.1$, $p < 0.001$).

Notice that the p-value alone tells you almost nothing useful here. The CI and
the mean tell you that the cohort runs about 10 mmHg higher -- that is the
finding.

## Two-Sample t-Test

Do male and female participants have different mean systolic BP? Before running
any test, always look at the data:

```{r}
framingham |>
  ggplot(aes(x = male, y = sysBP, fill = male)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = "Sex", y = "Systolic BP (mmHg)",
       title = "Systolic BP by Sex") +
  theme_bw()
```

```{r}
# Group-level summaries first
framingham |>
  group_by(male) |>
  summarize(
    n        = n(),
    mean_bp  = mean(sysBP, na.rm = TRUE),
    sd_bp    = sd(sysBP, na.rm = TRUE)
  )
```

```{r}
# Welch two-sample t-test (default -- does not assume equal variances)
t.test(sysBP ~ male, data = framingham)
```

::: {.callout-note}
The formula interface `y ~ group` is the cleanest way to run a two-sample test
on a data frame. It is equivalent to splitting the column yourself:
`t.test(females$sysBP, males$sysBP)`.
:::

Reading the output:

- The estimated difference is shown at the bottom: mean in Female minus mean in
  Male. Here females have lower mean systolic BP.
- The 95% CI is for the **difference** $\mu_\text{Female} - \mu_\text{Male}$,
  not for either mean individually.
- Because the CI does not contain 0, we reject $H_0: \mu_1 = \mu_2$.

A complete written interpretation:

> Mean systolic BP was 131.0 mmHg (SD = 22.0) in females and 134.5 mmHg
> (SD = 17.6) in males. Males had significantly higher mean systolic BP
> (difference = −3.5 mmHg, 95% CI: −5.1 to −1.9; $t_{3912} = −4.3$,
> $p < 0.001$).

Note that the difference of 3.5 mmHg is statistically significant — because
$n$ is large — but modest in clinical terms. Always report the effect size
alongside the p-value.

---

## Paired t-Test

Systolic and diastolic BP are measured on the same person. Pulse pressure
(their difference) reflects arterial stiffness. We want to know whether mean
systolic BP and mean diastolic BP differ — and because they come from the same
individuals, the comparison is inherently paired.

```{r}
# Compute the per-person differences first and look at them
framingham <- framingham |>
  mutate(pulse_pressure = sysBP - diaBP)

framingham |>
  ggplot(aes(x = pulse_pressure)) +
  geom_histogram(binwidth = 3, fill = "steelblue", color = "white") +
  labs(x = "Pulse Pressure = Systolic − Diastolic (mmHg)",
       y = "Count",
       title = "Distribution of Per-Person Differences") +
  theme_bw()
```

The paired t-test is a one-sample t-test on the differences. The two calls
below are exactly equivalent:

```{r}
# Using the pre-computed difference
t.test(framingham$pulse_pressure, mu = 0)

# Using the paired argument directly
t.test(framingham$sysBP, framingham$diaBP, paired = TRUE)
```

To show why pairing matters, compare the paired result to the naive unpaired
test on the same data:

```{r}
# Unpaired (wrong here -- treats the two columns as independent groups)
t.test(framingham$sysBP, framingham$diaBP, paired = FALSE)
```

Both tests give $p < 0.001$ here because the sample is so large, but notice
that the unpaired test produces a **much wider confidence interval**. In a
smaller study the unpaired test might fail to detect a real difference entirely,
because it is wasting power on between-person variability that the paired design
already removed. Using the wrong test costs you precision.

---

## Chi-Square Test

Is smoking status associated with prevalent hypertension? Both variables are
categorical, so we need a chi-square test.

```{r}
# Always look at the raw counts first
tab <- table(Smoker = framingham$currentSmoker,
             Hypertension = framingham$prevalentHyp)
tab
```

```{r}
# Proportions within each smoking group (row proportions)
prop.table(tab, margin = 1) |> round(3)
```

```{r}
# Visualize before testing
framingham |>
  filter(!is.na(currentSmoker), !is.na(prevalentHyp)) |>
  group_by(currentSmoker, prevalentHyp) |>
  summarize(n = n(), .groups = "drop") |>
  group_by(currentSmoker) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = currentSmoker, y = pct, fill = prevalentHyp)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Smoking Status", y = "Percent (%)",
       fill = "Hypertension",
       title = "Hypertension Prevalence by Smoking Status") +
  theme_bw()
```

```{r}
result_chisq <- chisq.test(tab)
result_chisq
```

```{r}
# Always check expected counts -- all should be >= 5
result_chisq$expected
```

All expected counts are well above 5, so the chi-square approximation is valid.

Reading the output:

- $\chi^2$ measures total discrepancy between observed and expected counts
  under independence.
- `df` = (rows − 1) × (cols − 1) = 1 × 1 = 1 for a 2×2 table.
- A small p-value means the observed pattern is very unlikely if smoking and
  hypertension were independent.

::: {.callout-important}
The chi-square test only tells you **whether** an association exists, not how
large it is. Always report the proportions from `prop.table()` alongside the
test result so the reader can judge the magnitude.
:::

A complete written interpretation:

> Hypertension was present in 32.1% of non-smokers and 29.8% of smokers.
> The association between smoking status and hypertension was statistically
> significant ($\chi^2_1 = 4.28$, $p = 0.039$), though the absolute difference
> in prevalence was small (2.3 percentage points).

---

## Fisher's Exact Test

Fisher's exact test is used instead of the chi-square test when expected cell
counts are small (any cell < 5). To illustrate, we look at the association
between prevalent stroke and diabetes — both rare in this cohort, so cell
counts will be small.

```{r}
tab_rare <- table(Stroke   = framingham$prevalentStroke,
                  Diabetes = framingham$diabetes)
tab_rare
```

```{r}
chisq.test(tab_rare)$expected
```

The "Yes/Yes" cell has an expected count below 5. The chi-square approximation
is unreliable here. Use Fisher's exact test instead:

```{r}
fisher.test(tab_rare)
```

Fisher's exact test computes the exact probability of the observed table (and
all more extreme tables) under $H_0$, without relying on any large-sample
approximation. It also returns an **odds ratio** with a 95% CI as the effect
size measure.

Reading the output:

- `p-value`: exact probability; no approximation involved.
- `odds ratio`: the ratio of the odds of stroke in diabetic vs. non-diabetic
  participants. An OR > 1 means higher odds in the first group.
- The 95% CI for the OR excludes 1, confirming a significant association.

::: {.callout-note}
For 2×2 tables, `fisher.test()` always works and is never wrong to use. The
chi-square test is fine when expected counts are all ≥ 5 and you have a larger
table where Fisher's exact becomes computationally intensive.
:::

---

## Wilcoxon Rank-Sum Test (Two Independent Groups)

Glucose values in the Framingham data are right-skewed, particularly in
diabetic participants. When Normality is questionable and $n$ is not large
enough for the CLT to rescue us, the Wilcoxon rank-sum test is the safer
choice.

```{r}
framingham |>
  filter(!is.na(diabetes)) |>
  ggplot(aes(x = diabetes, y = glucose, fill = diabetes)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = "Diabetes", y = "Glucose (mg/dL)",
       title = "Glucose by Diabetes Status") +
  theme_bw()
```

```{r}
# Separate the two groups
glucose_no  <- framingham |> filter(diabetes == "No")  |> pull(glucose)
glucose_yes <- framingham |> filter(diabetes == "Yes") |> pull(glucose)

# Distribution check
par(mfrow = c(1, 2))
hist(glucose_no,  main = "Non-diabetic", xlab = "Glucose", col = "steelblue")
hist(glucose_yes, main = "Diabetic",     xlab = "Glucose", col = "firebrick")
par(mfrow = c(1, 1))
```

The diabetic group is clearly right-skewed with outliers. The Wilcoxon test
works on ranks, so these extreme values have much less influence than they
would in a t-test.

```{r}
wilcox.test(glucose ~ diabetes, data = framingham,
            conf.int = TRUE)   # request the Hodges-Lehmann CI
```

Reading the output:

- `W` is the rank-sum statistic.
- `p-value`: probability of a rank-sum this extreme under $H_0$ (no systematic
  difference between groups).
- `difference in location`: the Hodges-Lehmann estimator -- the median of all
  pairwise differences between one group and the other. This is the effect size
  for a Wilcoxon test.
- The 95% CI is for this location shift, not for the difference in means.

Compare to what the t-test would give on the same data:

```{r}
t.test(glucose ~ diabetes, data = framingham)
```

Both tests give $p < 0.001$ here, but the t-test CI is for the difference in
**means** while the Wilcoxon CI is for the **location shift** (closer to the
difference in medians). With heavily skewed data, the median shift is often the
more honest summary.

---

## Wilcoxon Signed-Rank Test (Paired)

The signed-rank test is the paired version of the Wilcoxon test. Use it when
you have matched pairs and the differences are not Normally distributed.

To illustrate, we compare total cholesterol at baseline to a hypothetical
follow-up. Since we only have one time point in Framingham, we will simulate
a follow-up measurement by adding a small random change to each person's
cholesterol — just enough to demonstrate the mechanics.

```{r}
set.seed(42)
n <- sum(!is.na(framingham$totChol))

chol_baseline <- framingham |>
  filter(!is.na(totChol)) |>
  pull(totChol)

# Simulate follow-up: a 5 mg/dL mean reduction with right-skewed noise
chol_followup <- chol_baseline - rexp(n, rate = 0.2) + 2

# Look at the differences
diff_chol <- chol_followup - chol_baseline

tibble(diff = diff_chol) |>
  ggplot(aes(x = diff)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(x = "Change in Cholesterol (follow-up − baseline, mg/dL)",
       y = "Count",
       title = "Distribution of Paired Differences",
       subtitle = "Right-skewed — Wilcoxon signed-rank is appropriate") +
  theme_bw()
```

```{r}
# Wilcoxon signed-rank test on the paired differences
wilcox.test(chol_followup, chol_baseline,
            paired   = TRUE,
            conf.int = TRUE)
```

```{r}
# For comparison -- the paired t-test on the same data
t.test(chol_followup, chol_baseline, paired = TRUE)
```

Because the differences are right-skewed (the simulated noise is exponential),
the Wilcoxon test is the more trustworthy result here. In practice, always
check the distribution of the **differences** (not the raw values) when
deciding between a paired t-test and the signed-rank test.

::: {.callout-note}
**Summary: which test when?**

| Situation | Normal / large $n$? | Test |
|---|---|---|
| One group vs. reference | Yes | One-sample t |
| Two independent groups | Yes | Welch two-sample t |
| Two paired groups | Yes | Paired t |
| Two independent groups | No | Wilcoxon rank-sum |
| Two paired groups | No | Wilcoxon signed-rank |
| Two categorical variables, large counts | — | Chi-square |
| Two categorical variables, small counts | — | Fisher's exact |
:::


---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document. Title it `"Day 5 Lab: Inference"`,
select HTML, click **Create**.

**Step 2.** Replace the template with this YAML and setup chunk:

````markdown
---
title: "Day 5 Lab: Inference"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{{r}}
library(tidyverse)
library(this.path)
this.dir <- this.dir()
setwd(this.dir)

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male            = factor(male,            levels = c(0, 1),
                                              labels = c("Female", "Male")),
    currentSmoker   = factor(currentSmoker,   levels = c(0, 1),
                                              labels = c("Non-smoker", "Smoker")),
    BPMeds          = factor(BPMeds,          levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentStroke = factor(prevalentStroke, levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentHyp    = factor(prevalentHyp,    levels = c(0, 1),
                                              labels = c("No", "Yes")),
    diabetes        = factor(diabetes,        levels = c(0, 1),
                                              labels = c("No", "Yes")),
    TenYearCHD      = factor(TenYearCHD,      levels = c(0, 1),
                                              labels = c("No", "Yes")),
    education       = factor(education,       levels = 1:4,
                                              labels = c("Some HS", "HS Diploma",
                                                         "Some College",
                                                         "College Degree"))
  )
```
````

**Step 3.** Work through Exercises 1--6. For each:

- Add a code chunk and write your code
- Write 2--3 sentences below the chunk with a plain-language interpretation of
  your result -- not just what the numbers are, but what they mean biologically
  or clinically
- Always make a plot before running a test

**Step 4.** Render when finished and fix any errors.

::: {.callout-tip}
**Before running any test**, always: (1) make a histogram or boxplot to look at
the data, (2) check whether the assumptions are plausible, (3) then run the
test. This is not just good practice -- it will catch data problems that would
otherwise produce misleading results.
:::

::: {.callout-note}
Exercises 1--4 cover one test each and are the core material. Exercise 5
requires you to choose and justify the right test yourself. Exercise 6 is a
challenge that brings in visualization and multiple comparisons.
:::

---

**Exercise 1 — One-Sample t-Test.**

The normal resting diastolic blood pressure for adults is considered to be
around 80 mmHg.

a. Make a histogram of `diaBP` in the Framingham cohort. Does the distribution
   look approximately Normal? Add a vertical line at 80 mmHg.

b. Run a one-sample t-test testing whether mean diastolic BP in the cohort
   differs from 80 mmHg. Report the test statistic, degrees of freedom,
   p-value, and 95% CI.

c. Write a 2--3 sentence interpretation of the result. Is the difference
   statistically significant? Is it practically meaningful? (Think about what
   a difference of a few mmHg means clinically.)

```{r}
# your code here
```

---

**Exercise 2 — Two-Sample t-Test.**

Do male and female participants have different mean systolic blood pressure?

a. Make side-by-side boxplots of `sysBP` by `male`. Do the distributions look
   approximately Normal within each group? Are the spreads similar?

b. Run a Welch two-sample t-test comparing mean `sysBP` between males and
   females. Report the estimated difference, 95% CI for the difference,
   t-statistic, and p-value.

c. Write a plain-language interpretation. Make sure to report the actual
   estimated difference in mmHg, not just the p-value.

d. **Assumption check:** Make a Q-Q plot for each group separately using the
   code below. Does Normality look reasonable?

```{r}
# Q-Q plot template -- fill in the group name
framingham |>
  filter(male == "Female") |>
  pull(sysBP) |>
  qqnorm(main = "Q-Q Plot: Females")
qqline(framingham |> filter(male == "Female") |> pull(sysBP))
```

```{r}
# your test code here
```

---

**Exercise 3 — Paired t-Test.**

Systolic and diastolic BP are measured on the same person, so any comparison
between them is inherently paired. Pulse pressure (the difference between
systolic and diastolic BP) is a clinically meaningful measure of arterial
stiffness.

a. Create a new variable `pulse_pressure = sysBP - diaBP`. Make a histogram
   of `pulse_pressure`. What is its mean and standard deviation?

b. Run a paired t-test using `t.test(framingham$sysBP, framingham$diaBP, paired = TRUE)`.
   Confirm that the mean of the differences in the output matches the mean
   `pulse_pressure` you computed in (a).

c. The null hypothesis here is $H_0: \mu_{\text{sys}} - \mu_{\text{dia}} = 0$,
   i.e., no difference on average. Does the result surprise you? Why or why not?

d. Now run an **unpaired** two-sample t-test on the same data:
   `t.test(framingham$sysBP, framingham$diaBP)`.
   Compare the p-values and confidence intervals. Which test is more
   appropriate and why?

```{r}
# your code here
```

---

**Exercise 4 — Chi-Square Test.**

Is smoking status associated with prevalent hypertension in this cohort?

a. Create a contingency table of `currentSmoker` by `prevalentHyp` using
   `table()`. What are the cell counts? Do all expected counts look large
   enough for a chi-square test (rule: all expected counts ≥ 5)?

b. Run `chisq.test()` on the table. Report $\chi^2$, degrees of freedom, and
   p-value.

c. Compute the proportion with hypertension separately for smokers and
   non-smokers using `group_by()` and `summarize()`. How large is the
   difference in proportions?

d. Make a grouped bar chart showing hypertension prevalence by smoking status.
   Use `geom_col()` after computing proportions with `dplyr`.

e. Write a 2--3 sentence interpretation. Is the association statistically
   significant? Is the effect size meaningful?

```{r}
# your code here
```

---

**Exercise 5 — Choose the Right Test.**

For each of the following questions, (i) state which test you would use and
why, (ii) check the relevant assumption(s), and (iii) run the test and
interpret the result.

**Question A.** Is mean BMI different between diabetic and non-diabetic
participants?

**Question B.** Is mean glucose different between diabetic and non-diabetic
participants? (Make a histogram of `glucose` within each group first. Does the
distribution look Normal? Does your answer change which test you use?)

**Question C.** Is education level associated with smoking status? (Both
variables are categorical -- what test applies here?)

```{r}
# Question A
```

```{r}
# Question B
```

```{r}
# Question C
```

---

**Exercise 6 (Challenge) — Multiple Comparisons and Effect Sizes.**

We will compare mean total cholesterol (`totChol`) across all four education
levels.

a. Make a boxplot of `totChol` by `education`. Order the groups by median
   cholesterol using `reorder()`.

b. Run pairwise two-sample t-tests across all education groups using:

```{r}
#| eval: false
pairwise.t.test(framingham$totChol, framingham$education,
                p.adjust.method = "BH")  # Benjamini-Hochberg correction
```

The result is a matrix of adjusted p-values. Which pairs of education groups
differ significantly (adjusted $p < 0.05$)?

c. For each pair that is significant, compute the difference in means and a
   95% CI using a standard `t.test()`. Are the differences large enough to be
   clinically meaningful? (A difference of about 10 mg/dL in total cholesterol
   is generally considered clinically notable.)

d. Why do we need to adjust p-values when running multiple tests? What would
   happen to the false positive rate if we ran 6 pairwise tests each at
   $\alpha = 0.05$ without adjustment?

```{r}
# your code here
```

---

# Quick Reference Card

## Distributions in R

| Distribution | Random draws | Probability $\leq q$ | Value at quantile $p$ |
|---|---|---|---|
| Normal | `rnorm(n, mean, sd)` | `pnorm(q, mean, sd)` | `qnorm(p, mean, sd)` |
| Binomial | `rbinom(n, size, prob)` | `pbinom(q, size, prob)` | `qbinom(p, size, prob)` |
| Poisson | `rpois(n, lambda)` | `ppois(q, lambda)` | `qpois(p, lambda)` |
| t | `rt(n, df)` | `pt(q, df)` | `qt(p, df)` |

## Hypothesis Tests

| Situation | Function | Key arguments |
|---|---|---|
| One group mean vs. reference | `t.test(x, mu = value)` | `mu`: null value |
| Two independent group means | `t.test(x, y)` | Welch by default |
| Two paired measurements | `t.test(x, y, paired = TRUE)` | vectors must be same length and order |
| Two categorical variables | `chisq.test(table(var1, var2))` | expects a table of counts |
| Small expected counts | `fisher.test(table(var1, var2))` | use when any expected count < 5 |
| Non-Normal, two groups | `wilcox.test(x, y)` | add `conf.int = TRUE` for CI |
| Non-Normal, paired | `wilcox.test(x, y, paired = TRUE)` | |
| Multiple group comparisons | `pairwise.t.test(x, g, p.adjust.method = "BH")` | |

## Reading t-test Output

```r
t.test(x, mu = 0)

#   One Sample t-test
# t = 3.45, df = 99, p-value = 0.0008
# 95 percent confidence interval: 1.2  4.8
# sample estimates: mean of x = 3.0
```

- **t**: test statistic (sample mean − null value) / standard error
- **df**: degrees of freedom ($n - 1$ for one-sample)
- **p-value**: probability of a result this extreme if $H_0$ is true
- **95% CI**: plausible range for the true population mean
- Always report: estimated mean (or difference), CI, t, df, p

## Assumption Checking

```r
# Histogram
ggplot(df, aes(x = var)) + geom_histogram()

# Q-Q plot (base R -- points near the line = approximately Normal)
qqnorm(x); qqline(x)

# Check expected counts after chi-square
chisq.test(tab)$expected   # all should be >= 5
```

## Common Mistakes to Avoid

- Reporting only a p-value without the estimated effect and CI
- Saying the p-value is "the probability the null hypothesis is true" -- it is not
- Using a paired test on independent data, or an independent test on paired data
- Forgetting `na.rm = TRUE` in `mean()`, `sd()`, etc.
- Choosing your test *after* seeing which one gives $p < 0.05$
