---
title: "Day 10 (PM) Lab: Survival Analysis"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This lab puts the Day 10 morning material into practice. You will work with
time-to-event data, construct Kaplan-Meier curves, run log-rank tests, and
fit Cox proportional hazards models using the `lung` dataset.

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** Follow along as the instructor works
through a complete KM and Cox analysis using a different grouping variable
than the morning example.
- **Part 2 (on your own, ~90 min):** Five exercises of increasing difficulty.
Create a new `.qmd` file, work through each exercise, render when done.

The **Quick Reference Card** in Section 3 has all the key functions.

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(this.path)
library(survival)
setwd(this.dir())

data(lung)
lung <- as_tibble(lung) |>
  mutate(
    sex_label  = factor(sex,      levels = c(1, 2),
                        labels = c("Male", "Female")),
    pat.karno.gr  = factor(
      ifelse(pat.karno > 80, "good", "poor")),
    ecog_group = factor(ph.ecog,  levels = c(0, 1, 2, 3),
                        labels = c("0 (fully active)",
                                   "1 (ambulatory)",
                                   "2 (in bed <50%)",
                                   "3 (in bed >50%)"))
  )
```

## Step 1: Examine the Data

We will look at survival by ECOG performance score group.

```{r}
lung |>
  count(ecog_group, status) |>
  mutate(status_label = if_else(status == 2, "Dead", "Censored")) |>
  select(-status) |>
  pivot_wider(names_from = status_label, values_from = n, values_fill = 0)
```

Most patients in this dataset have ECOG scores of 0, 1, or 2. Very few have
score 3. As a result, we'll combine those with a score of 3 to those with a score of 2:


```{r}
lung <- lung |>
  filter(!is.na(ecog_group)) |> 
  mutate(
    ecog_group = fct_collapse(ecog_group, 
                               "0 (fully active)" = "0 (fully active)",
                               "1 (ambulatory)" = "1 (ambulatory)",
                               "2+ (non-ambulatory)" = c("2 (in bed <50%)",
                               "3 (in bed >50%)"))
    )
lung |>
  count(ecog_group, status) |>
  mutate(status_label = if_else(status == 2, "Dead", "Censored")) |>
  select(-status) |>
  pivot_wider(names_from = status_label, values_from = n, values_fill = 0)
```

## Step 2: KM Curves by ECOG Group

```{r}
km_ecog <- survfit(Surv(time, status == 2) ~ ecog_group, data = lung)
print(km_ecog)
```

```{r}
#| fig-width: 8
#| fig-height: 5
plot(km_ecog,
     xlab      = "Time (days)",
     ylab      = "Survival probability",
     main      = "Kaplan-Meier Curves by ECOG Performance Score",
     col       = c("steelblue", "darkgreen", "firebrick"),
     lwd       = 2,
     conf.int  = FALSE,
     mark.time = TRUE)
legend("topright",
       legend = c("0 (fully active)", "1 (ambulatory)",
                  "2+ (non-ambulatory)"),
       col    = c("steelblue", "darkgreen", "firebrick"),
       lwd    = 2,
       cex    = 0.85)
abline(h = 0.5, lty = 2, col = "gray50")
```

There is a clear ordering: patients with better performance scores (lower ECOG)
survive longer. Median survival decreases substantially with each step up in
ECOG score.

## Step 3: Log-Rank Test

```{r}
survdiff(Surv(time, status == 2) ~ ecog_group, data = lung)
```

The log-rank test compares survival across all four ECOG groups simultaneously.
A small p-value confirms that survival differs across groups well beyond what
chance alone would produce.

## Step 4: Simple Cox Model for ECOG

```{r}
cox_ecog <- coxph(Surv(time, status == 2) ~ ph.ecog, data = lung)
exp(cbind(HR = coef(cox_ecog), confint(cox_ecog))) |> round(3)
```

Treating ECOG as continuous, each one-unit increase in score is associated
with approximately 50% higher hazard of death. This is the unadjusted effect.

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document: File -> New File -> Quarto Document.
Title it `"Day 10 Lab: Survival Analysis"`, select HTML, click **Create**.

**Step 2.** Replace the template content with this YAML and setup chunk:

````markdown
---
title: "Day 10 Lab: Survival Analysis"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
html:
toc: true
theme: cosmo
execute:
echo: true
warning: false
message: false
---

```{{r}}
library(tidyverse)
library(this.path)
library(survival)
setwd(this.dir())

data(lung)
lung <- as_tibble(lung) |>
mutate(
sex_label = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
ecog_group = factor(ph.ecog, levels = c(0, 1, 2, 3),
labels = c("0 (fully active)", "1 (ambulatory)",
"2 (in bed <50%)", "3 (in bed >50%)"))
)
```
````

**Step 3.** Work through Exercises 1--5 in order. For each one, add a code
chunk, write and run your code, and add 2--3 sentences interpreting the result
in plain language.

**Step 4.** Render when finished.

::: {.callout-tip}
**Before asking for help**, check the **Quick Reference Card** in Section 3.
The most common errors in survival analysis are: (1) forgetting `status == 2`
inside `Surv()`, and (2) passing a factor instead of a numeric variable to
`cox_ph` when you intend a continuous predictor.
:::

::: {.callout-note}
Exercises 1--3 cover the core KM and log-rank material. Exercise 4 builds a
Cox model and interprets hazard ratios. Exercise 5 is a full workflow tying
everything together.
:::

---

**Exercise 1 -- Exploring the Data.**

a. Use `count()` to tabulate the number of events (status == 2) and censored
observations (status == 1) separately for males and females. What proportion
of each sex group died during the study?

b. Use `group_by()` and `summarize()` to compute the mean and median follow-up
time separately for those who died and those who were censored. How do they
compare? What does this tell you about the timing of censoring? **(Hint: see section 4.2 of the morning notes)**

c. Produce a histogram of follow-up times (`time`), colored by event status.
Use `fill = if_else(status == 2, "Dead", "Censored")` inside `aes()` and
`scale_fill_manual()` to use meaningful colors.

```{r}
# your code here
```

---

**Exercise 2 -- Kaplan-Meier Curve for the Full Cohort.**

a. Fit a KM estimator for the overall cohort using
`survfit(Surv(time, status == 2) ~ 1, data = lung)`. Assign it to an object
and use `print()` to see the summary.

b. What is the estimated median survival time and its 95% confidence interval?
In plain language, what does the median survival time mean?

c. Plot the KM curve with `plot()`. Include `conf.int = TRUE` and
`mark.time = TRUE`. Add a horizontal dashed line at 0.5 using `abline()`.
Label the axes clearly.

d. From the curve, approximately what is $\hat{S}(365)$, the estimated
probability of surviving more than one year? (Read it from the plot or use
`summary(km_fit, times = 365)`.)

```{r}
# your code here
```

---

**Exercise 3 -- Comparing Survival Curves.**

a. Fit separate KM curves by sex using
`survfit(Surv(time, status == 2) ~ pat.karno.gr, data = lung)`.
Use `print()` to get median survival for each group. Which group has
longer median survival?

b. Plot the two KM curves on the same figure using different colors. Add a
legend. Do the curves separate early, late, or consistently throughout
follow-up?

c. Run a log-rank test using `survdiff(Surv(time, status == 2) ~ pat.karno.gr,
data = lung)`. Report the chi-squared statistic and p-value. Is there
statistically significant evidence that survival differs by sex?

d. **Interpretation.** The log-rank test gives a p-value. What does a small
p-value actually tell us? What does it NOT tell us that the Cox model will?

```{r}
# your code here
```

---

**Exercise 4 -- Cox Proportional Hazards Model.**

a. Fit a Cox model with `pat.karno.gr` as the only predictor:
`coxph(Surv(time, status == 2) ~ pat.karno.gr, data = lung)`. Use
`exp(cbind(HR = coef(model), confint(model)))` to get the hazard ratio
and 95% CI.

b. Interpret the HR for sex in one sentence. Remember: HR > 1 means higher
hazard (worse survival), HR < 1 means lower hazard (better survival).

c. Now add age and sex to the model. After adjusting for these sex,
does the HR for `pat.karno.gr` change? 

d. Report and interpret the adjusted HR for `pat.karno.gr`. 

e. Use `anova(reduced_model, full_model)` to formally test whether age and
sex jointly add significant prognostic information beyond `pat.karno.gr` alone.

```{r}
# your code here
```

---

**Exercise 5 (Challenge) -- Full Workflow: Building a Prognostic Model.**

Build a multivariable Cox model for overall survival using several predictors.

a. Restrict to complete cases on `time`, `status`, `sex`, `age`, `ph.ecog`,
and `wt.loss` using `drop_na()`. How many patients remain?

b. Check which variables are associated with survival by fitting a separate
simple Cox model for each predictor: `sex_label`, `age`, `ph.ecog`, and
`wt.loss`. Report the unadjusted HR and 95% CI for each.

c. Fit a full Cox model with all four predictors and report the adjusted HRs.
Which predictors remain significant after adjustment?

d. Use `anova()` to test whether `wt.loss` adds significant information beyond
the other three predictors. Report the LRT p-value.

e. Check the proportional hazards assumption for sex by plotting the KM curves
on a log-log scale (`fun = "cloglog"` in `plot(survfit(...))`). Do the
curves look approximately parallel?

f. **Discussion.** In the morning lecture, the sex HR was about 0.59 in the
simple model and 0.58 in the adjusted model. Based on your results in this
exercise, do the additional variables (age, ECOG, weight loss) appear to
confound the sex-survival relationship? How do you know?

```{r}
# your code here
```

---

# Quick Reference Card

## Setup

```r
library(tidyverse)
library(this.path)
library(survival)
setwd(this.dir())

data(lung)
lung <- as_tibble(lung)
```

## Creating a Survival Object

```r
# status == 2 means event occurred; status == 1 means censored
Surv(time, status == 2)

# Add to dataset
df <- df |> mutate(surv = Surv(time, status == 2))
```

## Kaplan-Meier Estimation

```r
# Overall KM curve
km_fit <- survfit(Surv(time, status == 2) ~ 1, data = df)

# KM curve by group
km_by_group <- survfit(Surv(time, status == 2) ~ group, data = df)

# Print median survival
print(km_fit)
summary(km_fit)$table

# Survival probability at specific times
summary(km_fit, times = c(180, 365, 730))
```

## Plotting KM Curves

```r
# Basic plot
plot(km_fit,
xlab      = "Time (days)",
ylab      = "Survival probability",
col       = "steelblue",
lwd       = 2,
conf.int  = TRUE,
mark.time = TRUE)   # tick marks for censored observations

# Add horizontal line at median (S = 0.5)
abline(h = 0.5, lty = 2, col = "gray50")

# Two-group plot
plot(km_sex,
col  = c("steelblue", "firebrick"),
lwd  = 2,
conf.int  = FALSE,
mark.time = TRUE)
legend("topright",
legend = c("Group 1", "Group 2"),
col    = c("steelblue", "firebrick"),
lwd    = 2)
```

## Log-Rank Test

```r
survdiff(Surv(time, status == 2) ~ group, data = df)
# Reports chi-squared statistic and p-value
```

## Cox Proportional Hazards Model

```r
# Simple Cox model
cox_model <- coxph(Surv(time, status == 2) ~ sex + age, data = df)
summary(cox_model)

# Hazard ratios and 95% CIs
exp(cbind(HR = coef(cox_model), confint(cox_model)))

# Rescale a continuous HR to meaningful units
exp(10 * coef(cox_model)["age"])       # HR for 10-year age increase
exp(2  * coef(cox_model)["ph.ecog"])   # HR for 2-unit ECOG increase

# Compare nested models (LRT)
anova(cox_reduced, cox_full)

# Complete cases (for fair model comparisons)
vars <- c("time", "status", "sex", "age", "ph.ecog")
df_cc <- df |> drop_na(all_of(vars))
```

## Checking the Proportional Hazards Assumption

```r
# Log-log plot: curves should be approximately parallel
plot(km_by_group,
fun  = "cloglog",
xlab = "Log time",
ylab = "log(-log(S(t)))")
```

## Interpreting Hazard Ratios

| HR | Meaning |
|---|---|
| HR = 1 | No difference in hazard |
| HR = 2.0 | Twice the hazard (worse survival) |
| HR = 0.5 | Half the hazard (better survival) |
| HR = 1.5 | 50% higher hazard |
| HR = 0.7 | 30% lower hazard |
| CI excludes 1 | Statistically significant at alpha = 0.05 |

## Key Vocabulary

| Term | Meaning |
|---|---|
| Event | The outcome of interest (death, relapse, etc.) |
| Censored | Participant left follow-up without the event |
| S(t) | Probability of surviving beyond time t |
| h(t) | Instantaneous event rate at time t |
| Median survival | Time at which S(t) = 0.5 |
| HR | Multiplicative effect on the hazard |
| Log-rank test | Tests H0: survival curves are equal |
| PH assumption | HR is constant over time |
