---
title: "Summer Course Materials"
output:
  pdf_document: default
  html_document: default
date: "2026-05-13"
---


# Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# tidyverse package required, uncomment to install if needed 
# install.packages("tidyverse")

library(tidyverse)
```

---

# 1. Airway Dataset

## Source

Bioconductor package: `airway`

## Description

This package provides a RangedSummarizedExperiment object of read counts in genes for an RNA-Seq experiment on human airway smooth muscle cell lines treated with dexamethasone. There are 8 cell samples (4 treatment and 4 control) with 63,677 genes in each sample. This is best suited for material from days 11-14.

## Load Test

```{r airway-load, message=FALSE, warning=FALSE}

# uncomment to install packages required for 'airway'

# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")

# BiocManager::install("airway")

library(airway)

data(airway)

airway
```

## Exploratory Checks

```{r airway-checks}
class(airway)
dim(airway)

airway_counts <- assay(airway)
head(airway_counts)

# metadata for samples and genes
sample_info <- as.data.frame(colData(airway))
gene_info <- as.data.frame(rowData(airway))

# check for missing values
colSums(is.na(airway_counts))
```

## Cleaning Notes

assay(airway) extracted the actual gene expression data from the larger RangedSummarizedExperiment container. Metadata must be extracted and stored in separate variables. 

---

# 2. Framingham Dataset

## Source

BIOS 735 course materials (framingham.csv)

## Description

This dataset contains demographic and cardiovascular risk factor information collected for the Framingham Heart Study, a longitudinal study investigating predictors of coronary heart disease. It is well-suited for linear and logistic regression models, covered in days 1-8 of the course.

## Load Test

```{r framingham-load}
# .csv file download required
library(rstudioapi)
this.dir <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(this.dir)
framingham <- read.csv("framingham.csv")

head(framingham)
```

## Exploratory Checks

```{r framingham-checks}
str(framingham)
summary(framingham)

# check for missing values
colSums(is.na(framingham))
```

## Cleaning Notes

There are some missing values to be aware of when using for analysis (education, cigsPerDay, BPMeds, totChol, BMI, heartRate, and glucose variables). 

---

# 3. InsectSprays Dataset

## Source

Built-in base R dataset

## Description

This dataset tracks effectiveness of six different insecticides. It contains 72 observations with 2 variables: count of insects in each agricultural experimental unit and type of insecticide used (A-F). It is best suited for Poisson regression or things involving count data. Specifically, this could be used in days 1-5 and for GLM/Poisson examples on day 8.


## Load Test

```{r insectsprays-load}
data(InsectSprays)

head(InsectSprays)
```

## Exploratory Checks

```{r insectsprays-checks}
str(InsectSprays)
summary(InsectSprays)

table(InsectSprays$spray)

# check for missing values
colSums(is.na(InsectSprays))
```

## Cleaning Notes

N/A

---

# 4. Sleep Study Dataset

## Source

Package: `lme4`

## Description

This dataset contains the average reaction time on a series of tests given each day to each subject in a sleep deprivation study. On day 0, subjects had normal amount of sleep. Starting that night, they were restricted to 3 hours of sleep per night for 9 nights. It includes repeated measures on subjects, so this would be particularly useful for days 4 and 10.



## Load Test

```{r sleepstudy-load}
library(lme4)
data("sleepstudy")

head(sleepstudy)
```
## Exploratory Checks

```{r sleepstudy-checks}

str(sleepstudy)
summary(sleepstudy)

# check for missing values
colSums(is.na(sleepstudy))
```

## Cleaning Notes

N/A

---

# 5. Lung Dataset

## Source

Package: `survival`

## Description

This survival dataset includes observations from patients with advanced stage lung cancer from the North Central Cancer Treatment Group.It includes demographic variables and performance scores that rate how well a patient can perform usual daily activities.


## Load Test

```{r lung-load}
# survival package required, uncomment to install if needed 
# install.packages("survival")

library(survival)

head(lung)
```

## Exploratory Checks

```{r lung-checks}

str(lung)
summary(lung)

# check for missing values
colSums(is.na(lung))
```

## Cleaning Notes

There are some missing values to be aware of when using for analysis (inst, ph.ecog, ph.karno, pat.karno, meal.cal, and wt.loss variables). 
